\documentclass[11pt,a4paper,openright,twoside]{memoir}

\input{preamble}

\begin{document}

% === Front Matter ===
\frontmatter

\input{frontmatter/cover}

\input{frontmatter/titlepage}

\cleardoublepage
\tableofcontents

\cleardoublepage
\input{frontmatter/preface}

\cleardoublepage
\input{frontmatter/notation}

% === Main Matter ===
\mainmatter

% --- Part I: Foundations ---
\part{Foundations}
\label{part:foundations}

\input{chapters/ch01-what-is-doxonomics}
\input{chapters/ch02-ontology-of-public-belief}
\input{chapters/ch03-prehistory-of-doxonomics}
\input{chapters/ch04-core-units-of-analysis}
\input{chapters/ch05-belief-production-chain}

\begin{interlude}
You now have the foundational vocabulary: \emph{doxa}, the belief production chain, epistemic capital, narrative infrastructure, and the doxonomic system as a formal object.
You can identify the core units of analysis (funding, institutions, narratives, exposure, belief) and trace them through a production chain.
Part~II develops these concepts into formal models.
\end{interlude}

% --- Part II: Concepts and Models ---
\part{Concepts and Models}
\label{part:concepts}

\input{chapters/ch06-epistemic-capital}
\input{chapters/ch07-narrative-infrastructure}
\input{chapters/ch08-common-sense-capture}
\input{chapters/ch09-doxonomic-power}
\input{chapters/ch10-economics-of-belief-production}
\input{chapters/ch11-formal-models}
\input{chapters/ch12-contradiction-and-coherence}

\begin{interlude}
You now have the conceptual toolkit: epistemic capital, narrative infrastructure, common-sense capture, doxonomic power, belief markets, dynamical models, and the sheaf-theoretic framework for coherence and contradiction.
These tools let you \emph{model} doxonomic systems formally.
Part~III asks: how would you \emph{measure} them empirically?
The methods chapters are the scientific backbone of the book: where the framework meets data.
\end{interlude}

% --- Part III: Methods ---
\part{Methods}
\label{part:methods}

\input{chapters/ch13-doxometric-measurement}
\input{chapters/ch14-belief-flow-accounting}
\input{chapters/ch15-network-analysis}
\input{chapters/ch16-causal-inference}
\input{chapters/ch17-text-media-discourse}
\input{chapters/ch18-historical-archival}
\input{chapters/ch19-experimental-doxonomics}
\input{chapters/ch20-model-validation}

\begin{interlude}
You now have both the models and the methods.
You can define a doxonomic system, measure its components, trace causal effects, analyze texts, reconstruct histories, design experiments, and assess the robustness of your conclusions.
Part~IV applies this toolkit to the belief regimes that matter most: selfishness, meritocracy, consumerism, inequality, race, austerity, work, nation, technology, and crisis.
These chapters are case studies, not proofs.
The framework illuminates each case; each case tests the framework.
Where the fit is strong, it increases our confidence.
Where it is weak, it marks the boundaries of what doxonomics can explain.
\end{interlude}

% --- Part IV: Domains of Application ---
\part{Domains of Application}
\label{part:applications}

\input{chapters/ch21-human-nature-narrative}
\input{chapters/ch22-meritocracy}
\input{chapters/ch23-consumerism}
\input{chapters/ch24-inequality-legitimacy}
\input{chapters/ch25-race-coloniality-gender}
\input{chapters/ch26-austerity}
\input{chapters/ch27-work-discipline}
\input{chapters/ch28-nation-myth}
\input{chapters/ch29-technology-inevitability}
\input{chapters/ch30-crisis-doxonomics}

\begin{interlude}
The applications have shown what doxonomics can see: how institutional investment produces common sense in domain after domain.
But diagnosis without remedy is incomplete.
Part~V asks the forward-looking questions: how do belief regimes break, how can they be challenged, what would a more democratic belief ecology look like, what ethical constraints bind the field, and what remains to be discovered?
These chapters are the most speculative and the most important.
\end{interlude}

% --- Part V: Resistance, Repair, and Reconstruction ---
\part{Resistance, Repair, and Reconstruction}
\label{part:resistance}

\input{chapters/ch31-how-belief-regimes-fracture}
\input{chapters/ch32-counter-doxonomic-strategies}
\input{chapters/ch33-epistemic-democracy}
\input{chapters/ch34-better-belief-ecologies}
\input{chapters/ch35-ethics-of-doxonomics}
\input{chapters/ch36-toward-social-self-understanding}

% === Back Matter ===
\backmatter

% --- Appendices ---
\appendix

\input{appendices/appA-glossary}
\input{appendices/appB-math-notation}
\input{appendices/appC-data-sources}
\input{appendices/appD-survey-instruments}
\input{appendices/appE-network-templates}
\input{appendices/appF-datasets}
\input{appendices/appG-replication}
\input{appendices/appH-reading}

% --- Bibliography ---
\printbibliography[heading=bibintoc,title={References}]

% --- Index ---
\printindex

\end{document}
