\chapter{Formal Models of Doxonomic Systems}
\label{ch:formal-models}

\epigraph{A theory has only the alternative of being right or wrong. A model has a third possibility: it may be right, but irrelevant.}{Manfred Eigen}

\noindent
This chapter develops the mathematical core of doxonomics.
The previous chapters introduced concepts (belief stocks, narrative flows, thresholds, lock-in, capture) using a mixture of formal definitions and qualitative argument.
Here we build the dynamical models that give these concepts predictive structure.

Five model families are presented, each illuminating a different aspect of doxonomic dynamics: stock-and-flow models for aggregate belief trajectories, epidemiological models for narrative spread, Bayesian models for rational updating under asymmetric exposure, cascade models for sudden collective shifts, and competitive models for belief-versus-belief dynamics.
The chapter closes with a guide to choosing among them.

A warning: all models in this chapter are stylized.
They are designed to clarify mechanisms and generate testable predictions, not to reproduce empirical data with quantitative precision.
Calibrating these models to real doxonomic systems is a major open problem.
The Piatra Institute's computational playgrounds allow readers to explore each model's behavior under varying parameters.


\section{Stock-and-Flow Models}
\label{sec:ch11-stock-flow}

The simplest dynamical model treats belief as a stock that accumulates or decays depending on narrative inflow and natural erosion.

\begin{definition}[Belief Stock and Narrative Flow]
\label{def:stock-flow}
\index{stock-and-flow model}
The \emph{belief stock} $\belief{i}(t) \in [0,1]$ is the fraction of the population holding belief $i$ at time $t$.
The \emph{narrative flow} $\phi_i(t) \in \R_{\geq 0}$ is the rate at which narrative content supporting $i$ is produced and circulated.
The dynamics are:
\begin{equation}
\label{eq:stock-flow}
  \frac{d\belief{i}}{dt} = \underbrace{\alpha \cdot \phi_i(t) \cdot (1 - \belief{i}(t))}_{\text{institutional persuasion}} - \underbrace{\delta \cdot \belief{i}(t)}_{\text{natural decay}} + \underbrace{\beta \cdot \belief{i}(t)(1 - \belief{i}(t))}_{\text{social reinforcement}}
\end{equation}
where $\alpha$ is the persuasion rate (effectiveness of institutional narrative per unit flow), $\delta$ is the natural decay rate (belief erodes without reinforcement), and $\beta$ is the social reinforcement rate (peer effects amplify belief through social proof).
\end{definition}

Each term has a distinct doxonomic interpretation:
\begin{itemize}[nosep]
  \item The persuasion term $\alpha\phi(1-\belief{})$ captures institutional production.
  It is proportional to narrative flow $\phi$ and to the remaining persuadable fraction $(1-\belief{})$: as belief approaches 1, there are fewer people left to persuade.
  \item The decay term $-\delta\belief{}$ captures the natural erosion of belief absent reinforcement: memories fade, new generations arrive uninstructed, and contrary experience accumulates.
  \item The social reinforcement term $\beta\belief{}(1-\belief{})$ captures peer effects: holding a belief makes you more likely to transmit it to non-holders.
  This term is logistic, strongest when $\belief{} \approx 0.5$ and vanishing at the extremes.
\end{itemize}

\subsection{Stability Analysis}

\begin{proposition}[Steady States and Stability]
\label{prop:steady-states}
Setting $d\belief{i}/dt = 0$ and writing $B = \belief{i}$ for brevity, the steady states satisfy:
\[
  \alpha\phi(1-B) - \delta B + \beta B(1-B) = 0
\]
which is a quadratic in $B$:
\[
  -\beta B^2 + (\alpha\phi + \beta - \delta)B - \alpha\phi \cdot 0 = 0
\]
Rearranging: $\beta B^2 - (\alpha\phi + \beta - \delta)B - \alpha\phi = 0$.
The steady states are:
\begin{equation}
\label{eq:steady-states}
  B^* = \frac{(\alpha\phi + \beta - \delta) \pm \sqrt{(\alpha\phi + \beta - \delta)^2 + 4\beta\alpha\phi}}{2\beta}
\end{equation}
\end{proposition}

\begin{remark}[Bistability Condition]
\label{rem:bistability}
\index{bistability}
The discriminant $\Delta = (\alpha\phi + \beta - \delta)^2 + 4\beta\alpha\phi$ is always positive (since $\beta, \alpha, \phi \geq 0$), so two real roots always exist when $\beta > 0$.
However, only roots in $[0,1]$ are physically meaningful.
The system exhibits \emph{bistability} (two stable equilibria separated by an unstable threshold) when $\beta$ is sufficiently large relative to $\delta$.
Specifically, when $\beta > \delta$ and $\phi$ is in an intermediate range, there exist a low-belief stable equilibrium $B_L^*$ and a high-belief stable equilibrium $B_H^*$, separated by an unstable equilibrium $B_U^*$.
The basin of attraction of each stable equilibrium determines the system's response to perturbations.
\end{remark}

The stability of each equilibrium is determined by the sign of $f'(B^*)$ where $f(B) = \alpha\phi(1-B) - \delta B + \beta B(1-B)$.
We have:
\[
  f'(B) = -\alpha\phi - \delta + \beta(1 - 2B)
\]
An equilibrium $B^*$ is stable if $f'(B^*) < 0$ and unstable if $f'(B^*) > 0$.

\subsection{Bifurcation Under Changing Funding}

\begin{proposition}[Funding Bifurcation]
\label{prop:bifurcation}
\index{bifurcation!funding}
As the narrative flow $\phi$ increases from zero, the system undergoes a saddle-node bifurcation.
Below a critical $\phi^*$, only the low-belief equilibrium exists.
At $\phi = \phi^*$, a pair of new equilibria appears (one stable, one unstable).
Above $\phi^*$, the system is bistable: the high-belief equilibrium is reachable if the initial condition or a temporary perturbation pushes $B$ above the unstable threshold.
\end{proposition}

This has a direct doxonomic interpretation: there is a \emph{minimum funding level} below which no amount of social reinforcement can sustain a belief.
Above that level, the belief can be ``tipped'' into a self-sustaining high-adoption equilibrium by a temporary burst of investment.
Once tipped, the belief persists even if funding is subsequently reduced (as long as $\phi$ remains above the level needed to maintain the equilibrium, which may be much lower than the tipping level).

\begin{example}[Numerical Example]
\label{ex:stock-flow-numerical}
With $\alpha = 0.3$, $\beta = 0.5$, $\delta = 0.1$:
\begin{itemize}[nosep]
  \item At $\phi = 0$: only $B^* = 0$ is stable (belief dies without support).
  \item At $\phi = 0.1$: $B_L^* \approx 0.08$ (low equilibrium), $B_H^* \approx 0.87$ (high equilibrium).
  The unstable threshold is $B_U^* \approx 0.35$.
  \item At $\phi = 0.5$: $B_H^* \approx 0.95$ (near-complete adoption); the low equilibrium has vanished.
\end{itemize}
The critical finding: at $\phi = 0.1$, a temporary shock that pushes adoption above 35\% will cause the system to cascade to the high equilibrium, and stay there even after the shock ends.
This is the formal model of the tipping-point phenomenon from Chapter~\ref{ch:common-sense-capture}.
\end{example}


\section{Epidemiological Diffusion Models}
\label{sec:ch11-sir}

Belief spread shares structural features with disease transmission: an ``infected'' individual can transmit the belief to ``susceptible'' contacts, and exposure can come from both peer contact and institutional sources.

\begin{model}[SIR Model for Belief]
\label{mod:sir-belief}
\index{SIR model}
Partition the population into three compartments:
\begin{itemize}[nosep]
  \item $S(t)$: susceptible (have not encountered the narrative),
  \item $I(t)$: infected (actively hold and transmit the belief),
  \item $R(t)$: recovered (have encountered and rejected the belief, or lost interest).
\end{itemize}
With $S + I + R = 1$, the dynamics are:
\begin{align}
  \frac{dS}{dt} &= -\beta S I - \alpha \phi(t) S \label{eq:sir-s}\\
  \frac{dI}{dt} &= \beta S I + \alpha \phi(t) S - \gamma I \label{eq:sir-i}\\
  \frac{dR}{dt} &= \gamma I \label{eq:sir-r}
\end{align}
where $\beta$ is the peer transmission rate, $\alpha\phi(t)$ is the institutional transmission rate (from narrative infrastructure), and $\gamma$ is the recovery rate (skepticism, counter-narrative exposure, or simple loss of interest).
\end{model}

\begin{definition}[Basic Reproduction Number]
\label{def:R0}
\index{basic reproduction number}
The \emph{doxonomic reproduction number} is
\[
  R_0 = \frac{\beta + \alpha\phi}{\gamma}
\]
If $R_0 > 1$, the belief can spread epidemically from a small seed.
If $R_0 < 1$, each ``infected'' individual transmits to fewer than one new adopter on average, and the belief dies out.
\end{definition}

A crucial difference from epidemiology: institutions can maintain $\phi > 0$ indefinitely, sustaining beliefs that would otherwise die out.
A belief with $\beta/\gamma < 1$ (peer transmission alone is insufficient) can persist if $\alpha\phi$ makes up the difference.
This is the formal basis of the claim that some beliefs persist not because they are persuasive but because they are subsidized.

\subsection{Endemic Equilibrium}

\begin{proposition}[Endemic Belief]
\label{prop:endemic}
\index{endemic equilibrium}
When $R_0 > 1$ and institutional support is sustained ($\phi = \text{const}$), the belief reaches an \emph{endemic equilibrium} where a constant fraction of the population holds the belief:
\[
  I^* = 1 - \frac{1}{R_0} = 1 - \frac{\gamma}{\beta + \alpha\phi}
\]
This is the doxonomic analogue of an endemic disease: the belief is never universally held but never disappears, maintained by the combination of peer transmission and institutional support.
\end{proposition}

\begin{example}[Endemic vs.\ Epidemic Belief]
\label{ex:endemic}
Consider two scenarios:
\begin{itemize}[nosep]
  \item \textbf{Epidemic belief} (high $R_0$): a viral conspiracy theory with $\beta = 0.3$, $\gamma = 0.1$, $\alpha\phi = 0$.
  $R_0 = 3$; the belief sweeps through the susceptible population rapidly but then decays as the recovery rate takes effect.
  Peak $I$ is high but transient.
  \item \textbf{Endemic belief} (moderate $R_0$, sustained $\phi$): the belief ``markets are efficient'' with $\beta = 0.05$, $\gamma = 0.02$, $\alpha\phi = 0.03$.
  $R_0 = 4$; the belief reaches a stable endemic level $I^* = 0.75$ and stays there indefinitely.
  It never goes viral but never fades either.
\end{itemize}
The epidemic belief is dramatic but ephemeral.
The endemic belief is quiet but permanent.
Doxonomic systems aim for endemic stability, not epidemics.
\end{example}

\subsection{The SIRV Model: Inoculation}

\begin{model}[SIRV Model with Inoculation]
\label{mod:sirv}
\index{SIRV model}
\index{inoculation!model}
Adding a ``vaccinated'' compartment $V(t)$ representing people who have received counter-narrative inoculation (Chapter~\ref{ch:counter-doxonomic}):
\begin{align}
  \frac{dS}{dt} &= -\beta S I - \alpha \phi S - \nu S \\
  \frac{dV}{dt} &= \nu S - \eta V I \\
  \frac{dI}{dt} &= \beta S I + \alpha \phi S + \eta V I - \gamma I \\
  \frac{dR}{dt} &= \gamma I
\end{align}
where $\nu$ is the inoculation rate and $\eta \ll \beta$ is the reduced transmission rate for inoculated individuals ($\eta < \beta$ because inoculation provides partial immunity but is not perfect).

The effective reproduction number with inoculation is:
\[
  R_{\text{eff}} = \frac{(\beta + \alpha\phi) S^* + \eta V^*}{\gamma}
\]
Inoculation reduces $R_{\text{eff}}$ by moving people from $S$ to $V$, where they are harder to infect.
\end{model}

\begin{remark}[Doxonomic Herd Immunity]
\label{rem:herd-immunity}
\index{herd immunity!doxonomic}
If inoculation is sufficiently widespread, $R_{\text{eff}}$ drops below 1 and the belief cannot sustain itself even with institutional support.
The \emph{doxonomic herd immunity threshold} is the fraction of the population that must be inoculated to prevent endemic belief:
\[
  V^* > 1 - \frac{1}{R_0}
\]
This provides a quantitative target for counter-doxonomic educational programs: inoculate more than $1 - 1/R_0$ of the population, and the belief cannot persist at endemic levels.
\end{remark}


\section{Bayesian Updating Under Unequal Exposure}
\label{sec:ch11-bayesian}

The epidemiological models treat belief adoption as binary (hold or don't hold).
The Bayesian framework models belief as a continuous credence that evolves as agents receive signals from their information environment.

\begin{model}[Asymmetric Information Environment]
\label{mod:asymmetric-bayes}
\index{Bayesian updating!asymmetric}
Agent $j$ observes a sequence of signals $\{s_t^{(j)}\}_{t=1}^{T}$ where each $s_t^{(j)}$ is drawn from
\[
  s_t^{(j)} \sim \begin{cases} \nu^+_j & \text{with probability } q_j \\ \nu^-_j & \text{with probability } 1 - q_j \end{cases}
\]
and $q_j$ depends on $j$'s media environment.
Agents in channel $k$ have $q_j = q_k$, determined by the channel's editorial stance.
The parameter $q_k$ captures the \emph{exposure bias} of channel $k$: $q_k > 1/2$ means the channel favors proposition $p$; $q_k < 1/2$ means it favors $\neg p$; $q_k = 1/2$ means balanced exposure.
\end{model}

\begin{theorem}[Polarization Under Asymmetric Exposure]
\label{thm:polarization}
\index{polarization}
If two groups $A$ and $B$ have $q_A > 1/2$ and $q_B < 1/2$ (their channels systematically favor opposite conclusions), then as $T \to \infty$:
\[
  \pi_A^{(T)}(p) \to 1 \quad \text{and} \quad \pi_B^{(T)}(p) \to 0
\]
almost surely, even if both groups are perfectly rational Bayesian updaters starting from the same prior.
\end{theorem}

\begin{proof}
For group $A$, the log-likelihood ratio after $T$ signals is:
\[
  \Lambda_A^{(T)} = \sum_{t=1}^{T} \log \frac{L(s_t \mid p)}{L(s_t \mid \neg p)}
\]
Since $q_A > 1/2$, the expected log-likelihood ratio per signal is:
\[
  \E[\log L(s_t \mid p) / L(s_t \mid \neg p)] = q_A \log\frac{q_A}{1-q_A} + (1-q_A)\log\frac{1-q_A}{q_A} = D_{\mathrm{KL}}(q_A \| 1-q_A) > 0
\]
By the strong law of large numbers, $\Lambda_A^{(T)} / T \to D_{\mathrm{KL}}(q_A \| 1-q_A) > 0$ almost surely, so $\Lambda_A^{(T)} \to \infty$ and $\pi_A^{(T)}(p) \to 1$.
The symmetric argument applies to group $B$ with $q_B < 1/2$.
The divergence rate is proportional to $D_{\mathrm{KL}}(q_A \| 1-q_A) + D_{\mathrm{KL}}(q_B \| 1-q_B)$.
\end{proof}

\begin{remark}
This theorem is the formal foundation for the claim that doxonomic polarization does not require irrationality, only asymmetric evidence supply.
Two perfectly rational populations in different information environments will arrive at opposite conclusions with certainty.
The doxonomic system controls $q_k$ (the exposure bias of each channel); the rest is Bayes' rule doing its job.
\end{remark}

\begin{example}[Numerical Polarization]
\label{ex:polarization-numerical}
Two groups start with the same prior $\pi^{(0)} = 0.5$.
Group $A$ watches Channel X ($q_A = 0.7$): 70\% of signals support $p$.
Group $B$ watches Channel Y ($q_B = 0.3$): only 30\% of signals support $p$.
After $T = 50$ signals:
\begin{itemize}[nosep]
  \item Group $A$: expected $\pi_A^{(50)} \approx 0.98$ (near-certain $p$ is true).
  \item Group $B$: expected $\pi_B^{(50)} \approx 0.02$ (near-certain $p$ is false).
\end{itemize}
Both groups are fully rational.
Both have seen 50 pieces of ``evidence.''
They disagree completely because they live in different information environments.
If asked, each would say the other group is ``ignoring the evidence.'' From their perspective, they would be right.
This is doxonomic polarization in its purest form.
\end{example}

\begin{remark}[Correlated Signals and Bounded Rationality]
\label{rem:correlated}
The theorem assumes conditionally independent signals.
In practice, media content is highly correlated (outlets repeat each other, social media creates echo chambers, institutional sources are shared).
Correlated signals carry less information per signal than independent ones, which means:
\begin{itemize}[nosep]
  \item the \emph{perceived} evidence is stronger than the \emph{actual} evidence (agents treat correlated signals as independent, over-counting),
  \item the polarization occurs faster than the theorem predicts (because agents are over-confident in their media-derived beliefs),
  \item de-polarization requires not just exposure to counter-signals but recognition that prior signals were correlated, a meta-cognitive step that most agents do not take spontaneously.
\end{itemize}
This is one reason why ``just expose people to the other side'' is a weaker intervention than doxonomic analysis might naively suggest.
The problem is not insufficient exposure; it is the structure of the prior evidence environment.
\end{remark}


\section{Cascade Models}
\label{sec:ch11-cascades}

\begin{model}[Information Cascade]
\label{mod:cascade-threshold}
\index{information cascade}
Following \textcite{banerjee1992} and \textcite{bikhchandani1992}, agents act sequentially, observing predecessors' actions but not their private signals.
Each agent $j$ receives a private signal $s_j \in \{+, -\}$ (favoring $p$ or $\neg p$) and observes the actions $a_1, \ldots, a_{j-1}$ of all predecessors.
Agent $j$'s optimal action is:
\[
  a_j = \begin{cases} + & \text{if } \pi_j(p \mid s_j, a_1, \ldots, a_{j-1}) > 1/2 \\ - & \text{otherwise} \end{cases}
\]
An \emph{information cascade} occurs when the accumulated public evidence from predecessors' actions is so strong that agent $j$'s optimal action is independent of $j$'s private signal: the public evidence overwhelms private information.
\end{model}

\begin{proposition}[Cascade Fragility]
\label{prop:cascade-fragility}
\index{information cascade!fragility}
Information cascades are \emph{informationally fragile}: they aggregate very little information.
In a cascade, each agent's action reveals no new information (because the action is independent of the private signal).
Therefore, the cascade is based on the signals of the few agents who acted before the cascade began, potentially as few as two or three.
A single credible revelation of contradictory information can shatter the cascade.
\end{proposition}

\begin{remark}[Doxonomic Interpretation]
In doxonomic terms, a cascade occurs when the accumulated weight of institutional narrative is so strong that individual experience becomes irrelevant to expressed belief.
Everyone acts as if $p$ is true, not because the evidence is overwhelming, but because everyone else is acting as if $p$ is true.
This is precisely the pluralistic-ignorance configuration (Definition~\ref{def:pluralistic-ignorance}).

The fragility result (Proposition~\ref{prop:cascade-fragility}) has an important implication for regime fracture (Chapter~\ref{ch:how-belief-regimes-fracture}): cascade-based consensus is dramatically more vulnerable to disruption than evidence-based consensus.
A single whistleblower, a single scandal, or a single credible defection can trigger a \emph{reverse cascade} (a rapid collapse of apparent consensus, as each agent, seeing others defect, reveals their own private doubt).
This is the mechanism behind ``preference cascades'' and the ``element of surprise in revolutions'' \autocite{kuran1995}.
\end{remark}


\section{Competitive Belief Dynamics}
\label{sec:ch11-competitive}

Most real doxonomic systems involve not one belief spreading through a population but multiple beliefs competing for the same audience.

\begin{model}[Lotka-Volterra Belief Competition]
\label{mod:lotka-volterra}
\index{Lotka-Volterra model}
Let $\belief{1}(t)$ and $\belief{2}(t)$ be two competing beliefs, not necessarily summing to 1 (an agent can hold neither or both to some degree).
The dynamics are:
\begin{align}
  \frac{d\belief{1}}{dt} &= \belief{1}(\alpha_1 \phi_1 - \gamma_1 - \beta_{12}\belief{2}) \label{eq:lv1}\\
  \frac{d\belief{2}}{dt} &= \belief{2}(\alpha_2 \phi_2 - \gamma_2 - \beta_{21}\belief{1}) \label{eq:lv2}
\end{align}
where $\phi_k$ is the narrative flow for belief $k$, $\gamma_k$ is the intrinsic decay rate, and $\beta_{ij}$ is the competitive displacement rate (the degree to which belief $j$ suppresses belief $i$).
\end{model}

\subsection{Equilibrium Analysis}

The system has four possible equilibria:
\begin{enumerate}[nosep]
  \item $(0, 0)$: both beliefs extinct.
  Always exists; stable when both $\alpha_k\phi_k < \gamma_k$.
  \item $(\belief{1}^*, 0)$: belief 1 dominates, belief 2 extinct.
  $\belief{1}^* = (\alpha_1\phi_1 - \gamma_1)/\beta_{12}$ if positive.
  \item $(0, \belief{2}^*)$: belief 2 dominates, belief 1 extinct.
  \item $(\belief{1}^*, \belief{2}^*)$: coexistence.
\end{enumerate}

\begin{proposition}[Competitive Exclusion]
\label{prop:competitive-exclusion}
\index{competitive exclusion}
If $\beta_{12}\beta_{21} > (\alpha_1\phi_1 - \gamma_1)(\alpha_2\phi_2 - \gamma_2) / (\belief{1}^*\belief{2}^*)$, the coexistence equilibrium is unstable: one belief must exclude the other.
The winner depends on initial conditions and funding levels.
This is the doxonomic analogue of the \emph{competitive exclusion principle} in ecology: two beliefs competing for the same ``niche'' (the same proposition space) cannot coexist indefinitely unless they differentiate.
\end{proposition}

\begin{proposition}[Funding Advantage and Dominance]
\label{prop:funding-advantage}
\index{funding advantage}
If beliefs 1 and 2 have identical intrinsic properties ($\gamma_1 = \gamma_2$, $\beta_{12} = \beta_{21}$) but asymmetric funding ($\phi_1 > \phi_2$), then belief 1 dominates in the long run.
Specifically, the critical condition for belief 1 to exclude belief 2 is:
\[
  \alpha_1\phi_1 - \gamma_1 > \beta_{21} \cdot \frac{\alpha_2\phi_2 - \gamma_2}{\beta_{12}}
\]
When the beliefs are symmetric in all respects except funding, this reduces to $\phi_1 > \phi_2$: the better-funded belief wins, regardless of intrinsic merit.
\end{proposition}

\begin{remark}[When the Under-Funded Belief Survives]
\label{rem:underfunded-survival}
The under-funded belief ($\phi_2 < \phi_1$) can survive or even win if:
\begin{itemize}[nosep]
  \item it has lower intrinsic decay ($\gamma_2 < \gamma_1$), meaning it is more consistent with lived experience and requires less maintenance,
  \item it has stronger competitive displacement ($\beta_{21} > \beta_{12}$), meaning it is more effective at undermining the dominant belief per unit of contact,
  \item it occupies a different niche: rather than directly contradicting belief 1, it addresses a different aspect of the proposition space, enabling coexistence.
\end{itemize}
These conditions provide guidance for counter-doxonomic strategy (Chapter~\ref{ch:counter-doxonomic}): when outspent, invest in resilience ($\gamma$ reduction), efficiency ($\beta$ enhancement), and niche differentiation rather than trying to match the dominant belief's funding.
\end{remark}


\section{Agent-Based Models}
\label{sec:ch11-abm}

When analytical models become intractable (because of heterogeneous agents, network structure, multiple interacting beliefs, or complex institutional dynamics), agent-based models (ABMs) provide a computational alternative.

\begin{model}[Doxonomic Agent-Based Model]
\label{mod:abm}
\index{agent-based model}
A doxonomic ABM consists of:
\begin{itemize}[nosep]
  \item $N$ agents on a network $G = (V, E)$, each with a belief vector $\bm{\pi}_j \in [0,1]^n$ and attributes (group membership, resistance profile, media exposure),
  \item $m$ institutional agents producing narrative at rates determined by funding,
  \item an exposure rule: at each time step, agent $j$ is exposed to narratives from their network neighbors and from institutional channels,
  \item an update rule: agent $j$ updates $\bm{\pi}_j$ according to the augmented internalization model~\eqref{eq:augmented-update} from Chapter~\ref{ch:belief-production-chain},
  \item a feedback rule: aggregate beliefs influence institutional funding (closing the loop).
\end{itemize}
\end{model}

ABMs sacrifice analytical tractability for realism.
They allow exploration of phenomena that are difficult to capture in closed-form models:
\begin{itemize}[nosep]
  \item \textbf{network effects}: beliefs spread differently on scale-free networks (hubs amplify) versus random networks (uniform diffusion),
  \item \textbf{community structure}: polarization emerges naturally from clustered networks where intra-group connections are denser than inter-group connections,
  \item \textbf{heterogeneous resistance}: different population segments respond differently to the same narrative, and the aggregate outcome depends on the distribution of resistance profiles,
  \item \textbf{strategic institutional behavior}: funders can adapt their strategies in response to observed belief dynamics, creating complex co-evolutionary dynamics.
\end{itemize}

\begin{remark}[ABMs and the Playgrounds]
\label{rem:playgrounds}
The Piatra Institute's computational playgrounds implement several doxonomic ABMs with adjustable parameters.
Readers can explore: how network topology affects polarization speed, how funding asymmetry determines competitive outcomes, how inoculation strategies perform against different epidemic dynamics, and how threshold distributions shape the conditions for capture.
These simulations are designed to complement the analytical models of this chapter, not to replace them.
\end{remark}


\section{Model Comparison and Selection}
\label{sec:ch11-comparison}

With five model families available, the researcher must choose which is appropriate for a given doxonomic question.
The following table provides guidance:

\begin{center}
\begin{tabular}{p{3cm}p{3.5cm}p{3.5cm}p{3.5cm}}
\toprule
\textbf{Model} & \textbf{Best for} & \textbf{Key assumption} & \textbf{Limitation} \\
\midrule
Stock-and-flow & Aggregate belief trajectories, tipping points, bifurcation analysis & Homogeneous population, well-mixed & No network structure, no individual heterogeneity \\
SIR/SIRV & Narrative spread dynamics, endemic equilibria, inoculation strategy & Compartmental (agents are S, I, or R) & Binary belief, no credence gradients \\
Bayesian & Polarization, filter bubbles, rational divergence under asymmetric exposure & Rational agents, independent signals & Ignores motivated reasoning, identity, affect \\
Cascade & Sudden collective shifts, fragile consensus, regime collapse & Sequential action, observable predecessors & Stylized; hard to calibrate to real populations \\
Lotka-Volterra & Belief competition, competitive exclusion, counter-narrative resilience & Competing beliefs interact through displacement & Aggregate; no within-population structure \\
ABM & All of the above with heterogeneity and network structure & Computational; requires parameter choices & Not analytically tractable; results are simulation-specific \\
\bottomrule
\end{tabular}
\end{center}

\begin{principle}[Model Selection]
\label{prin:model-selection}
\index{model selection}
Choose the simplest model that captures the mechanism under study.
Use stock-and-flow for aggregate dynamics, SIR for spread phenomena, Bayesian for polarization, Lotka-Volterra for competition, cascades for sudden shifts, and ABMs when heterogeneity and network structure are essential.
When in doubt, start with the analytical model and move to ABM only when the analytical model's assumptions are clearly violated by the empirical setting.
\end{principle}


\section{Notes and References}
\label{sec:ch11-notes}

Stock-and-flow models adapt systems dynamics; for an introduction, see \textcite{sterman2000}.
The SIR framework for information diffusion is developed in \textcite{watts2002}; for a comprehensive treatment of epidemic models on networks, see \textcite{pastor-satorras2015}.
On the distinction between simple and complex contagion (behaviors that require multiple reinforcing contacts to spread), see \textcite{centola2007, centola2018}.
Continuous-state opinion-dynamics models (the closest neighbor to the stock-and-flow and Bayesian families) are surveyed in \textcite{degroot1974} and \textcite{hegselmann2002}.
Bayesian polarization under asymmetric information is treated in the economics of media bias \autocite{gentzkow2010}; on rational herding, see \textcite{banerjee1992}.
Cascade models originate with \textcite{banerjee1992} and \textcite{bikhchandani1992}; on cascade fragility, see also \textcite{chamley2004}.
Lotka-Volterra models for cultural competition appear in the cultural evolution literature \autocite{nowak2006}; the competitive exclusion principle is standard in mathematical ecology.
Agent-based models in social science are surveyed in \textcite{salganik2018} and \textcite{epstein2006}.
Preference cascades and the element of surprise in regime collapse draw on \textcite{kuran1995}.
Bifurcation theory for social tipping points connects to \textcite{schelling1978}.
On the relationship between signal correlation and belief over-confidence, see \textcite{glaeser2009}.
On the fragility of Bayesian agreement when agents disagree about the likelihoods of signals (not only about priors), see \textcite{acemoglu2016}.


\section{Exercises}

\begin{exercise}
For the stock-and-flow model~\eqref{eq:stock-flow} with $\alpha = 0.3$, $\delta = 0.1$, $\beta = 0.5$, find the minimum narrative flow $\phi^*$ needed for a stable high-adoption equilibrium to exist.
Verify your answer by computing the steady states at $\phi = \phi^* - 0.01$ and $\phi = \phi^* + 0.01$.
\end{exercise}

\begin{exercise}
Compute $R_0$ for a belief spread through social media ($\beta = 0.05$, $\gamma = 0.2$) with institutional support $\alpha\phi = 0.1$.
Can the belief sustain itself without institutional support ($\phi = 0$)?
What is the endemic equilibrium $I^*$ with and without institutional support?
\end{exercise}

\begin{exercise}
Simulate the Bayesian polarization theorem: generate $T = 100$ signals for two groups with $q_A = 0.7$ and $q_B = 0.3$, starting from the same prior $\pi^{(0)} = 0.5$.
Plot the posterior trajectories.
Now repeat with correlated signals (each signal is a copy of the previous one with probability 0.5).
How does correlation affect the speed and confidence of polarization?
\end{exercise}

\begin{exercise}
In the Lotka-Volterra model with $\gamma_1 = \gamma_2 = 0.1$, $\beta_{12} = \beta_{21} = 0.3$, and $\alpha_1 = \alpha_2 = 0.2$, find the critical funding ratio $\phi_1 / \phi_2$ above which belief 1 excludes belief 2.
Interpret: how much must the dominant belief outspend the challenger to guarantee competitive exclusion?
\end{exercise}

\begin{exercise}
Extend the SIR model to include a ``vaccinated'' compartment $V(t)$ using Model~\ref{mod:sirv}.
If $R_0 = 3$, what fraction of the population must be inoculated to achieve doxonomic herd immunity?
What inoculation rate $\nu$ achieves this within 5 years?
\end{exercise}

\begin{exercise}
\label{ex:ch11-cascade-fragility}
In a cascade of 20 agents, suppose agent 15 receives a highly credible counter-signal (a whistleblower revelation).
Model the reverse cascade: if agent 15 defects, under what conditions do agents 16--20 follow?
Under what conditions does the cascade withstand the shock?
\end{exercise}

\begin{exercise}
\label{ex:ch11-abm}
Design a doxonomic ABM with 1000 agents on a small-world network.
Specify: two competing beliefs, one institutional funder for each, heterogeneous resistance profiles, and a feedback rule.
Describe (qualitatively) what you expect to observe as you vary the funding ratio $\phi_1/\phi_2$ from 1:1 to 10:1.
If you have access to a simulation environment, implement and run the model.
\end{exercise}

\begin{exercise}
The model comparison table provides guidance on when to use each model.
For each of the following doxonomic questions, identify the most appropriate model family and justify your choice:
(a)~How does Fox News affect climate belief in the U.S.?
(b)~Can a counter-narrative survive against a 5:1 funding disadvantage?
(c)~Why did public support for austerity collapse so quickly in some countries after 2008?
(d)~What is the minimum inoculation rate to prevent anti-vaccine belief from reaching endemic levels?
\end{exercise}
