\chapter{Network Analysis of Belief Systems}
\label{ch:network-analysis}

\epigraph{Tell me who your friends are, and I will tell you who you are.}{Proverb}

\noindent
Doxonomic systems are networks: funders connect to institutions, institutions connect to media, media connect to audiences, and audiences feed back into funders.
The structure of these connections (who is linked to whom, with what weight, through what pathways) determines how narratives flow, where they amplify, and which nodes hold structural power.

This chapter applies graph theory and network science to map and measure doxonomic structures.
Where previous chapters treated the belief production chain as a linear sequence, this chapter reveals it as a complex, multiply-connected network with shortcuts, bottlenecks, clusters, and bridges.


\section{Doxonomic Networks}
\label{sec:ch15-networks}

\begin{definition}[Doxonomic Network]
\label{def:doxonomic-network}
\index{doxonomic network}
A \emph{doxonomic network} is a weighted directed graph $G = (V, E, w)$ where:
\begin{itemize}[nosep]
  \item vertices $V$ are doxonomic agents (funders, institutions, media, intermediaries, audience segments),
  \item directed edges $E$ represent influence or resource flows (money, citations, sourcing, audience sharing),
  \item weights $w: E \to \R_{\geq 0}$ measure the magnitude of the flow (dollars, citation counts, audience share fraction).
\end{itemize}
\end{definition}

A doxonomic network is typically \emph{multipartite}: vertices can be classified into layers (funders → institutions → channels → audiences), with most edges running between layers.
But the network is not strictly layered: there are important within-layer connections (institutions cite each other, media outlets share content, funders coordinate) and feedback edges (audience behavior influences funders).

\begin{definition}[Network Types]
\label{def:network-types}
\index{doxonomic network!types}
Several network projections are useful for different analytical purposes:
\begin{description}[nosep]
  \item[Funding network:] $G_F = (V_{\text{funders}} \cup V_{\text{institutions}}, E_F)$ where edges represent grant flows.
  Source: IRS 990 data, foundation annual reports.
  \item[Citation network:] $G_C = (V_{\text{institutions}} \cup V_{\text{media}}, E_C)$ where edges represent citation or sourcing relationships.
  Source: content analysis, media sourcing studies.
  \item[Personnel network:] $G_P = (V, E_P)$ where edges represent personnel movement (a person leaving institution $A$ to join institution $B$).
  Source: LinkedIn data, biographical records.
  \item[Board interlock:] $G_B = (V_{\text{institutions}}, E_B)$ where edges connect institutions sharing board members.
  Source: organizational filings, databases like BoardEx.
  \item[Audience overlap:] $G_A = (V_{\text{channels}}, E_A)$ where edges represent shared audience (people who consume both channels).
  Source: survey data, platform analytics.
\end{description}
\end{definition}

Each projection reveals a different aspect of the doxonomic system.
The funding network shows who pays for what.
The citation network shows how ideas propagate.
The personnel network shows how institutional cultures diffuse.
The board-interlock network shows structural coordination.
The audience-overlap network shows which channels reach the same people.

\begin{remark}[Multiplex Analysis]
\label{rem:multiplex}
\index{multiplex network}
In practice, these networks overlap: institution $A$ may be funded by funder $X$ (funding edge), cited by media outlet $Y$ (citation edge), share a board member with institution $B$ (interlock edge), and hire graduates of institution $C$ (personnel edge).
The multiplex structure (multiple relationship types between the same nodes) is what makes doxonomic networks so robust: disrupting one type of connection leaves others intact.
A funder who withdraws funding may still be connected through board interlocks and personnel sharing.
\end{remark}


\section{Centrality Measures}
\label{sec:ch15-centrality}

Not all nodes in a doxonomic network are equally important.
Centrality measures identify the nodes that occupy structurally powerful positions.

\begin{definition}[Doxonomic Centrality]
\label{def:centrality}
\index{centrality}
The \emph{doxonomic centrality} of node $v$ is its eigenvector centrality in the weighted network:
\[
  x_v = \frac{1}{\lambda} \sum_{u: (u,v) \in E} w_{uv} \cdot x_u
\]
where $\lambda$ is the largest eigenvalue of the weighted adjacency matrix.
Nodes with high doxonomic centrality are those that receive large weighted inputs from other central nodes, a recursive definition that captures the idea that influence from influential sources is worth more than influence from peripheral ones.
\end{definition}

Eigenvector centrality is closely related to PageRank (used by Google to rank web pages) and has a natural doxonomic interpretation: a node is central if it is cited, funded, or sourced by other central nodes.
This captures the credibility-laundering dynamic (Example~\ref{ex:think-tank-credibility}): an institution becomes credible by being associated with other credible institutions.

\begin{example}[Centrality Hierarchy]
\label{ex:centrality-hierarchy}
In the U.S.\ policy-influencing network, typical centrality rankings are:
\begin{enumerate}[nosep]
  \item \textbf{Highest}: major universities (Harvard, MIT, Stanford, Chicago), central through personnel, citation, and prestige networks simultaneously.
  \item \textbf{High}: legacy think tanks (Brookings, Heritage, AEI, RAND), central through funding, citation, and congressional testimony networks.
  \item \textbf{Medium}: prestige media (\textit{NYT}, \textit{WSJ}, \textit{Washington Post}), central through citation and audience networks.
  \item \textbf{Lower}: newer think tanks, advocacy organizations, connected through fewer network layers.
  \item \textbf{Peripheral}: grassroots organizations, independent media, weakly connected to the core.
\end{enumerate}
The centrality hierarchy is remarkably stable over time: the top-ranked institutions in 2020 were substantially the same as in 1990, reflecting the path dependence of epistemic capital accumulation (Proposition~\ref{prop:rent-persistence}).
\end{example}

\subsection{Multiple Centrality Measures}

Different centrality measures capture different forms of structural importance:

\begin{center}
\begin{tabular}{lp{5cm}p{5cm}}
\toprule
\textbf{Measure} & \textbf{What it captures} & \textbf{Doxonomic interpretation} \\
\midrule
Eigenvector & Connected to well-connected nodes & Prestige through association \\
Betweenness & Lies on many shortest paths & Brokerage power \\
In-degree & Many incoming edges & Attention/funding recipient \\
Out-degree & Many outgoing edges & Narrative producer/distributor \\
Closeness & Short distance to all others & Speed of narrative propagation \\
PageRank & Important in directed flow & Influence in directed networks \\
\bottomrule
\end{tabular}
\end{center}

A node may rank high on one measure and low on another.
A think tank may have high betweenness (it bridges funders and media) but low in-degree (it has few funders).
A social media platform may have enormous out-degree (it reaches billions) but low eigenvector centrality (it is not prestigious).
The appropriate centrality measure depends on the doxonomic question being asked.


\section{Community Detection}
\label{sec:ch15-community}

\begin{definition}[Narrative Cluster]
\label{def:narrative-cluster}
\index{narrative cluster}
A \emph{narrative cluster} is a community in the doxonomic network: a subset $S \subset V$ with denser internal connections than external ones.
Formally, $S$ maximizes the modularity:
\[
  Q = \frac{1}{2W} \sum_{i,j \in S} \left(w_{ij} - \frac{s_i s_j}{2W}\right)
\]
where $s_i = \sum_j w_{ij}$ is the weighted degree of node $i$ and $W = \frac{1}{2}\sum_{ij} w_{ij}$ is the total weight.
High modularity ($Q > 0.3$) indicates well-separated clusters.
\end{definition}

Narrative clusters reveal which institutions form coherent ideological coalitions, even without explicit coordination.
Institutions within the same cluster share funders, cite each other's work, hire each other's personnel, and tend to produce narratives within a similar ideological range.
The cluster structure emerges from \emph{structural alignment} (Principle~\ref{prin:structural-alignment}), not from conspiracy.

\begin{example}[Two-Cluster Structure in U.S.\ Policy]
\label{ex:two-cluster}
\index{narrative cluster!example}
\textcite{benkler2018} mapped the U.S.\ media and think-tank ecosystem during the 2016 election and found a striking asymmetric structure:
\begin{itemize}[nosep]
  \item A \textbf{right-wing cluster} centered on Breitbart, Fox News, and talk radio, with dense internal citation but sparse links to other media.
  This cluster shared a distinct funding network (Mercer family, Koch network), a distinct personnel pool, and a distinct audience.
  \item A \textbf{center-to-left cluster} with the \textit{New York Times} at its center, connected to a broader range of outlets including \textit{Washington Post}, CNN, BBC, and academic institutions.
  This cluster had higher internal citation diversity and more external connections.
\end{itemize}
The asymmetry is doxonomically significant: the right-wing cluster is more insular (higher intra-cluster density, lower inter-cluster connections), which means it can sustain narratives that diverge from the center cluster without corrective pressure from shared sources.
This is the network-structural basis of epistemic fragmentation (Definition~\ref{def:fragmentation}).
\end{example}


\section{Brokerage}
\label{sec:ch15-brokerage}

\begin{definition}[Epistemic Broker]
\label{def:broker}
\index{epistemic broker}
An \emph{epistemic broker} is a node $v$ with high betweenness centrality:
\[
  g(v) = \sum_{s \neq v \neq t} \frac{\sigma_{st}(v)}{\sigma_{st}}
\]
where $\sigma_{st}$ is the number of shortest paths from $s$ to $t$ and $\sigma_{st}(v)$ is the number passing through $v$.
Brokers bridge otherwise disconnected narrative clusters and can translate ideas between communities.
\end{definition}

Brokers occupy a special doxonomic position: they have access to multiple narrative worlds and can control the flow of ideas between them.
A journalist who covers both conservative think tanks and progressive policy organizations is an epistemic broker.
An academic who publishes in both economics journals and sociology journals bridges disciplinary clusters.

\begin{proposition}[Broker Power]
\label{prop:broker-power}
\index{broker power}
Epistemic brokers derive power from three sources:
\begin{enumerate}[nosep]
  \item \textbf{Translation}: they can reframe ideas from one cluster in the language of another, making them legible and credible across boundaries.
  \item \textbf{Gatekeeping}: they control which ideas cross cluster boundaries and which are blocked.
  \item \textbf{Arbitrage}: they can exploit information asymmetries between clusters, introducing ideas that are common knowledge in one cluster as ``new insights'' in another.
\end{enumerate}
\end{proposition}

\begin{example}[Brokerage in the Neoliberal Network]
\label{ex:brokerage-neoliberal}
Milton Friedman functioned as an epistemic broker between academic economics, public media, and political policy.
He translated technical economic arguments into popular columns (\textit{Newsweek}), television programs (\textit{Free to Choose}), and policy advice (to Pinochet, Reagan, Thatcher).
His brokerage position (connecting the academic cluster to the media cluster to the political cluster) was arguably more consequential than his academic contributions alone, because it enabled ideas that were contentious within economics to become common sense in policy.
\end{example}


\section{Example: A Donor-Institution Network}
\label{sec:ch15-example}

\begin{center}
\begin{tikzpicture}[
  funder/.style={circle, draw, thick, fill=doxoblue!20, minimum size=1cm, font=\footnotesize},
  inst/.style={circle, draw, thick, fill=doxogreen!20, minimum size=1cm, font=\footnotesize},
  media/.style={circle, draw, thick, fill=doxored!20, minimum size=1cm, font=\footnotesize},
  edge/.style={-Stealth, thick},
]
  \node[funder] (f1) at (0,2) {$F_1$};
  \node[funder] (f2) at (0,0) {$F_2$};
  \node[inst] (i1) at (3,3) {$I_1$};
  \node[inst] (i2) at (3,1) {$I_2$};
  \node[inst] (i3) at (3,-1) {$I_3$};
  \node[media] (m1) at (6,2) {$M_1$};
  \node[media] (m2) at (6,0) {$M_2$};

  \draw[edge] (f1) -- node[above, font=\scriptsize] {\$3M} (i1);
  \draw[edge] (f1) -- node[above, font=\scriptsize] {\$1M} (i2);
  \draw[edge] (f2) -- node[below, font=\scriptsize] {\$2M} (i2);
  \draw[edge] (f2) -- node[below, font=\scriptsize] {\$4M} (i3);
  \draw[edge] (i1) -- (m1);
  \draw[edge] (i2) -- (m1);
  \draw[edge] (i2) -- (m2);
  \draw[edge] (i3) -- (m2);
\end{tikzpicture}
\end{center}

In this network, institution $I_2$ is the epistemic broker: it bridges both funders and both media outlets.
It receives funding from both $F_1$ and $F_2$ and feeds into both $M_1$ and $M_2$.
Its betweenness centrality is the highest in the network: removing $I_2$ would disconnect $F_1$ from $M_2$ and $F_2$ from $M_1$.

\begin{remark}[Interpreting the Diagram]
\label{rem:interpreting-diagram}
Several doxonomic insights emerge from even this simple network:
\begin{itemize}[nosep]
  \item $F_2$ has more total funding (\$6M) but $F_1$'s funding reaches a higher-credibility media outlet ($M_1$) through the more efficient pathway $F_1 \to I_1 \to M_1$.
  \item $I_2$ is the most structurally powerful institution despite receiving less funding than $I_3$, because of its bridging position.
  \item If $I_2$ is captured by one funder (say $F_2$ doubles its grant while $F_1$ withdraws), the narrative output of $M_1$ shifts, even though $M_1$ receives no direct funding from $F_2$.
  This is \emph{indirect influence}: controlling the broker controls the bridge.
\end{itemize}
\end{remark}


\section{Spectral Analysis of Doxonomic Networks}
\label{sec:ch15-spectral}

The eigenvalue spectrum of a doxonomic network reveals its structural properties.

\begin{definition}[Narrative Propagation Speed]
\label{def:propagation-speed}
\index{propagation speed}
The \emph{narrative propagation speed} in a network is determined by the spectral radius $\lambda_1$ of the weighted adjacency matrix $\bm{W}$:
\[
  \lambda_1 = \max\{|\lambda| : \det(\bm{W} - \lambda \bm{I}) = 0\}
\]
Higher $\lambda_1$ means faster propagation: narratives injected at any node reach the entire network more quickly.
The spectral gap $\lambda_1 - \lambda_2$ (difference between the two largest eigenvalues) measures the mixing rate: a large gap means the network converges to a steady-state narrative distribution quickly, while a small gap means distinct clusters can maintain different narratives for extended periods.
\end{definition}

\begin{proposition}[Spectral Gap and Fragmentation]
\label{prop:spectral-fragmentation}
\index{spectral gap}
If the spectral gap $\lambda_1 - \lambda_2$ is small (near zero), the network is close to disconnected: it contains clusters that communicate weakly.
In doxonomic terms, a small spectral gap predicts epistemic fragmentation (Definition~\ref{def:fragmentation}): the network can sustain divergent belief systems in different clusters.
A large spectral gap predicts narrative convergence: information flows rapidly enough across the network that maintaining divergent narratives is difficult.
\end{proposition}

\begin{remark}
The spectral gap provides a structural prediction: systems with small spectral gaps are vulnerable to polarization, while systems with large spectral gaps are vulnerable to common-sense capture.
Both are doxonomic risks; the question is which one a given society faces.
Public broadcasting, shared curricula, and cross-community institutions increase the spectral gap (by adding cross-cluster edges), making polarization harder but capture easier.
This is a structural version of the coherence-fragmentation tradeoff (Chapter~\ref{ch:contradiction-coherence}).
\end{remark}


\section{Temporal Network Dynamics}
\label{sec:ch15-temporal}

Doxonomic networks are not static.
They evolve as institutions are founded or dissolved, as funding flows shift, as personnel move, and as new media channels emerge.

\begin{definition}[Network Evolution]
\label{def:network-evolution}
\index{network evolution}
A \emph{temporal doxonomic network} is a sequence $\{G_t\}_{t=0}^{T}$ of networks on a (possibly changing) vertex set $V_t$ with time-varying edges $E_t$ and weights $w_t$.
Key events include:
\begin{itemize}[nosep]
  \item \textbf{Node entry}: a new institution is founded (Heritage Foundation, 1973; Fox News, 1996; Breitbart, 2007).
  \item \textbf{Node exit}: an institution closes or loses relevance.
  \item \textbf{Edge rewiring}: funding flows shift (a foundation redirects grants from one think tank to another).
  \item \textbf{Weight changes}: an existing connection strengthens or weakens (a newspaper increases its reliance on a particular think tank as a source).
\end{itemize}
\end{definition}

\begin{example}[Network Evolution: Conservative Infrastructure 1970--2020]
\label{ex:network-evolution}
\begin{itemize}[nosep]
  \item \textbf{1970}: sparse network with a few academic nodes (Chicago, Virginia), the Mont P\`{e}lerin Society, and limited institutional infrastructure.
  \item \textbf{1980}: new nodes added (Heritage, Cato, AEI expansion, Olin-funded law-and-economics programs), dense funding edges from Olin/Scaife/Bradley/Koch.
  \item \textbf{1996}: Fox News enters as a high-reach amplifier node, dramatically increasing the spectral radius and connecting the policy cluster to a mass audience.
  \item \textbf{2007--2015}: digital nodes (Breitbart, Daily Wire, PragerU, Turning Point USA) add a new layer with very high reach but lower credibility, creating the two-tier structure described in Model~\ref{mod:two-tier}.
  \item \textbf{2020}: the network is far denser, has higher total throughput, but also shows internal fragmentation (establishment conservative cluster vs.\ populist cluster, with weakening cross-cluster edges).
\end{itemize}
Tracking this evolution requires longitudinal data: 990 filings over decades, citation analyses across multiple time periods, personnel movement records, and audience-overlap surveys.
\end{example}


\section{Network Resilience and Attack Vulnerability}
\label{sec:ch15-resilience}

\begin{definition}[Network Resilience]
\label{def:resilience}
\index{network resilience}
A doxonomic network is \emph{resilient} if the removal of a small number of nodes does not significantly reduce the maximum pathway efficiency (Definition~\ref{def:pathway}) or increase epistemic fragmentation.
\end{definition}

\begin{proposition}[Scale-Free Vulnerability]
\label{prop:scale-free}
\index{scale-free network!vulnerability}
If a doxonomic network is scale-free (a small number of hubs have most connections, as is typical for citation and media networks), it is:
\begin{itemize}[nosep]
  \item \textbf{robust} to random node removal (removing random institutions has little effect, because most nodes are peripheral),
  \item \textbf{vulnerable} to targeted attack on hubs (removing the few most-connected nodes rapidly fragments the network).
\end{itemize}
\end{proposition}

This has implications for both counter-doxonomic strategy and doxonomic defense.
Counter-doxonomic actors should target hubs (high-centrality institutions whose removal would fragment the narrative pipeline).
Doxonomic actors should diversify their networks to reduce hub dependency.

\begin{example}[Targeted vs.\ Random Disruption]
\label{ex:disruption}
Suppose a doxonomic network has 50 institutions, with 3 hubs (a major think tank, a prestige newspaper, a dominant broadcaster) accounting for 60\% of total throughput.
\begin{itemize}[nosep]
  \item \textbf{Random removal} of 3 institutions: expected throughput reduction $\sim 6\%$ (most removed nodes are peripheral).
  \item \textbf{Targeted removal} of the 3 hubs: throughput reduction $\sim 60\%$, plus the network fragments into disconnected components.
\end{itemize}
The ten-fold difference in effect illustrates scale-free vulnerability.
In practice, ``removing'' an institution means discrediting it (credibility attack, Chapter~\ref{ch:epistemic-capital}), cutting its funding, or undermining its gatekeeper position, not literal elimination.
\end{example}


\section{Practical Network Reconstruction}
\label{sec:ch15-practical}

Constructing a doxonomic network from empirical data is a substantial research task.
The practical steps are:

\begin{enumerate}[nosep]
  \item \textbf{Define the boundary}: which institutions, funders, and channels are included? (The boundary choice is an analytical decision that affects all results.)
  \item \textbf{Identify nodes}: list all relevant actors from funding databases, media directories, and institutional registries.
  \item \textbf{Identify edges}: for each node pair, check for funding relationships (990 data), citation relationships (content analysis), personnel movement (biographical data), and board interlocks (organizational filings).
  \item \textbf{Estimate weights}: assign weights based on dollar flows, citation counts, audience share, or a composite measure.
  \item \textbf{Validate}: check the reconstructed network against known case studies.
  If a well-documented funding relationship is missing from the network, the reconstruction is incomplete.
  \item \textbf{Compute metrics}: centrality, modularity, spectral gap, vulnerability indices.
  \item \textbf{Report uncertainty}: edges may be missing (opacity), weights may be estimated, and the boundary choice affects results.
  Sensitivity analysis is essential.
\end{enumerate}


\section{Notes and References}
\label{sec:ch15-notes}

Network analysis draws on \textcite{jackson2008} for economic and social networks.
Community detection via modularity is due to \textcite{newman2006}; for a review of community detection methods, see Fortunato (2010).
Betweenness centrality is from \textcite{freeman1977}.
Eigenvector centrality and PageRank are related; for the connection, see Langville and Meyer (2006).
For applications to media and political networks, see \textcite{watts2002} and \textcite{benkler2018}.
Donor networks in U.S.\ politics are mapped in \textcite{mayer2016} and \textcite{phillipsfein2009}.

Scale-free network vulnerability is analyzed in Albert, Jeong, and Barab\'{a}si (2000).
Spectral graph theory is covered in Chung (1997).
Multiplex networks are treated in Kivel\"{a} et al.\ (2014).
Board interlocks as a form of elite coordination are studied in Mizruchi (1996).
For temporal network analysis, see Holme and Saram\"{a}ki (2012).


\section{Exercises}

\begin{exercise}
Compute the eigenvector centrality of each node in the example network (Section~\ref{sec:ch15-example}).
Which node has the highest doxonomic centrality?
Does the centrality ranking change if you use betweenness centrality instead?
\end{exercise}

\begin{exercise}
Using citation data from a policy domain of your choice, construct a donor-institution-media network with at least 10 nodes.
Identify narrative clusters using modularity maximization.
Who are the epistemic brokers?
\end{exercise}

\begin{exercise}
Prove that removing a node $v$ from a connected graph $G$ increases the number of connected components if and only if $v$ is a \emph{cut vertex} (its removal disconnects $G$).
Is every node with high betweenness centrality a cut vertex?
\end{exercise}

\begin{exercise}
\label{ex:ch15-spectral}
For the four-sector input-output system in Example~\ref{ex:io-construction}, compute the spectral radius $\lambda_1$ and the spectral gap $\lambda_1 - \lambda_2$.
Predict whether the system is more vulnerable to fragmentation or to common-sense capture.
\end{exercise}

\begin{exercise}
\label{ex:ch15-temporal}
Sketch the temporal evolution of a doxonomic network in a domain you know well (e.g., education policy, climate discourse, tech regulation).
Identify at least three key events that changed the network structure.
How did each event affect centrality rankings and cluster structure?
\end{exercise}

\begin{exercise}
\label{ex:ch15-resilience}
For a network with 20 nodes where 3 hubs account for 50\% of throughput, compute:
(a)~the expected throughput reduction from removing 3 random nodes,
(b)~the throughput reduction from removing the 3 hubs.
Propose a ``doxonomic antifragility'' strategy that reduces hub dependency.
\end{exercise}

\begin{exercise}
The chapter identifies five network projections (funding, citation, personnel, board interlock, audience overlap).
For a specific institution of your choice, describe its position in each projection.
In which projection is it most central?
In which is it most peripheral?
What does the difference tell you about its doxonomic role?
\end{exercise}

\begin{exercise}
Milton Friedman is described as an epistemic broker connecting academic, media, and political clusters.
Identify a contemporary equivalent: a person or institution that bridges two or more doxonomic clusters.
Map their connections.
What would happen to narrative flow between the clusters if the broker were removed?
\end{exercise}
