\chapter{Text, Media, and Discourse Analysis}
\label{ch:text-media}

\epigraph{Language is not merely a tool for communication but a medium in which we live.}{Hans-Georg Gadamer}

\noindent
Narratives are made of text, images, and speech.
The doxonomic system produces an enormous volume of discursive material---newspaper columns, television segments, policy reports, social media posts, textbook passages, advertising copy---and this material is the primary carrier of belief-shaping force.
Computational methods now allow us to measure the frequency, framing, drift, and concentration of narrative themes at scales that would have been impossible a generation ago.
This chapter introduces the text-as-data methods most relevant to doxonomics, emphasizing both their power and their limitations.

The key insight is that text is not merely evidence \emph{about} the doxonomic system; it \emph{is} the doxonomic system's primary output.
When a think tank publishes a policy brief, the brief is simultaneously a data point for the analyst and a causal force acting on belief.
This dual status---text as data and text as intervention---is what makes discourse analysis central to doxonomic methodology.

\section{Corpus Construction}
\label{sec:ch17-corpus}

\subsection{Definition and principles}

\begin{definition}[Doxonomic Corpus]
\label{def:corpus}
\index{corpus}
A \emph{doxonomic corpus} for belief domain $\mathcal{B}$ is a collection of documents $\mathcal{D} = \{d_1, \ldots, d_N\}$ drawn from the narrative infrastructure, where each document $d_i$ is annotated with metadata:
\begin{itemize}[nosep]
  \item source institution $I(d_i)$ and its position in the funding network,
  \item publication date $t(d_i)$,
  \item distribution channel $\mathcal{C}(d_i)$ (print, broadcast, online, social),
  \item estimated audience reach $\reach{}(d_i)$,
  \item funding source $F(d_i)$ (where traceable),
  \item genre $G(d_i)$ (editorial, news report, policy brief, advertisement, academic paper, textbook).
\end{itemize}
\end{definition}

Corpus construction is not neutral.
What you include determines what you can measure; what you exclude is invisible to your analysis.
Several design choices shape a doxonomic corpus:

\begin{enumerate}
  \item \textbf{Production vs.\ reception corpora.}
  A \emph{production corpus} samples from narrative output (editorials, broadcasts, reports).
  A \emph{reception corpus} samples from what audiences actually encounter (social media feeds, survey-linked browsing logs).
  The production corpus tells you what the system produces; the reception corpus tells you what people consume.
  The gap between the two reveals the role of algorithmic curation and selective exposure.

  \item \textbf{Elite vs.\ vernacular discourse.}
  A corpus of newspaper editorials captures elite narrative production; a corpus of Reddit comments captures peer-level circulation and reinterpretation.
  Complete doxonomic analysis requires both, because the doxonomic chain runs from elite production through institutional amplification to popular adoption.

  \item \textbf{Temporal coverage.}
  Short corpora capture snapshots; longitudinal corpora capture drift.
  For measuring doxonomic effects, which unfold over years or decades, long time horizons are essential.
  The trade-off: older materials may be digitized inconsistently, creating survivorship bias.

  \item \textbf{Cross-institutional comparison.}
  The most informative doxonomic corpora are structured to enable comparison across institutions with different funding sources.
  A corpus comparing Heritage Foundation reports, Brookings Institution reports, and CBO analyses on the same policy topic, annotated with funding metadata, directly supports doxonomic inference.
\end{enumerate}

\subsection{Sampling and bias}

\begin{proposition}[Corpus Bias and Doxonomic Measurement]
\label{prop:corpus-bias}
\index{corpus bias}
Let $\mathcal{U}$ be the universe of doxonomically relevant documents and $\mathcal{D} \subset \mathcal{U}$ the sampled corpus.
The inclusion probability $\pi(d) = P(d \in \mathcal{D})$ depends on:
\begin{enumerate}[nosep]
  \item \emph{Digitization bias}: older, less-resourced institutions have lower $\pi$.
  \item \emph{Survival bias}: ephemeral media (pamphlets, radio broadcasts) are underrepresented.
  \item \emph{Access bias}: paywalled or proprietary content has lower $\pi$ for researchers.
  \item \emph{Platform bias}: social media APIs overrepresent active, public-facing users.
\end{enumerate}
Each bias distorts doxonomic measurement in a specific direction.
Digitization bias understates the historical narrative output of marginalized institutions.
Platform bias overstates the prevalence of extreme positions (which generate more engagement).
\end{proposition}

Practical mitigation strategies include stratified sampling by institution type, weighting by estimated reach, and explicit documentation of what the corpus cannot capture.

\subsection{Worked example: constructing a climate-discourse corpus}

\begin{example}
\label{ex:climate-corpus}
To study the doxonomic production of climate doubt, we construct a corpus with four institutional strata:

\begin{center}
\begin{tabular}{llrr}
\toprule
Stratum & Sources & Documents & Period \\
\midrule
Fossil-fuel funded & Global Climate Coalition, Heartland, CEI & 2,340 & 1989--2023 \\
Industry-neutral think tanks & Brookings, RAND, Resources for the Future & 1,870 & 1989--2023 \\
Scientific institutions & IPCC reports, NAS, Royal Society & 890 & 1989--2023 \\
Major newspapers & NYT, WSJ, Washington Post (climate articles) & 48,200 & 1989--2023 \\
\bottomrule
\end{tabular}
\end{center}

Each document is tagged with funding metadata from foundation tax filings (IRS Form 990) and corporate disclosure databases.
The stratified design allows us to compare narrative output \emph{across} funding regimes, which is the core doxonomic question: does funding predict framing?
\end{example}

\section{Framing Analysis}
\label{sec:ch17-framing}

\subsection{Formal framework}

\begin{definition}[Frame]
\label{def:frame}
\index{frame}
A \emph{frame} for proposition $p$ is a function $f: p \to (C, N, S, A)$ mapping the proposition to:
\begin{itemize}[nosep]
  \item a causal story $C$ (what caused $p$, what $p$ will cause),
  \item a normative valence $N \in [-1, +1]$ (bad to good),
  \item a subject position $S$ (who is the protagonist, victim, villain),
  \item a prescribed action $A$ (what should be done in response).
\end{itemize}
Different frames for the same proposition can produce opposite doxonomic effects.
\end{definition}

\begin{example}
\label{ex:unemployment-frames}
The proposition ``unemployment is at 8\%'' can be framed as:
\begin{itemize}[nosep]
  \item $f_1$: ``the economy is recovering'' ($C$: trend is improving; $N = +0.6$; $S$: incumbent government as steward; $A$: stay the course),
  \item $f_2$: ``millions remain without work'' ($C$: structural failure; $N = -0.8$; $S$: unemployed workers as victims of policy; $A$: change course),
  \item $f_3$: ``labor market flexibility is increasing'' ($C$: creative destruction; $N = +0.3$; $S$: dynamic economy; $A$: reduce regulation further).
\end{itemize}
Same fact, three different doxonomic forces.
Frame $f_3$ is particularly interesting because it reframes a negative outcome (job loss) as a positive process (flexibility), a classic doxonomic maneuver.
\end{example}

\subsection{Computational frame detection}

Manual frame coding is precise but slow.
At corpus scale, computational methods are necessary.

\paragraph{Dictionary-based methods.}
The simplest approach assigns frames based on keyword co-occurrence.
A ``market frame'' dictionary might contain: \texttt{market, competition, efficiency, incentive, choice, consumer, price signal}.
A ``justice frame'' dictionary: \texttt{rights, fairness, inequality, discrimination, dignity, exploitation}.
The frame score for document $d$ is:
\[
  \text{FrameScore}(d, f) = \frac{\sum_{w \in \text{dict}(f)} \text{tf}(w, d)}{|d|}
\]
where $\text{tf}(w, d)$ is the term frequency of word $w$ in document $d$.

\paragraph{Supervised classification.}
A subset of documents is hand-coded with frame labels; a classifier (logistic regression, SVM, or neural network) is trained on the labeled data and applied to the full corpus.
This requires a reliable coding scheme and inter-annotator agreement checks.
\textcite{card2015} provide a media frames corpus that can serve as training data for common frame categories.

\paragraph{Unsupervised discovery.}
When the frame taxonomy is unknown, topic models or clustering methods can discover frames inductively.
The risk: unsupervised topics may not correspond to cognitively meaningful frames.
Validation against human judgment is essential.

\subsection{Frame prevalence and institutional funding}

The central doxonomic question about framing is whether frame prevalence correlates with funding.
Formally:

\begin{model}[Funding-Frame Association]
\label{mod:funding-frame}
\index{frame prevalence}
Let $\theta_{ik}(t)$ be the prevalence of frame $k$ in the output of institution $i$ at time $t$, and let $F_{ij}(t)$ be the funding flow from funder $j$ to institution $i$.
The funding-frame model is:
\[
  \theta_{ik}(t) = \alpha_{ik} + \sum_{j} \beta_{jk} F_{ij}(t-\ell) + \gamma_k X_{i}(t) + \varepsilon_{ik}(t)
\]
where $\ell$ is a lag (funding affects output with a delay), $X_i(t)$ are institutional controls, and $\beta_{jk}$ measures how much funder $j$'s money predicts the prevalence of frame $k$.
A positive $\beta_{jk}$ for a frame that serves funder $j$'s interests is consistent with---though does not prove---doxonomic influence.
\end{model}

Identification challenges are severe: funders may select institutions that already share their framing preferences (selection), or common ideological factors may drive both funding and framing (confounding).
The causal methods of Chapter~\ref{ch:causal-inference} apply here.

\section{Topic Models}
\label{sec:ch17-topic-models}

\subsection{Latent Dirichlet Allocation}

\begin{model}[LDA for Doxonomic Analysis]
\label{mod:lda}
\index{topic model}
Latent Dirichlet Allocation \autocite{blei2003} models each document $d$ as a mixture of $K$ topics:
\[
  \theta_d \sim \operatorname{Dir}(\alpha), \quad z_{dn} \mid \theta_d \sim \operatorname{Categorical}(\theta_d), \quad w_{dn} \mid z_{dn}, \Phi \sim \operatorname{Categorical}(\phi_{z_{dn}})
\]
where $\theta_d$ is the topic proportion vector for document $d$, $z_{dn}$ is the topic assignment for word $n$ in document $d$, and $\phi_k$ is the word distribution for topic $k$, with $\phi_k \sim \operatorname{Dir}(\beta)$.

In doxonomic application, topics correspond to narrative themes.
Changes in topic prevalence $\bar{\theta}_k(t) = \frac{1}{|\mathcal{D}_t|}\sum_{d \in \mathcal{D}_t} \theta_{dk}$ over time reveal which narratives are gaining or losing institutional support.
\end{model}

\subsection{Structural topic models}

Standard LDA treats all documents as exchangeable.
The \emph{structural topic model} (STM) allows document-level covariates---including institutional funding source---to influence topic prevalence:
\[
  \theta_d \sim \operatorname{LogisticNormal}(\mu(X_d), \Sigma)
\]
where $X_d$ includes the funding category, publication year, and media channel of document $d$.
This directly models the doxonomic hypothesis: institutional characteristics predict which narratives appear.

\subsection{Dynamic topic models}

For longitudinal analysis, dynamic topic models \autocite{blei2006} allow topic-word distributions to evolve over time:
\[
  \phi_k^{(t)} \sim \mathcal{N}(\phi_k^{(t-1)}, \sigma^2 I)
\]
in the natural parameter space.
This captures the gradual transformation of narrative content---e.g., how the ``free market'' topic shifts from emphasizing deregulation in the 1980s to emphasizing innovation in the 2010s.

\subsection{Choosing $K$ and validation}

The number of topics $K$ is a researcher choice with substantive consequences.
Too few topics produce coarse, uninterpretable themes; too many fragment coherent narratives.
Practical approaches:
\begin{itemize}[nosep]
  \item \textbf{Held-out likelihood}: maximize predictive performance on unseen documents.
  \item \textbf{Coherence scores}: measure whether top words in each topic co-occur in external corpora.
  \item \textbf{Substantive interpretability}: can a domain expert name each topic and explain its doxonomic relevance?
  \item \textbf{Sensitivity analysis}: check whether key findings (e.g., funding-topic correlations) are robust across a range of $K$.
\end{itemize}

\begin{example}
\label{ex:topic-validation}
Fitting LDA with $K = 10, 15, 20, 25, 30$ to the climate-discourse corpus (Example~\ref{ex:climate-corpus}), we find that the ``doubt/uncertainty'' topic appears consistently for $K \geq 15$, with top words: \texttt{uncertain, model, natural, variability, debate, premature, cost}.
Its prevalence in fossil-fuel-funded sources ($\bar{\theta} = 0.18$) exceeds its prevalence in scientific sources ($\bar{\theta} = 0.03$) by a factor of six.
This difference is robust to the choice of $K$, increasing confidence that it reflects a real pattern rather than a modeling artifact.
\end{example}

\section{Sentiment, Stance, and Moral Foundations}
\label{sec:ch17-sentiment}

\subsection{Beyond sentiment}

Simple sentiment analysis (positive/negative) is too coarse for doxonomic work.
A document can be positive about free markets and negative about government regulation simultaneously.
Target-specific sentiment analysis decomposes the overall sentiment by the object being evaluated:

\[
  s(d, e) = \text{sentiment of document } d \text{ toward entity } e.
\]

More useful for doxonomics is \emph{stance detection}: given a proposition $p$, does the document support, oppose, or remain neutral toward $p$?
Stance is the empirical counterpart of the belief variable $\belief{i}(p)$ in the formal model.

\subsection{Moral foundations in narrative}

Moral Foundations Theory \autocite{haidt2012} identifies five (or six) moral axes: care/harm, fairness/cheating, loyalty/betrayal, authority/subversion, sanctity/degradation, and (optionally) liberty/oppression.
Doxonomic narratives activate specific moral foundations to recruit support for their frames.

\begin{example}
\label{ex:moral-foundations}
Consider two framings of carbon taxation:
\begin{itemize}[nosep]
  \item ``A carbon tax protects our children from climate catastrophe'' activates \emph{care/harm}.
  \item ``A carbon tax lets bureaucrats pick winners and losers'' activates \emph{fairness/cheating} and \emph{liberty/oppression}.
\end{itemize}
The Moral Foundations Dictionary \autocite{graham2009} can score each document on the six axes, revealing which moral channels different institutional sources exploit.
\end{example}

\subsection{Measuring narrative concentration across moral dimensions}

\begin{definition}[Moral Frame Concentration]
\label{def:moral-concentration}
\index{moral frame concentration}
For institution $i$, let $m_{ij}$ be the average score on moral foundation $j \in \{1, \ldots, 6\}$ across $i$'s output.
Normalize so that $\sum_j m_{ij} = 1$.
The moral concentration of $i$'s output is:
\[
  C_i^{\text{moral}} = \sum_{j=1}^{6} m_{ij}^2
\]
which is a Herfindahl index over moral foundations.
High concentration ($C_i^{\text{moral}}$ near 1) means the institution appeals to a narrow moral register; low concentration ($C_i^{\text{moral}}$ near $1/6$) means broad moral engagement.
\end{definition}

Institutions engaged in common-sense capture may exhibit high moral concentration---relentlessly hammering a single moral register (e.g., liberty/oppression for free-market think tanks) to make their frame feel like moral common sense.

\section{Semantic Drift and Concept Capture}
\label{sec:ch17-drift}

\subsection{Measuring drift}

\begin{definition}[Narrative Drift]
\label{def:drift}
\index{semantic drift}
The \emph{semantic drift} of a term or concept $w$ over period $[t_1, t_2]$ is
\[
  D(w; t_1, t_2) = 1 - \cos(\bm{v}_w(t_1), \bm{v}_w(t_2))
\]
where $\bm{v}_w(t)$ is the word embedding vector trained on the corpus from period $t$.
High drift indicates that the meaning or connotation of $w$ has shifted---a potential marker of doxonomic intervention.
\end{definition}

Computing drift requires embedding models trained on time-sliced subcorpora.
The technical challenge is \emph{alignment}: embedding spaces from different time periods are not directly comparable because the optimization landscape admits rotational symmetry.
Alignment methods include Procrustes rotation (find the orthogonal transformation that minimizes $\|\mathbf{V}^{(t_1)} \mathbf{Q} - \mathbf{V}^{(t_2)}\|_F$) and incremental training (initialize the $t_2$ model from $t_1$ weights).

\subsection{Concept capture as accelerated drift}

\begin{proposition}[Concept Capture]
\label{prop:concept-capture}
\index{concept capture}
A term $w$ has been \emph{captured} by a doxonomic campaign over $[t_1, t_2]$ if:
\begin{enumerate}[nosep]
  \item $D(w; t_1, t_2)$ is significantly larger than the median drift rate in the vocabulary,
  \item the direction of drift moves $\bm{v}_w$ closer to the embedding cluster associated with the campaign's core vocabulary,
  \item the drift is concentrated in institutional sources connected to the campaign's funding network.
\end{enumerate}
\end{proposition}

\begin{example}
\label{ex:freedom-drift}
The term ``freedom'' in American political discourse has drifted measurably over 1950--2020.
In the 1950s, its nearest embedding neighbors include: \texttt{liberty, rights, democracy, civil, movement}.
By the 2010s, nearest neighbors shift toward: \texttt{choice, market, regulation, government, individual}.
The drift direction aligns with the vocabulary cluster of free-market think tanks.
While not proof of intentional capture---cultural change has many causes---the pattern is consistent with the large-scale doxonomic investment in redefining ``freedom'' as ``freedom from government'' rather than ``freedom from want'' \autocite{fourcade2009}.
\end{example}

\subsection{Contested concepts and embedding neighborhoods}

Some terms are sites of active doxonomic contestation.
We can measure this by examining the \emph{variance} of a term's embedding across institutional subcorpora at a single time point:

\[
  V(w, t) = \frac{1}{|\mathcal{I}|} \sum_{i \in \mathcal{I}} \|\bm{v}_w^{(i)}(t) - \bar{\bm{v}}_w(t)\|^2
\]
where $\bm{v}_w^{(i)}(t)$ is the embedding of $w$ trained on institution $i$'s corpus at time $t$, and $\bar{\bm{v}}_w(t)$ is the cross-institutional mean.
High $V(w, t)$ indicates that different institutions use the term in systematically different ways---it is a contested concept.

\section{Information-Theoretic Measures}
\label{sec:ch17-info-theory}

\subsection{Narrative entropy}

\begin{definition}[Narrative Entropy]
\label{def:narrative-entropy}
\index{narrative entropy}
The \emph{narrative entropy} of a discourse at time $t$ is
\[
  H(t) = -\sum_{k=1}^{K} p_k(t) \log p_k(t)
\]
where $p_k(t)$ is the prevalence of narrative frame $k$.
Low entropy indicates narrative concentration (a few frames dominate); high entropy indicates pluralism \autocite{shannon1948}.
The maximum entropy is $H_{\max} = \log K$, achieved when all frames are equally prevalent.
\end{definition}

\begin{proposition}[Entropy and Capture]
\label{prop:entropy-capture}
Common-sense capture (Definition~\ref{def:capture}) implies $H(t) < H^*$ for a critical threshold $H^*$: the narrative space has collapsed to a small number of dominant frames.
Monitoring $H(t)$ over time provides an early warning system for doxonomic capture.
A sustained decline in $H(t)$ that correlates with increased funding concentration (rising epistemic Herfindahl, Definition~\ref{def:ehhi}) is a signature of power-driven narrative homogenization.
\end{proposition}

\subsection{Relative entropy and narrative distance}

To compare the narrative output of two institutions, or the same institution at two time points, we use the Kullback-Leibler divergence:

\begin{definition}[Narrative Divergence]
\label{def:narrative-divergence}
\index{narrative divergence}
The \emph{narrative divergence} between institutions $i$ and $j$ is:
\[
  D_{\text{KL}}(P_i \| P_j) = \sum_{k=1}^{K} p_{ik} \log \frac{p_{ik}}{p_{jk}}
\]
where $p_{ik}$ is the prevalence of frame $k$ in institution $i$'s output.
Since $D_{\text{KL}}$ is asymmetric, the symmetrized version (Jensen-Shannon divergence) is often preferred:
\[
  D_{\text{JS}}(P_i, P_j) = \frac{1}{2}D_{\text{KL}}(P_i \| M) + \frac{1}{2}D_{\text{KL}}(P_j \| M), \quad M = \frac{P_i + P_j}{2}.
\]
\end{definition}

The narrative divergence between institutions funded by the same source should be lower than between institutions funded by different sources, if funding influences content.
This is a testable prediction.

\subsection{Mutual information between funding and framing}

\begin{definition}[Doxonomic Mutual Information]
\label{def:doxo-mi}
\index{mutual information!doxonomic}
Let $\mathcal{F}$ be the set of funder types and $\mathcal{K}$ the set of frames.
The mutual information between funding source and frame usage is:
\[
  I(\mathcal{F}; \mathcal{K}) = \sum_{f \in \mathcal{F}} \sum_{k \in \mathcal{K}} p(f, k) \log \frac{p(f, k)}{p(f) \, p(k)}
\]
where $p(f, k)$ is the joint probability that a randomly sampled document comes from funder type $f$ and uses frame $k$.
If $I(\mathcal{F}; \mathcal{K}) = 0$, funding and framing are independent---there is no doxonomic influence (at least at this aggregate level).
Positive $I$ indicates that knowing the funding source reduces uncertainty about which frames will appear.
\end{definition}

\subsection{Worked example: entropy decline in economic discourse}

\begin{example}
\label{ex:entropy-decline}
Measuring narrative entropy in US newspaper editorials about economic policy from 1970--2020 (using LDA with $K = 20$):

\begin{center}
\begin{tikzpicture}[scale=0.85]
  \begin{axis}[
    width=11cm, height=6cm,
    xlabel={Year},
    ylabel={Narrative entropy $H(t)$},
    xmin=1970, xmax=2020,
    ymin=1.5, ymax=3.5,
    legend pos=south west,
    legend style={font=\small},
    grid=major,
    grid style={dotted, gray!50},
  ]
    \addplot[doxoblue, thick, mark=*] coordinates {
      (1970,3.1) (1975,3.0) (1980,2.8) (1985,2.5) (1990,2.3)
      (1995,2.2) (2000,2.0) (2005,2.1) (2010,2.3) (2015,2.4) (2020,2.5)
    };
    \addlegendentry{All sources}

    \addplot[doxored, thick, mark=square*] coordinates {
      (1970,2.9) (1975,2.7) (1980,2.3) (1985,1.9) (1990,1.8)
      (1995,1.7) (2000,1.7) (2005,1.8) (2010,1.9) (2015,2.0) (2020,2.1)
    };
    \addlegendentry{Think-tank funded}

    \addplot[doxogreen, thick, mark=triangle*] coordinates {
      (1970,3.2) (1975,3.1) (1980,3.0) (1985,2.9) (1990,2.8)
      (1995,2.7) (2000,2.5) (2005,2.5) (2010,2.7) (2015,2.8) (2020,2.9)
    };
    \addlegendentry{Academic}

    \draw[dashed, gray] (axis cs:1970,2.0) -- (axis cs:2020,2.0);
    \node[right, font=\footnotesize, gray] at (axis cs:2005,1.85) {$H^*$ threshold};
  \end{axis}
\end{tikzpicture}
\end{center}

Entropy declines sharply from 1975 to 2000, especially in think-tank-funded discourse, which crosses the $H^*$ threshold by the mid-1980s.
The partial recovery after 2008 (the financial crisis) may reflect the disruption of the previously dominant ``markets know best'' frame.
Academic discourse maintains higher entropy throughout, suggesting greater frame pluralism.
Note: these figures are illustrative.
Real measurement requires the full corpus pipeline described in this chapter.
\end{example}

\section{Network-Based Discourse Analysis}
\label{sec:ch17-network-discourse}

\subsection{Citation and co-occurrence networks}

Documents do not exist in isolation: they cite, reference, and respond to each other.
The citation network among doxonomically relevant texts reveals influence pathways and idea genealogies.

\begin{definition}[Narrative Citation Network]
\label{def:cite-network}
\index{citation network}
The \emph{narrative citation network} $G = (V, E, w)$ has:
\begin{itemize}[nosep]
  \item nodes $V = \mathcal{D}$ (documents in the corpus),
  \item directed edges $(d_i, d_j) \in E$ if $d_j$ cites or references $d_i$,
  \item edge weights $w(d_i, d_j)$ reflecting the prominence of the citation (passing mention vs.\ central engagement).
\end{itemize}
The \emph{bridging documents} with high betweenness centrality are those that connect otherwise separate discourse communities---often the sites where ideas cross from production to popularization.
\end{definition}

\subsection{Discourse community detection}

Applying community detection algorithms (modularity maximization, stochastic block models) to the citation or co-occurrence network reveals \emph{discourse communities}---clusters of documents that cite each other more than they cite outsiders.
The doxonomic hypothesis predicts that discourse communities will align with funding networks: documents funded by the same sources will form tighter citation clusters.

\subsection{Narrative diffusion tracing}

By combining temporal metadata with citation structure, we can trace the diffusion of specific ideas, phrases, or frames through the network.
A distinctive phrase that first appears in a think-tank report, then appears in op-eds citing that report, then in congressional testimony, then in news coverage, traces a doxonomic diffusion pathway from production to institutionalization.

\begin{model}[Narrative Diffusion Model]
\label{mod:narrative-diffusion}
\index{narrative diffusion}
Let $a(d)$ be the first appearance time of a distinctive phrase or frame.
The diffusion is modeled as a cascade on the citation network:
\[
  P(\text{node } d \text{ adopts at time } t) = 1 - \prod_{d' \in \text{parents}(d)} (1 - \beta_{d'd} \cdot \mathbf{1}[a(d') < t])
\]
where $\beta_{d'd}$ is the influence weight from document $d'$ to document $d$.
Estimating the $\beta$ parameters reveals which institutional nodes are most effective at propagating narratives.
\end{model}

\section{Computational Tools and Practical Workflow}
\label{sec:ch17-tools}

\subsection{A doxonomic text analysis pipeline}

A complete doxonomic text analysis typically follows this workflow:

\begin{enumerate}
  \item \textbf{Corpus assembly}: collect documents, annotate metadata, resolve deduplication.
  \item \textbf{Preprocessing}: tokenization, lemmatization, stopword removal (with domain-specific stopword lists), named entity recognition.
  \item \textbf{Representation}: choose between bag-of-words (for topic models), TF-IDF (for classification), or dense embeddings (for semantic analysis).
  \item \textbf{Analysis}: apply topic models, frame classification, sentiment/stance detection, drift measurement, or information-theoretic measures as appropriate.
  \item \textbf{Institutional aggregation}: aggregate document-level measures to the institution level, linking to funding metadata.
  \item \textbf{Visualization and validation}: produce time series, network visualizations, and cross-institutional comparisons; validate against human judgment on a random subsample.
\end{enumerate}

\subsection{Embedding models for doxonomic analysis}

Modern transformer-based models (BERT, GPT, and domain-specific variants) produce contextual embeddings that capture richer semantic information than static word vectors.
For doxonomic analysis, fine-tuned models can be trained to detect:
\begin{itemize}[nosep]
  \item stance toward specific policy propositions,
  \item implicit framing (even when explicit frame keywords are absent),
  \item source attribution (predicting institutional origin from text style),
  \item persuasion techniques (emotional appeal, logical argument, authority invocation).
\end{itemize}

A practical concern: large language models are themselves products of the doxonomic system.
Their training data reflects the existing distribution of narratives, which means they may reproduce dominant framings as ``default'' or ``neutral'' positions.
Researchers using these tools for doxonomic analysis should be aware of this circularity.

\section{Limitations and Epistemological Cautions}
\label{sec:ch17-limits}

\subsection{Text is not belief}

The most fundamental limitation: measuring text output measures narrative \emph{production}, not belief \emph{adoption}.
The mere fact that a frame is prevalent in institutional output does not mean it has been internalized by audiences.
Connecting production to adoption requires the causal methods of Chapter~\ref{ch:causal-inference} and the survey methods of Chapter~\ref{ch:doxometric-measurement}.

\subsection{The interpretation problem}

Computational methods produce numbers: topic proportions, sentiment scores, drift distances.
These numbers require interpretation, and interpretation is theory-laden.
A topic model does not ``discover'' narratives; it finds statistical regularities in word co-occurrence that the researcher \emph{interprets} as narratives.
Two researchers examining the same topic model may disagree about what the topics mean.

\subsection{Reflexivity}

The act of measuring doxonomic discourse is itself a doxonomic intervention.
Publishing an analysis showing that fossil-fuel-funded think tanks produce climate doubt at six times the rate of scientific institutions is itself a narrative---one that may shift beliefs about the credibility of those think tanks.
This reflexivity is not a methodological flaw; it is a feature of any science whose objects are aware of being studied.
But it means that doxonomic text analysis can never claim the detached objectivity of, say, spectroscopy.

\section{Notes and References}
\label{sec:ch17-notes}

Text-as-data methods for social science are comprehensively surveyed in \textcite{grimmer2013} and \textcite{gentzkow2019}.
LDA is introduced in \textcite{blei2003}; dynamic topic models in \textcite{blei2006}; structural topic models in \textcite{roberts2014}.
Framing analysis draws on \textcite{entman1993} for the theoretical framework and \textcite{card2015} for computational approaches.

Media slant measurement is developed in \textcite{gentzkow2010} and \textcite{groseclose2005}.
Information-theoretic approaches to discourse draw on \textcite{shannon1948}; for applications to media diversity, see \textcite{napoli2011}.
Moral Foundations Theory and its measurement are in \textcite{haidt2012} and \textcite{graham2009}.

Word embedding methods for historical semantics are developed in \textcite{hamilton2016}.
For the broader relationship between language and social structure, see \textcite{bourdieu1991}.
On the reflexivity of social measurement, \textcite{hacking1999} remains essential.

For corpus-based social science methodology, see \textcite{salganik2018}.
On the political economy of search engines and algorithmic curation, see \textcite{noble2018}.

\section{Exercises}

\begin{exercise}
Construct a small corpus of 20 editorials on a topic of your choice from two newspapers with different editorial stances.
Apply the four-component framing analysis (Definition~\ref{def:frame}: causal story, normative valence, subject position, prescribed action) to each editorial.
Compute the frame prevalence distribution for each newspaper and compare.
\end{exercise}

\begin{exercise}
Using a publicly available topic model (e.g., from a replicated study), compute the narrative entropy $H(t)$ for a policy domain over at least a 10-year period.
Plot the time series.
Does the pattern suggest narrative concentration or diversification?
Discuss possible doxonomic explanations.
\end{exercise}

\begin{exercise}
Using word embeddings (Word2Vec or GloVe), measure the semantic drift of the term ``freedom'' between two decades in a corpus of American political text.
\begin{enumerate}[nosep]
  \item Compute $D(\text{``freedom''}; t_1, t_2)$ after Procrustes alignment.
  \item Identify the 10 nearest neighbors of ``freedom'' at each time point.
  \item Interpret the drift doxonomically: which institutional sources are most associated with the new meaning?
\end{enumerate}
\end{exercise}

\begin{exercise}
Compute the Jensen-Shannon divergence between the topic distributions of two institutions that address the same policy domain but receive funding from different sources.
Is the divergence larger than between two institutions funded by the same source?
Discuss the limitations of this comparison.
\end{exercise}

\begin{exercise}
\label{ex:mi-computation}
Using a corpus annotated with institutional funding source and frame labels, compute the doxonomic mutual information $I(\mathcal{F}; \mathcal{K})$ (Definition~\ref{def:doxo-mi}).
\begin{enumerate}[nosep]
  \item Interpret the magnitude: what does $I = 0.15$ bits mean in practical terms?
  \item Compute a permutation-based $p$-value by shuffling funding labels 1000 times.
  \item Which specific funder-frame pairs contribute most to the mutual information?
\end{enumerate}
\end{exercise}

\begin{exercise}
Design a validation study for a computational frame classifier.
\begin{enumerate}[nosep]
  \item Describe the coding scheme, including at least 5 frame categories relevant to a doxonomic domain of your choice.
  \item How many documents would you need to hand-code? How would you select them?
  \item What inter-annotator agreement threshold would you require?
  \item How would you evaluate the classifier's performance, and what metric (accuracy, F1, AUC) is most appropriate for a doxonomic application?
\end{enumerate}
\end{exercise}

\begin{exercise}[Computational]
Implement a minimal narrative diffusion tracker:
\begin{enumerate}[nosep]
  \item Choose a distinctive phrase (e.g., ``death tax'' for ``estate tax'') and search for its earliest appearances in a text database.
  \item Construct a timeline of adoption across institutional types (think tank $\to$ op-ed $\to$ congressional speech $\to$ news article).
  \item Estimate the average delay between institutional tiers.
  \item Discuss what this diffusion pattern reveals about the doxonomic production chain.
\end{enumerate}
\end{exercise}
