\chapter{Historical and Archival Methods}
\label{ch:historical-archival}

\epigraph{History is the memory of states.}{Henry Kissinger}

\noindent
Doxonomic systems operate over decades and centuries.
Computational methods capture the present; historical methods reveal how we arrived here, and reveal it in a form that no regression can provide, because the causal chains that produce today's common sense were laid down by specific actors making specific decisions in specific institutional contexts.
This chapter develops the tools for doxonomic history: genealogy of concepts, institutional archive analysis, long-duration narrative change, comparative-historical methods, and the integration of historical evidence with the computational and statistical methods developed in Chapters~\ref{ch:causal-inference}--\ref{ch:text-media}.

The central premise is that doxonomic processes leave archival traces (in foundation grant records, corporate strategy memos, propaganda agency files, curriculum committee minutes, and media ownership ledgers), and that these traces, when assembled systematically, reveal the material infrastructure behind what presents itself as ``just common sense.''

\section{Genealogy of Concepts}
\label{sec:ch18-genealogy}

\subsection{Definition and scope}

\begin{definition}[Doxonomic Genealogy]
\label{def:genealogy}
\index{genealogy}
A \emph{doxonomic genealogy} of belief $p$ is a reconstruction of:
\begin{enumerate}[nosep]
  \item when $p$ was first articulated as an explicit claim,
  \item which agents promoted it and with what material resources,
  \item through which institutional channels it gained credibility,
  \item what alternative beliefs it displaced and by what mechanisms,
  \item at what point it achieved common-sense status (if ever),
  \item what residual traces of the displaced alternatives survive.
\end{enumerate}
\end{definition}

This extends \textcite{foucault1971} from discourse analysis to \emph{funded} discourse analysis: not just \emph{who said what when}, but \emph{who paid for it to be said, through what institutions, with what effect on the distribution of beliefs in the population}.
The genealogical method is not simply intellectual history (the history of ideas) but institutional-material history (the history of the infrastructure that gave certain ideas reach and persistence while denying those resources to competitors).

\subsection{Genealogy vs.\ intellectual history}

Standard intellectual history traces ideas through their logical development: Locke influenced Smith who influenced Mill who influenced Hayek.
Doxonomic genealogy adds the material dimension: \emph{how} did Hayek's ideas achieve mass influence when other equally rigorous economists (e.g., Joan Robinson, Nicholas Kaldor) did not?
The answer is not in the quality of the arguments but in the institutional infrastructure (Mont P\`{e}lerin Society, the network of free-market think tanks, corporate-funded university chairs) that gave Hayek's framework reach, repetition, and persistence that Robinson's never had.

\begin{proposition}[Genealogical Distinctness Criterion]
\label{prop:genealogy-criterion}
A doxonomic genealogy differs from standard intellectual history if and only if the material conditions of belief production appear as explanatory factors: that is, if removing the funding and institutional analysis from the genealogy would change the explanation of why belief $p$ achieved its current prevalence.
\end{proposition}

\subsection{Constructing a genealogy: method}

A doxonomic genealogy proceeds in five stages:

\begin{enumerate}
  \item \textbf{Identify the target belief and establish its current status.}
  What is the belief?
  Is it common sense, contested, marginal?
  What is the evidence for its current prevalence (surveys, policy outcomes, textbook content)?

  \item \textbf{Trace the intellectual precursors.}
  When was the belief first articulated?
  By whom?
  In what context?
  This is standard intellectual history.

  \item \textbf{Map the institutional infrastructure.}
  Which institutions promoted the belief?
  When were they founded?
  By whom, and with what funding?
  What channels did they use (publications, conferences, curricula, media, policy advisory)?

  \item \textbf{Identify the critical amplification events.}
  At what moments did the belief's prevalence change sharply?
  What combination of institutional action and external events (crises, elections, technological change) produced the shift?

  \item \textbf{Document the displaced alternatives.}
  What beliefs did the target belief replace?
  Were those alternatives refuted, defunded, or simply drowned out?
  Do archival traces of the alternatives survive?
\end{enumerate}

\subsection{Worked example: the genealogy of ``markets are efficient''}

\begin{example}
\label{ex:genealogy-emh}
The belief that financial markets efficiently aggregate information (the efficient markets hypothesis) provides a compact genealogy:

\paragraph{Intellectual precursors.}
Bachelier (1900) on random walks in bond prices; Samuelson (1965) on the impossibility of beating the market; \textcite{fama1970} on the three forms of market efficiency.

\paragraph{Institutional infrastructure.}
The University of Chicago economics department, funded in part by conservative foundations (Olin, Bradley, Scaife) from the 1950s onward.
The Graduate School of Business (now Booth) trained a generation of finance scholars and MBA students.
The Journal of Financial Economics provided a publication outlet.
Corporate partnerships (Wells Fargo's index fund, founded 1971) created commercial incentives to promote the hypothesis.

\paragraph{Amplification.}
The success of index funds in the 1980s--90s provided apparent empirical vindication.
Deregulatory policy (repeal of Glass-Steagall, 1999) was justified partly by efficient-markets reasoning.
Business media (CNBC, Bloomberg) adopted efficient-markets language as default vocabulary.

\paragraph{Displacement.}
Keynesian and institutionalist approaches to finance (which emphasized speculation, herd behavior, and regulatory necessity) were marginalized in leading economics departments by the 1980s, though they survived in heterodox programs.

\paragraph{Crisis and partial genealogical reversal.}
The 2008 financial crisis disrupted the genealogy: behavioral finance gained institutional footing, and former proponents (including Greenspan) acknowledged that the hypothesis had been taken too far.
But the institutional infrastructure (departments, journals, textbooks, curricula) remained largely intact, and the hypothesis recovered much of its common-sense status within a decade.
\end{example}

\section{Institutional Archives}
\label{sec:ch18-archives}

\subsection{Types of archival sources}

Doxonomic history depends on archives that conventional intellectual historians rarely consult.
The key categories:

\begin{enumerate}
  \item \textbf{Foundation records.}
  Annual reports, grant databases, IRS Form 990 filings (for US nonprofits), board meeting minutes.
  These reveal who funded what, how much, and through what intermediaries.
  Major digitized sources include the Foundation Directory Online and GuideStar/Candid.

  \item \textbf{Corporate archives.}
  Advertising strategies, public relations campaigns, internal strategy memos, lobbying expenditure records.
  Many are accessible only through litigation discovery (e.g., tobacco industry documents at the UCSF Truth Tobacco Documents Library) or leak (e.g., ExxonMobil climate research memos).

  \item \textbf{Government propaganda records.}
  State information agencies (USIA, BBC World Service, Soviet-era Agitprop), censorship bureaus, educational ministry curriculum directives.
  National archives in most democracies declassify propaganda records after 25--50 years.

  \item \textbf{Educational archives.}
  Textbook editions, curriculum committee records, accreditation standards, university hiring and tenure records.
  Particularly valuable for tracing the institutionalization of beliefs in the education system.

  \item \textbf{Media ownership records.}
  FCC filings (US), Ofcom records (UK), media concentration databases.
  These document changes in narrative infrastructure ownership.

  \item \textbf{Think-tank publications.}
  Policy briefs, reports, congressional testimony, media appearances.
  Most think tanks maintain publication archives; many are now digitized.
\end{enumerate}

\subsection{Archival methodology for doxonomics}

\begin{model}[Archival Triangulation]
\label{mod:triangulation}
\index{archival triangulation}
Because no single archive captures the full doxonomic chain (from funding decision to belief change), doxonomic archival research requires triangulation across at least three source types:
\begin{enumerate}[nosep]
  \item \emph{Production archives}: what was produced (think-tank publications, media content).
  \item \emph{Infrastructure archives}: who funded the production (foundation records, corporate filings).
  \item \emph{Reception evidence}: what changed in the population (polling data, behavioral indicators, policy outcomes).
\end{enumerate}
The doxonomic claim (that funding shaped narrative which shaped belief) requires evidence from all three levels.
Evidence from production alone shows what was said but not why.
Evidence from infrastructure alone shows who paid but not what resulted.
Evidence from reception alone shows what changed but not what caused the change.
\end{model}

\subsection{Source criticism for doxonomic archives}

Archival sources require critical evaluation, and doxonomic archives present distinctive challenges:

\begin{itemize}
  \item \textbf{Survivorship bias.}
  Archives that survive are disproportionately those of successful, well-funded institutions.
  Failed doxonomic campaigns, suppressed counter-narratives, and defunded organizations leave fewer traces.
  The archive systematically over-represents the winners of doxonomic competition.

  \item \textbf{Strategic self-presentation.}
  Foundation annual reports and corporate public relations materials present a sanitized version of institutional activity.
  Internal memos and litigation discovery documents are more candid but harder to access.

  \item \textbf{Opacity by design.}
  Some doxonomic funding is deliberately obscured through donor-advised funds, pass-through organizations (``dark money''), and offshore foundations.
  The archival record may have intentional gaps \autocite{mayer2016}.

  \item \textbf{Volume and heterogeneity.}
  The archival record for even a single doxonomic campaign (e.g., climate doubt) can comprise millions of pages across dozens of institutional archives.
  Systematic coverage requires computational assistance (OCR, topic modeling, named entity recognition applied to archival materials).
\end{itemize}

\subsection{Worked example: the tobacco documents}

\begin{example}
\label{ex:tobacco-archives}
The UCSF Truth Tobacco Industry Documents Library contains over 14 million documents released through litigation.
These documents revealed a systematic doxonomic campaign: tobacco companies funded research institutions, public relations firms, and front organizations to produce and circulate the narrative that the link between smoking and cancer was ``uncertain'' and ``debated'' \autocite{oreskes2010}.

Key archival findings:
\begin{itemize}[nosep]
  \item A 1969 Brown \& Williamson memo stated: ``Doubt is our product since it is the best means of competing with the `body of fact' that exists in the mind of the general public.''
  \item Funding records show that tobacco companies directed \$450+ million to the Council for Tobacco Research (1954--1998), which published research designed to create the appearance of scientific controversy.
  \item Media analysis showed that news coverage of the smoking-cancer link used ``balance'' framing (quoting both sides equally) for decades after the scientific consensus had solidified, partly because industry-funded scientists provided a steady supply of quotable ``experts.''
\end{itemize}

This case illustrates the archival triangulation model: production archives (industry-funded publications), infrastructure archives (funding records, internal memos), and reception evidence (media framing, public opinion surveys showing persistent doubt).
\end{example}

\section{Long-Duration Narrative Change}
\label{sec:ch18-longue-duree}

\subsection{Periodization of belief regimes}

\begin{model}[Periodization of Belief Regimes]
\label{mod:periodization}
\index{periodization}
A \emph{doxonomic period} is defined by the dominance of a coherent belief regime $\alpha$: a mutually reinforcing set of common-sense beliefs, institutional practices, and narrative infrastructure that together constitute the ``taken for granted'' framework for a domain.
Transitions between periods are marked by:
\begin{enumerate}[nosep]
  \item \emph{Funding shifts}: new donors enter, old donors exit, institutional vehicles change.
  \item \emph{Institutional realignment}: new think tanks, curriculum reforms, media ownership changes.
  \item \emph{Narrative crises}: events (wars, financial collapses, pandemics) that destabilize the dominant frame by making its predictions visibly fail.
  \item \emph{Measurable belief shifts}: changes in population belief distributions detectable through surveys, electoral behavior, or policy outcomes.
\end{enumerate}
\end{model}

The periodization framework draws on the \emph{longue dur\'{e}e} tradition of the Annales school (Braudel, Bloch) but adds a specifically doxonomic layer: the material infrastructure of belief production is treated as a structural feature comparable to geography, demography, or technology.

\subsection{A periodization of economic common sense, 1870--2020}

\begin{example}
\label{ex:econ-periodization}
We can identify five doxonomic periods for economic belief in the Anglophone world:

\begin{center}
\begin{tabular}{p{2.5cm}p{3.5cm}p{3cm}p{3.5cm}}
\toprule
Period & Dominant belief & Key institutions & Transition trigger \\
\midrule
Classical liberalism (1870--1929) & Self-regulating markets & Universities, business press, chambers of commerce & Great Depression \\[4pt]
Keynesian consensus (1933--1970) & Government must manage demand & Government agencies, Brookings, NBER, reformed economics depts.\ & Stagflation, oil shocks \\[4pt]
Neoliberal ascendancy (1970--2008) & Markets are efficient, government is the problem & Heritage, Cato, AEI, Chicago school, IMF, World Bank & 2008 financial crisis \\[4pt]
Contested interregnum (2008--2020) & No stable consensus; populist, technocratic, and progressive frames compete & Fragmented: social media, partisan media, revived heterodox econ & COVID-19, inequality crises \\[4pt]
Post-2020 (emerging) & Uncertain; industrial policy revival, climate urgency & Emerging institutions: climate funds, tech policy think tanks & TBD \\
\bottomrule
\end{tabular}
\end{center}

Each transition involved not just a change in ideas but a change in the institutional infrastructure that produced and circulated ideas.
The neoliberal transition, for instance, required a 30-year institution-building campaign before the political conditions (stagflation, Thatcher/Reagan elections) allowed the new beliefs to achieve policy dominance \autocite{mirowski2009, harvey2005, phillipsfein2009}.
\end{example}

\subsection{Temporal scales in doxonomic change}

Doxonomic processes operate at multiple temporal scales simultaneously:

\begin{enumerate}
  \item \textbf{Event time} (days to weeks): a crisis, scandal, or policy announcement can temporarily shift the narrative landscape.
  Media framing during a crisis can set the terms for subsequent debate.

  \item \textbf{Campaign time} (months to years): a coordinated doxonomic campaign (think-tank launch, advertising blitz, curriculum reform) unfolds over this timescale.
  Most natural experiments and quasi-experimental designs operate at this scale.

  \item \textbf{Institutional time} (decades): the creation, growth, and transformation of narrative institutions.
  Universities, media organizations, and think tanks operate on generational timescales.
  The lag between institutional founding and measurable belief change is typically 10--30 years.

  \item \textbf{Structural time} (generations to centuries): deep shifts in the taken-for-granted framework, from divine right to social contract, from mercantilist to liberal economic thought, from racial hierarchy to formal equality.
  These shifts involve wholesale replacement of the institutional infrastructure, not just modification.
\end{enumerate}

The historian's distinctive contribution to doxonomics is the ability to work across these temporal scales, tracing how event-level disruptions interact with institutional-level dynamics and structural-level transformations.

\section{Comparative-Historical Methods}
\label{sec:ch18-comparative}

\subsection{Mill's methods and doxonomic comparison}

Comparative-historical analysis applies Mill's methods of agreement and difference to historical cases:

\begin{definition}[Doxonomic Comparative Design]
\label{def:comparative}
\index{comparative-historical method}
\textbf{Method of agreement}: if two societies with different economic structures, political systems, and cultures both adopted belief $p$ after building similar narrative infrastructure, the infrastructure (not the background conditions) is the likely cause.

\textbf{Method of difference}: if two otherwise similar societies diverge in their beliefs after one builds narrative infrastructure that the other does not, the infrastructure is the likely cause of the divergence.
\end{definition}

\begin{example}
\label{ex:comparative-neoliberal}
The neoliberal turn occurred in both the US and UK (method of agreement) despite different political systems, welfare state traditions, and economic structures.
What they shared: both received heavy investment in free-market think tanks and both experienced stagflation crises that discredited the Keynesian alternative.

The neoliberal turn did \emph{not} occur in the same form in France or Germany during the same period (method of difference), where think-tank infrastructure was weaker and existing institutions (dirigisme, social market economy) were more robust.
This pattern is consistent with the hypothesis that institutional infrastructure, not just ideas or economic conditions, drove the transition.
\end{example}

\subsection{Process tracing}

Process tracing follows the causal chain from funding decision to belief change step by step within a single case, verifying that each link in the hypothesized mechanism actually operated.

\begin{model}[Doxonomic Process Trace]
\label{mod:process-trace}
\index{process tracing}
A complete doxonomic process trace documents:
\begin{enumerate}[nosep]
  \item the funding decision (who decided to invest, when, how much, and why),
  \item institutional creation or capture (which organizations received the funding and what they did with it),
  \item narrative production (what specific publications, broadcasts, or educational materials were produced),
  \item distribution and amplification (through which channels the narrative reached audiences),
  \item reception and adoption (evidence that the target population's beliefs actually changed),
  \item feedback (how the changed beliefs affected policy, which in turn affected funding for the next cycle).
\end{enumerate}
Each step must be documented with archival evidence.
Gaps in the chain weaken the causal claim proportionally.
\end{model}

\subsection{Counterfactual reasoning}

Historical counterfactuals are inherently speculative, but they serve an important analytical function: they help distinguish necessary from sufficient conditions.

\begin{example}
\label{ex:counterfactual}
\emph{Would neoliberalism have become common sense without the think-tank network?}
The counterfactual requires us to hold fixed the intellectual content (Hayek, Friedman, Buchanan) and the economic conditions (stagflation) while removing the institutional infrastructure.
Several considerations:
\begin{itemize}[nosep]
  \item Similar ideas existed in the 1930s (the original neoliberals) but failed to achieve policy influence without institutional infrastructure.
  \item Comparable economic crises in countries without the think-tank network produced different ideological responses (e.g., France's turn to \emph{dirigisme} rather than marketization after the 1970s crises).
  \item The speed and synchronization of the neoliberal turn across multiple policy domains (labor, finance, trade, welfare) is more consistent with coordinated institutional action than independent intellectual convergence.
\end{itemize}
The counterfactual suggests that the infrastructure was a necessary (though not sufficient) condition.
The ideas and the crisis were also necessary.
It was the \emph{combination} (prepared ideas, built infrastructure, and catalyzing crisis) that produced the regime change.
\end{example}

\section{Integrating Historical and Computational Evidence}
\label{sec:ch18-integration}

\subsection{The complementarity principle}

Historical and computational methods have complementary strengths:

\begin{center}
\begin{tabular}{lcc}
\toprule
Feature & Historical methods & Computational methods \\
\midrule
Temporal depth & Centuries & Decades (limited by digitization) \\
Causal mechanism & Detailed process tracing & Statistical association \\
Scale & Few cases, deep & Many cases, broad \\
Source type & Archives, documents & Digitized corpora, databases \\
Subjectivity & High (interpreter-dependent) & Lower (algorithm-dependent) \\
Counterfactuals & Qualitative reasoning & Formal modeling \\
\bottomrule
\end{tabular}
\end{center}

Neither method alone is sufficient for doxonomic analysis.
Historical methods provide causal mechanism and temporal depth but are limited in scale and reproducibility.
Computational methods provide scale and reproducibility but are limited in causal depth and temporal reach.
The strongest doxonomic studies combine both.

\subsection{Mixed-methods research designs}

\begin{model}[Sequential Mixed-Methods Design]
\label{mod:mixed-methods}
\index{mixed methods}
A productive sequence for doxonomic research:
\begin{enumerate}[nosep]
  \item \textbf{Computational scan}: use topic models and text analysis on a large corpus to identify patterns (e.g., a funding-frame correlation, a semantic drift event).
  \item \textbf{Historical deep dive}: select 2--3 cases where the pattern is strongest and conduct archival process tracing to verify the causal mechanism.
  \item \textbf{Computational verification}: return to the large corpus to check whether the mechanism identified in the deep dive holds more broadly (e.g., does the funding-frame correlation persist after controlling for institutional ideology?).
  \item \textbf{Generalization}: combine the historical mechanism with the computational evidence to formulate a general doxonomic claim, noting the scope conditions under which it holds.
\end{enumerate}
\end{model}

\begin{example}
\label{ex:mixed-methods}
A mixed-methods study of climate doubt:
\begin{enumerate}[nosep]
  \item \emph{Computational scan}: topic modeling of 50,000 climate-related documents shows that the ``uncertainty'' frame is six times more prevalent in fossil-fuel-funded sources than in scientific sources (cf.\ Example~\ref{ex:topic-validation}).
  \item \emph{Historical deep dive}: archival research on ExxonMobil's internal climate research (1977--1988) vs.\ its public communications (1989--2008) reveals a deliberate strategy to emphasize uncertainty despite internal knowledge of climate risks \autocite{oreskes2010}.
  \item \emph{Computational verification}: the funding-uncertainty correlation holds after controlling for institutional political alignment and publication year, suggesting it is not merely a proxy for conservative ideology.
  \item \emph{Generalization}: funded narrative production can sustain an ``uncertainty'' frame in public discourse for decades even when the scientific evidence is overwhelming, provided the funding is sufficient to maintain a steady stream of credentialed dissenting voices.
\end{enumerate}
\end{example}

\section{Oral History and Elite Interviews}
\label{sec:ch18-oral}

Written archives capture institutional output; they rarely capture the informal decision-making, personal networks, and tacit strategies behind doxonomic campaigns.
Oral history and elite interviews fill this gap.

\subsection{Interview design for doxonomic research}

Interviewing participants in doxonomic campaigns (foundation officers, think-tank directors, editors, academics who received targeted funding) requires attention to:
\begin{itemize}[nosep]
  \item \textbf{Retrospective bias}: participants may rationalize or sanitize their past decisions.
  Cross-check interview claims against archival records wherever possible.
  \item \textbf{Strategic non-disclosure}: participants may have ongoing relationships with funders and decline to discuss sensitive topics.
  Anonymization and embargo periods can help.
  \item \textbf{Survivor bias}: those willing to be interviewed are disproportionately those who are comfortable with their role.
  Dissenters, whistleblowers, and those who left the campaign may provide more candid accounts but are harder to locate.
\end{itemize}

\subsection{Snowball sampling through doxonomic networks}

Because doxonomic campaigns involve interconnected networks of actors, snowball sampling (asking each interviewee to suggest further contacts) is a natural recruitment strategy.
The risk: the sample follows the network's own structure, overrepresenting central actors and missing peripheral or dissenting voices.
Corrective measures include targeted recruitment of actors identified in archival records but not mentioned by earlier interviewees.

\section{Ethical Considerations in Doxonomic Historical Research}
\label{sec:ch18-ethics}

\subsection{The politics of genealogy}

Doxonomic genealogy is inherently political: it traces how beliefs that present themselves as natural truths were, in fact, produced by identifiable actors with identifiable interests.
This creates several ethical obligations:
\begin{itemize}[nosep]
  \item \textbf{Accuracy}: because genealogical claims can damage reputations and delegitimize institutions, the evidentiary standard must be high.
  Every claim should be traceable to specific archival evidence.
  \item \textbf{Symmetry}: genealogical analysis should be applied to all belief-producing institutions, not selectively to those the researcher opposes.
  Progressive foundations, public universities, and NGOs also engage in funded narrative production.
  \item \textbf{Proportionality}: the strength of the genealogical conclusion should be proportional to the evidence.
  ``Corporation X funded institution Y, which published document Z'' is a factual chain.
  ``Corporation X caused belief $p$ to become common sense'' is a causal claim that requires much more evidence.
\end{itemize}

\subsection{Access and power asymmetries}

Archival access is itself a doxonomic variable.
Well-funded institutions can restrict access to their archives, employ legal teams to resist FOIA requests, and control the narrative about their own history.
The tobacco and fossil-fuel archives became available primarily through litigation, not voluntary disclosure.
This creates a systematic bias: the doxonomic campaigns most in need of genealogical scrutiny are precisely those with the greatest capacity to prevent it.

\section{Notes and References}
\label{sec:ch18-notes}

Genealogical method draws on \textcite{foucault1971}; for an accessible introduction, see \textcite{gutting2005}.
The institutional history of neoliberalism is documented in \textcite{mirowski2009}, \textcite{harvey2005}, and \textcite{phillipsfein2009}.
On the Mont P\`{e}lerin Society specifically, see \textcite{slobodian2018}.

Long-duration institutional change is theorized in \textcite{north1990} and \textcite{acemoglu2012}.
The Annales school's \emph{longue dur\'{e}e} framework originates with Braudel (1949); for its application to institutional analysis, see \textcite{pierson2004}.

Comparative-historical methods are developed in \textcite{mahoney2003} and \textcite{skocpol1979}.
Process tracing is formalized in \textcite{bennett2015}.
Archival methods for social science are covered in \textcite{king1994}.

The tobacco industry documents case is analyzed in \textcite{oreskes2010} and Proctor (2012).
On fossil-fuel industry climate denial, see \textcite{oreskes2010} and Supran and Oreskes (2017).
On dark money and archival opacity, see \textcite{mayer2016}.

For mixed-methods research design, see \textcite{creswell2017}.
On oral history methodology, see Ritchie (2014).

\section{Exercises}

\begin{exercise}
Construct a doxonomic genealogy of the belief ``taxation is theft'' from its philosophical origins (Locke, Nozick) to its current institutional support.
Identify at least three institutional channels through which the belief gained popular currency.
What alternative beliefs about taxation did it displace?
\end{exercise}

\begin{exercise}
Using the periodization framework (Model~\ref{mod:periodization}), identify the major doxonomic periods for beliefs about gender roles in the 20th century.
For each transition, specify:
\begin{enumerate}[nosep]
  \item the funding shifts (which institutions gained or lost resources),
  \item the institutional realignments (new organizations, curriculum changes),
  \item the narrative crises (events that destabilized the prior regime),
  \item the measurable belief changes (survey evidence, behavioral indicators).
\end{enumerate}
\end{exercise}

\begin{exercise}
Design an archival research project to estimate the total spending on neoliberal narrative production from 1970 to 2000.
\begin{enumerate}[nosep]
  \item List the specific archival sources you would consult.
  \item Describe the main obstacles to accurate estimation (opacity, definitional ambiguity, international flows).
  \item Propose a lower-bound estimation strategy that uses only publicly available data.
  \item How would you handle the ``dark money'' problem: funding that flows through intermediaries designed to obscure the original source?
\end{enumerate}
\end{exercise}

\begin{exercise}
Apply the archival triangulation model (Model~\ref{mod:triangulation}) to a doxonomic campaign of your choice.
For each of the three archive types (production, infrastructure, reception), identify:
\begin{enumerate}[nosep]
  \item the specific archives or databases you would consult,
  \item what evidence you would look for,
  \item what limitations or gaps you would expect to encounter.
\end{enumerate}
\end{exercise}

\begin{exercise}
Write a complete doxonomic process trace (Model~\ref{mod:process-trace}) for the dissemination of a specific idea from academic publication to policy implementation.
Choose an idea where archival sources are available (e.g., the Laffer Curve, the ``broken windows'' theory of policing, the ``school choice'' movement).
Document each step with specific archival evidence and identify the weakest link in the causal chain.
\end{exercise}

\begin{exercise}
Compare the doxonomic infrastructure for economic policy ideas in two countries (e.g., the US and Germany, or the UK and France).
\begin{enumerate}[nosep]
  \item Map the major think tanks, their funding sources, and their institutional relationships in each country.
  \item Apply Mill's method of difference: how do differences in infrastructure correlate with differences in dominant economic beliefs?
  \item Discuss confounding factors: what other differences between the countries might explain the belief differences?
\end{enumerate}
\end{exercise}

\begin{exercise}
Design an interview protocol for oral history research on a doxonomic campaign.
\begin{enumerate}[nosep]
  \item Specify the target population (who would you interview?).
  \item Write 10 questions that probe the material infrastructure of belief production without leading the interviewee toward a predetermined conclusion.
  \item Describe your strategy for dealing with retrospective bias and strategic non-disclosure.
  \item How would you triangulate interview evidence against archival records?
\end{enumerate}
\end{exercise}
