\chapter{Experimental Doxonomics}
\label{ch:experimental}

\epigraph{The great tragedy of science---the slaying of a beautiful hypothesis by an ugly fact.}{Thomas Henry Huxley}

\noindent
Observational methods identify correlations; experiments identify causes.
The causal inference methods of Chapter~\ref{ch:causal-inference} work backward from observational data; experimental methods work forward from controlled intervention.
This chapter develops experimental designs for testing doxonomic hypotheses in the laboratory, the survey, and the field.

The fundamental doxonomic experiment asks: does exposure to a specific narrative, produced by a specific institution, \emph{cause} a measurable change in belief or behavior?
The question sounds simple.
In practice, it implicates every difficulty in social science experimentation (demand effects, ecological validity, ethical constraints, generalizability), plus several difficulties distinctive to the doxonomic setting, where ``treatments'' are complex discursive objects and ``outcomes'' are partially hidden mental states.

\section{Cooperation Games as Belief Measurement}
\label{sec:ch19-games}

\subsection{The behavioral proxy strategy}

\begin{definition}[Doxonomic Game]
\label{def:doxonomic-game}
\index{doxonomic game}
A \emph{doxonomic game} is an experimental economic game (public goods, dictator, trust, ultimatum) administered after exposure to a doxonomic treatment (narrative about human nature, institutional framing, economic pedagogy, etc.).
The outcome variable is the degree of cooperative or self-interested behavior, interpreted as a \emph{behavioral proxy} for internalized belief about human nature.
\end{definition}

The logic: if doxonomic theory is correct that narratives shape beliefs and beliefs shape behavior, then experimentally manipulating narrative exposure should produce measurable behavioral shifts in the direction predicted by the narrative.
The behavioral proxy strategy avoids the self-report problems discussed in Chapter~\ref{ch:doxometric-measurement}: subjects reveal their beliefs through action rather than declaration.

\subsection{Baseline findings}

The behavioral economics literature provides a crucial baseline.
\textcite{fischbacher2001} found that roughly 50\% of people are conditional cooperators (cooperate when they expect others to cooperate), 30\% are free riders, and 20\% display other patterns.
\textcite{henrich2001} showed that cooperation norms vary across cultures but that pure self-interest is nowhere the modal behavior.

These baselines matter for doxonomics because they establish what behavior looks like \emph{before} doxonomic intervention.
If a narrative treatment shifts the cooperation rate from 50\% to 35\%, we can estimate that the narrative converts roughly 15\% of the population from cooperative to self-interested behavior, a quantifiable doxonomic effect.

\subsection{Game selection}

Different games measure different aspects of belief:

\begin{center}
\begin{tabular}{p{3cm}p{4.5cm}p{4.5cm}}
\toprule
Game & What it measures & Doxonomic relevance \\
\midrule
Public goods & Willingness to contribute to collective benefit & Tests whether narratives about ``free-riding human nature'' reduce contributions \\[4pt]
Trust game & Willingness to make oneself vulnerable to another's choices & Tests whether institutional-distrust narratives reduce social trust \\[4pt]
Dictator game & Unilateral generosity without strategic incentive & Tests whether ``rational self-interest'' narratives reduce altruism \\[4pt]
Ultimatum game & Fairness norms and willingness to punish unfairness & Tests whether ``efficiency'' frames weaken fairness considerations \\[4pt]
Common-pool resource & Restraint in extracting from shared resources & Tests whether ``tragedy of the commons'' narratives become self-fulfilling \\
\bottomrule
\end{tabular}
\end{center}

\subsection{The economics education effect}

A well-documented finding provides natural evidence for doxonomic games: economics students behave more self-interestedly than students in other disciplines.
\textcite{frank1993} found that economics majors defected more in prisoner's dilemma games and contributed less in public goods games.

The causal question is contested: does economics education \emph{cause} selfishness (the indoctrination hypothesis), or do selfish people \emph{select into} economics (the selection hypothesis)?
Longitudinal designs that track students before and after economics courses suggest both mechanisms operate: there is a selection effect, but exposure to the rational-actor model also produces a treatment effect \autocite{bauman1996}.

\begin{example}
\label{ex:econ-education}
A pre-post design: measure cooperation in a public goods game at the beginning of the semester, then again after completing an introductory economics course.
The within-subject treatment effect is $\hat{\tau}_{\text{within}} = \bar{Y}_{\text{post}} - \bar{Y}_{\text{pre}}$.
Compare this to a control group taking introductory psychology over the same period.
The doxonomic effect of economics pedagogy is the DiD:
\[
  \hat{\tau}_{\text{DiD}} = (\bar{Y}^{\text{econ}}_{\text{post}} - \bar{Y}^{\text{econ}}_{\text{pre}}) - (\bar{Y}^{\text{psych}}_{\text{post}} - \bar{Y}^{\text{psych}}_{\text{pre}})
\]
If $\hat{\tau}_{\text{DiD}} < 0$, economics education has a net doxonomic effect of reducing cooperation beyond what would occur from maturation or other common factors.
\end{example}

\section{Priming and Framing Experiments}
\label{sec:ch19-priming}

\subsection{Basic design}

\begin{model}[Narrative Priming Experiment]
\label{mod:priming}
\index{priming experiment}
Randomly assign subjects to:
\begin{itemize}[nosep]
  \item \textbf{Treatment A} (selfishness narrative): read a text asserting that human nature is fundamentally self-interested (drawn from economics textbooks, evolutionary psychology popularizations, or Ayn Rand).
  \item \textbf{Treatment B} (cooperation narrative): read a text asserting that human nature is fundamentally cooperative (drawn from mutual-aid literature, Kropotkin, developmental psychology).
  \item \textbf{Control}: read a text of equal length on an unrelated topic (e.g., geography).
\end{itemize}
Then measure behavior in a public goods game or trust game.
The treatment effects are:
\begin{align}
  \hat{\tau}_A &= \bar{Y}_A - \bar{Y}_{\text{control}} \quad \text{(selfishness narrative effect)} \\
  \hat{\tau}_B &= \bar{Y}_B - \bar{Y}_{\text{control}} \quad \text{(cooperation narrative effect)} \\
  \hat{\tau}_{AB} &= \bar{Y}_A - \bar{Y}_B \quad \text{(frame contrast effect)}
\end{align}
\end{model}

\subsection{Dose-response designs}

A single brief exposure is a weak doxonomic treatment.
Real doxonomic effects accumulate through repeated exposure over extended periods.
Dose-response designs vary the intensity of exposure:

\begin{model}[Narrative Dose-Response]
\label{mod:dose-response}
\index{dose-response}
Assign subjects to $k$ treatment levels $D \in \{0, 1, 2, \ldots, k\}$ representing the number of narrative exposures (e.g., 0, 1, 3, 7, and 14 articles read over two weeks).
The dose-response function is:
\[
  Y(D) = \alpha + f(D) + \varepsilon
\]
where $f(\cdot)$ is estimated nonparametrically (e.g., by including indicators for each dose level) or parametrically (e.g., $f(D) = \beta \log(D + 1)$ for a concave dose-response).

Key questions:
\begin{enumerate}[nosep]
  \item Is there a \emph{threshold}: a minimum dose below which no effect is detectable?
  \item Is the response \emph{concave} (diminishing marginal returns to each additional exposure) or \emph{convex} (accelerating effects, consistent with cascade models)?
  \item Is there a \emph{saturation point} beyond which additional exposure has no further effect?
\end{enumerate}
\end{model}

\subsection{Source credibility manipulation}

The same narrative content may have different effects depending on the perceived credibility and institutional origin of the source.

\begin{model}[Source Attribution Experiment]
\label{mod:source-attribution}
\index{source credibility}
Present identical narrative content to all subjects but vary the attributed source:
\begin{itemize}[nosep]
  \item attributed to a prestigious university,
  \item attributed to a known think tank,
  \item attributed to a corporate-funded research institute,
  \item attributed to an anonymous blog post.
\end{itemize}
The treatment effect difference across conditions estimates the \emph{credibility premium}: how much does institutional prestige amplify (or attenuate) narrative effectiveness?
\end{model}

This design tests the epistemic capital model from Chapter~\ref{ch:epistemic-capital}: institutions with higher credibility should produce larger treatment effects from the same content, and revealing the funding source of a high-credibility institution should reduce its effect (a credibility destruction experiment).

\subsection{Persistence and decay}

A critical doxonomic question is how long experimental effects last.
A priming effect that decays within hours has different policy implications than one that persists for weeks or months.

\begin{model}[Persistence Design]
\label{mod:persistence}
\index{persistence}
After narrative exposure at time $t_0$, measure the outcome at multiple follow-up times $t_1, t_2, \ldots, t_m$:
\[
  \hat{\tau}(t_s) = \bar{Y}_{\text{treat}}(t_s) - \bar{Y}_{\text{control}}(t_s), \quad s = 1, \ldots, m.
\]
The decay function $\hat{\tau}(t)$ characterizes how quickly the effect fades.
Doxonomic theory predicts that single exposures produce rapidly decaying effects, but repeated exposures (as in the dose-response design) produce effects that persist longer, because repetition transitions the belief from working memory to long-term schema (cf.\ the internalization model in Chapter~\ref{ch:belief-production-chain}).
\end{model}

\begin{example}
\label{ex:persistence}
A hypothetical persistence study:
\begin{center}
\begin{tabular}{lccccc}
\toprule
& Immediate & 1 day & 1 week & 1 month & 3 months \\
\midrule
Single exposure & $-4.2^{**}$ & $-2.8^{*}$ & $-1.1$ & $-0.3$ & $+0.1$ \\
Seven exposures & $-6.1^{***}$ & $-5.4^{***}$ & $-4.8^{**}$ & $-3.2^{*}$ & $-2.1$ \\
\bottomrule
\end{tabular}
\end{center}
Effects on public goods contributions (percentage point change from control).
Stars denote significance: $^{*}p < 0.05$, $^{**}p < 0.01$, $^{***}p < 0.001$.
Single exposure decays to insignificance within a week; seven exposures persist for at least a month, consistent with the schema-internalization model.
(Illustrative data.)
\end{example}

\section{Norm Perception Experiments}
\label{sec:ch19-norms}

\subsection{The norm channel}

\begin{definition}[Perceived Social Norm]
\label{def:perceived-norm}
\index{perceived norm}
The \emph{perceived social norm} for behavior $x$ is agent $j$'s belief about the population's behavior: $\hat{\mu}_j(x) = \E_j[\bar{x}]$.
Doxonomic systems can shift behavior by shifting norm perceptions: making people think others are more selfish (or more cooperative) than they actually are.
\end{definition}

This channel is distinct from direct belief change.
A subject may not change their own moral convictions but still change their behavior because they believe the social environment has changed.
If I believe everyone else is defecting, conditional cooperation logic says I should defect too, not because I've become selfish, but because cooperation without reciprocity is exploitation.

\begin{proposition}[Norm Misperception Effect]
\label{prop:norm-misperception}
\index{norm misperception}
If agents are conditional cooperators (cooperate when they believe others cooperate), then a doxonomic intervention that increases $\hat{\mu}_j(\text{selfish})$ without changing actual behavior will reduce cooperation in subsequent interactions.
Formally, let $c_j = g(\hat{\mu}_j)$ be $j$'s cooperation level as a function of perceived norms, with $g' > 0$.
A doxonomic shift $\hat{\mu}_j \to \hat{\mu}_j + \delta$ (where $\delta$ increases perceived selfishness) reduces cooperation by $\Delta c_j \approx g'(\hat{\mu}_j) \cdot \delta$.
If agents observe each other's behavior, the reduced cooperation confirms the misperception, creating a \emph{self-fulfilling belief}:
\[
  \hat{\mu} \uparrow \;\Rightarrow\; c \downarrow \;\Rightarrow\; \text{observed selfishness} \uparrow \;\Rightarrow\; \hat{\mu} \uparrow \;\Rightarrow\; \cdots
\]
The system can tip from a cooperative equilibrium to a defection equilibrium without any change in underlying preferences.
\end{proposition}

\subsection{Measuring the norm channel}

To isolate the norm channel from direct belief change, we need designs that manipulate norm perceptions independently of personal beliefs.

\begin{model}[Norm Information Experiment]
\label{mod:norm-info}
\index{norm information experiment}
\begin{enumerate}[nosep]
  \item Measure subjects' personal beliefs about human nature (survey).
  \item Randomly provide accurate or distorted information about others' behavior:
  \begin{itemize}[nosep]
    \item \textbf{Accurate}: ``In previous sessions, 65\% of participants cooperated.''
    \item \textbf{Pessimistic distortion}: ``In previous sessions, only 25\% cooperated.''
    \item \textbf{Optimistic distortion}: ``In previous sessions, 90\% cooperated.''
  \end{itemize}
  \item Measure behavior in a public goods game.
  \item Re-measure personal beliefs (post-test).
\end{enumerate}
If behavior changes in step 3 but personal beliefs do not change in step 4, the effect operates through the norm channel.
If both change, the norm information also updates personal beliefs (a stronger doxonomic effect).
\end{model}

\begin{example}
\label{ex:norm-gap}
Suppose a norm perception study finds that 78\% of people believe ``most people are selfish,'' but behavioral games show only 30\% defect.
The \emph{norm misperception gap} is $0.78 - 0.30 = 0.48$: perceived selfishness exceeds actual selfishness by 48 percentage points.
This gap represents the doxonomic ``opportunity,'' the space in which narratives about human selfishness have shifted perceptions beyond reality.

The doxonomic prediction: closing the gap (by providing accurate norm information) should increase cooperation.
Several studies have found exactly this effect: providing accurate information about others' pro-social behavior increases cooperation, recycling, energy conservation, and charitable giving \autocite{cialdini2006}.
\end{example}

\section{Field Experiments}
\label{sec:ch19-field}

\subsection{Why laboratory results may not generalize}

Laboratory doxonomic experiments face threats to \emph{ecological validity}:
\begin{itemize}[nosep]
  \item subjects know they are in an experiment (demand effects),
  \item exposure is brief and isolated from the media environment,
  \item the stake sizes in games are small relative to real-world decisions,
  \item the subject pool (often university students) is not representative.
\end{itemize}

Field experiments address these threats by embedding doxonomic treatments in real-world settings.

\subsection{Media field experiments}

\begin{model}[Randomized Media Exposure]
\label{mod:media-field}
\index{media field experiment}
Recruit a panel of participants.
Randomly assign media diet modifications:
\begin{itemize}[nosep]
  \item \textbf{Treatment}: provide free subscription to a newspaper with known editorial stance (or pay subjects to read/watch specified media).
  \item \textbf{Control}: no media modification (or subscription to a neutral outlet).
\end{itemize}
Measure beliefs and behaviors at baseline, during treatment, and at follow-up intervals.

Design considerations:
\begin{enumerate}[nosep]
  \item \emph{Compliance}: subjects may not actually consume the assigned media. Intent-to-treat (ITT) and local average treatment effect (LATE) analyses are both needed.
  \item \emph{Contamination}: subjects receive media from many sources. The treatment adds to, rather than replaces, their existing media diet.
  \item \emph{Duration}: meaningful doxonomic effects may require months of exposure, making field experiments expensive.
\end{enumerate}
\end{model}

\begin{example}
\label{ex:media-field}
\textcite{dellavigna2007} used cable channel positioning as a natural experiment for Fox News exposure.
A randomized field experiment could improve on this:
\begin{itemize}[nosep]
  \item Recruit 2,000 participants in media markets without Fox News.
  \item Randomly give 1,000 free cable packages that include Fox News.
  \item Survey beliefs about economic policy, immigration, and government trust at baseline, 3 months, 6 months, and 12 months.
  \item Measure behavioral outcomes: voting behavior, charitable giving patterns, community participation.
\end{itemize}
The ITT estimate captures the effect of \emph{offering} the channel; the LATE (instrumenting actual viewership with random assignment) captures the effect of \emph{watching} it.
\end{example}

\subsection{Educational field experiments}

\begin{model}[Curriculum Experiment]
\label{mod:curriculum}
\index{curriculum experiment}
Randomly assign classrooms or schools to different pedagogical approaches for the same content area:
\begin{itemize}[nosep]
  \item \textbf{Treatment A}: economics course using standard textbook (emphasizing rational self-interest, efficiency, market equilibrium).
  \item \textbf{Treatment B}: economics course using behavioral/institutional textbook (emphasizing bounded rationality, cooperation, market failures).
  \item \textbf{Control}: no economics course (alternative elective).
\end{itemize}
Outcome measures: beliefs about human nature, economic beliefs, cooperative behavior in games, and real-world behavioral indicators (volunteering, political participation).
Follow-up at end of course, 1 year later, and (ideally) 5 years later.
\end{model}

This design directly tests the doxonomic claim that educational institutions are sites of belief production.
The randomization overcomes the selection problem that plagues observational studies of economics education.

\subsection{Information correction experiments}

\begin{model}[Doxonomic Correction Experiment]
\label{mod:correction}
\index{correction experiment}
Test whether providing counter-narratives or factual corrections can undo doxonomic effects:
\begin{enumerate}[nosep]
  \item Establish baseline beliefs.
  \item Expose all subjects to a doxonomic narrative (e.g., ``most economists agree that minimum wages destroy jobs'').
  \item Randomly assign correction:
  \begin{itemize}[nosep]
    \item \textbf{No correction} (control),
    \item \textbf{Factual correction} (``the evidence on minimum wages and employment is mixed''),
    \item \textbf{Source disclosure} (``this claim was funded by the National Restaurant Association''),
    \item \textbf{Both} correction and source disclosure.
  \end{itemize}
  \item Measure beliefs and behavioral intentions.
\end{enumerate}
This design estimates the ``de-doxonomification'' effect: how much belief recovery is possible after narrative exposure?
\end{model}

A key finding from the misinformation literature is that corrections often fail, especially when the original misinformation is congruent with prior beliefs \autocite{nyhan2010}.
The \emph{backfire effect} (corrections strengthening the original misbelief) has been documented in some (though not all) settings.
If doxonomic narratives are similarly resistant to correction, this has profound implications for counter-doxonomic policy.

\section{Survey Experiments}
\label{sec:ch19-survey}

\subsection{Conjoint designs for doxonomic attributes}

Conjoint experiments present respondents with profiles that vary multiple attributes simultaneously, allowing estimation of the causal effect of each attribute on choices or evaluations.

\begin{model}[Doxonomic Conjoint]
\label{mod:conjoint}
\index{conjoint experiment}
Present respondents with pairs of policy proposals that vary on:
\begin{itemize}[nosep]
  \item policy content (identical across conditions),
  \item source institution (think tank, university, government agency),
  \item funding disclosure (corporate-funded, foundation-funded, government-funded, undisclosed),
  \item endorsement (endorsed by economist, endorsed by community leader, no endorsement).
\end{itemize}
Ask respondents to choose which proposal they prefer.
The \emph{average marginal component effect} (AMCE) for each attribute estimates how much that attribute affects preference, holding all others constant.
\end{model}

This design estimates the ``credibility premium'' of institutional source and the ``credibility penalty'' of funding disclosure, directly testing the epistemic capital model.

\subsection{List experiments for sensitive beliefs}

Some doxonomically relevant beliefs are socially sensitive: respondents may not truthfully report racist beliefs, extreme distrust of institutions, or conspiracy theories.
List experiments (the item count technique) allow estimation of prevalence without requiring individual admission:

\begin{model}[List Experiment for Sensitive Beliefs]
\label{mod:list-experiment}
\index{list experiment}
Present respondents with a list of $k$ items and ask how many (not which) they agree with.
\begin{itemize}[nosep]
  \item \textbf{Control}: $k$ non-sensitive items.
  \item \textbf{Treatment}: same $k$ items plus one sensitive item.
\end{itemize}
The difference in means estimates the population prevalence of the sensitive belief:
\[
  \hat{\pi} = \bar{Y}_{\text{treat}} - \bar{Y}_{\text{control}}.
\]
This can be combined with doxonomic treatments (e.g., preceding the list experiment with narrative exposure) to estimate the effect of narrative on sensitive beliefs without direct questioning.
\end{model}

\section{Online and Platform Experiments}
\label{sec:ch19-online}

\subsection{Advantages and limitations}

Online platforms have become a major site for doxonomic experiments due to:
\begin{itemize}[nosep]
  \item large, diverse subject pools (Amazon Mechanical Turk, Prolific, Lucid),
  \item rapid iteration and large sample sizes,
  \item natural embedding in digital media environments,
  \item ability to simulate realistic media exposure.
\end{itemize}

Limitations include: lower attention and engagement than laboratory settings, inability to verify identity, self-selection into platforms, and the ``professional respondent'' problem.

\subsection{A/B testing as doxonomic evidence}

Technology companies routinely run A/B tests that are, in effect, large-scale doxonomic experiments: randomly varying the content shown to users and measuring behavioral responses.
These experiments generate valuable evidence about the causal effects of narrative exposure, but the results are typically proprietary.

\begin{proposition}[Platform Experiment Transparency]
\label{prop:platform-transparency}
The doxonomic information generated by platform A/B tests is a \emph{public good}: it reveals how narrative exposure affects belief and behavior at unprecedented scale.
Its private custody by platform companies constitutes an information asymmetry with doxonomic consequences: the companies that produce the most powerful narrative infrastructure are also the only ones who know how it works.
\end{proposition}

\subsection{Researcher-platform collaborations}

A growing number of studies use researcher-platform collaborations to overcome the access problem.
Facebook's ``social media and democracy'' research program, Twitter's (now X's) academic API, and YouTube's researcher access program have all produced doxonomically relevant experiments, though questions about independence and selective publication remain.

\section{Power Analysis and Sample Size}
\label{sec:ch19-power}

\subsection{Doxonomic effect sizes}

Experimental doxonomic effects are typically small to moderate by behavioral science standards.
A single narrative exposure might shift behavior by $d = 0.1$ to $0.3$ standard deviations.
Accumulated effects (multiple exposures, institutional-level treatments) may be larger, but they are harder to study experimentally.

\begin{proposition}[Sample Size for Doxonomic Experiments]
\label{prop:sample-size}
\index{power analysis}
For a two-group comparison detecting a standardized effect of $d$ with power $1 - \beta$ at significance level $\alpha$:
\[
  n_{\text{per group}} = \frac{2(z_{1-\alpha/2} + z_{1-\beta})^2}{d^2}.
\]
For $d = 0.2$ (a small-to-moderate doxonomic effect), $\alpha = 0.05$, power $= 0.80$:
\[
  n = \frac{2(1.96 + 0.84)^2}{0.04} = \frac{2(7.84)}{0.04} = 392 \text{ per group}.
\]
Multi-arm designs (3+ treatment conditions), moderation analyses (heterogeneous treatment effects), and mediation analyses (testing the norm channel vs.\ direct belief change) require substantially larger samples.
\end{proposition}

\subsection{Pre-registration}

Given the flexibility in doxonomic experiments (many possible outcomes, subgroups, model specifications), pre-registration is essential to prevent $p$-hacking.
A doxonomic experiment should pre-register:
\begin{enumerate}[nosep]
  \item primary outcome(s) and the direction of the predicted effect,
  \item secondary outcomes (clearly labeled as exploratory),
  \item planned subgroup analyses with theoretical justification,
  \item the statistical model (ITT vs.\ LATE, covariates, clustering),
  \item the multiple-testing correction procedure.
\end{enumerate}

\section{Ethical Constraints and Institutional Review}
\label{sec:ch19-ethics}

\subsection{The harm asymmetry}

Experimental doxonomics faces a distinctive ethical tension.
Standard IRB criteria ask whether the experiment could harm subjects.
A doxonomic experiment that successfully makes subjects more selfish ``works'' precisely by making them worse off (or making others worse off through their changed behavior).
Conversely, an experiment that makes subjects more cooperative or better-informed might be beneficial.

This creates an asymmetry: the experiments that would provide the strongest evidence for doxonomic theory (demonstrating that narrative exposure causes harmful belief change) are precisely those that are hardest to justify ethically.

\subsection{Specific ethical concerns}

\begin{enumerate}
  \item \textbf{Lasting belief effects.}
  If a brief experimental exposure can shift beliefs, the shift might persist beyond the experiment.
  Debriefing may not fully restore pre-experimental beliefs (the ``you can't un-ring a bell'' problem).

  \item \textbf{Deception about the social environment.}
  Norm perception experiments require providing false information about others' behavior.
  This is a form of deception that standard protocols require justification for and debriefing about.

  \item \textbf{Vulnerable populations.}
  If doxonomic effects are strongest on those with weak prior beliefs (as the formal model predicts), experimentally targeting such populations raises heightened ethical concerns.

  \item \textbf{Dual use.}
  Experimental knowledge about effective doxonomic techniques could be used by powerful actors to improve their narrative campaigns.
  The same research that reveals how beliefs are shaped could serve as an instruction manual for more effective manipulation.
\end{enumerate}

\subsection{Mitigation strategies}

\begin{itemize}[nosep]
  \item \textbf{Minimal deception}: use real rather than fabricated narrative materials where possible; attribute content accurately.
  \item \textbf{Thorough debriefing}: provide subjects with accurate information after the experiment, including the hypothesized direction of the effect.
  \item \textbf{Follow-up measurement}: check whether beliefs have returned to baseline at a follow-up assessment.
  \item \textbf{Equipoise}: design experiments where the direction of the effect is genuinely uncertain (the researcher doesn't know which treatment is ``harmful'').
  \item \textbf{Natural and observational alternatives}: when ethical constraints are binding, use natural experiments (Chapter~\ref{ch:causal-inference}) or observational designs that exploit existing variation in narrative exposure.
\end{itemize}

\section{Replication and Meta-Analysis}
\label{sec:ch19-replication}

\subsection{The replication crisis and doxonomic experiments}

Social psychology's replication crisis has particular relevance for experimental doxonomics.
Many priming effects that seemed robust in initial studies failed to replicate in larger, pre-registered replications.
Doxonomic experiments should be designed with replication in mind from the outset:
\begin{itemize}[nosep]
  \item pre-registered protocols with publicly available materials,
  \item sample sizes powered to detect realistic (not optimistic) effect sizes,
  \item multi-site designs that test generalizability across populations,
  \item sharing of raw data and analysis code.
\end{itemize}

\subsection{Meta-analytic synthesis}

As the experimental doxonomics literature grows, meta-analysis becomes the tool for synthesizing effects across studies.
Key questions for meta-analysis:
\begin{enumerate}[nosep]
  \item What is the average effect of narrative exposure on behavior across studies?
  \item Does the effect vary by narrative type, exposure duration, subject population, or outcome measure?
  \item Is there evidence of publication bias (selective reporting of significant results)?
  \item What moderates the effect: prior beliefs, education level, source credibility?
\end{enumerate}

A meta-analytic estimate of the doxonomic effect size, even if small ($d \approx 0.15$), would be substantively important if it represents the effect of a \emph{single} exposure.
The cumulative effect of thousands of exposures over a lifetime (the realistic doxonomic scenario) could be vastly larger, though extrapolation from single-exposure experiments to cumulative effects requires the dose-response models of Section~\ref{sec:ch19-priming}.

\section{Notes and References}
\label{sec:ch19-notes}

Public goods experiments are reviewed in \textcite{fischbacher2001} and Chaudhuri (2011).
Cross-cultural behavioral games are in \textcite{henrich2001}.
The economics education effect is documented in \textcite{frank1993} and \textcite{bauman1996}.

Priming effects on economic behavior are surveyed in \textcite{kahneman2011}.
On norm misperception and social influence, see \textcite{cialdini2006} and Bicchieri (2005).
Media effects experiments are reviewed in \textcite{dellavigna2010}.

The misinformation correction literature includes \textcite{nyhan2010} and Wood and Porter (2019).
On pre-registration and the replication crisis, see Nosek et al.\ (2015) and Open Science Collaboration (2015).
Experimental ethics in social science is discussed in \textcite{salganik2018}.

For conjoint experimental designs, see Hainmueller et al.\ (2014).
List experiments are developed in Blair and Imai (2012).
On platform experiments and transparency, see Guess et al.\ (2023).

\section{Exercises}

\begin{exercise}
Design a priming experiment to test whether exposure to ``greed is good'' narratives increases defection in a public goods game.
Specify: sample size (with power analysis), treatment arms, randomization procedure, primary and secondary outcome measures, and pre-registration plan.
\end{exercise}

\begin{exercise}
A norm perception study finds that 78\% of people believe ``most people are selfish,'' but behavioral games show only 30\% defect.
\begin{enumerate}[nosep]
  \item Compute the norm misperception gap.
  \item Using the conditional cooperation model ($c_j = g(\hat{\mu}_j)$ with $g(\hat{\mu}) = 1 - \hat{\mu}$), predict the equilibrium cooperation rate if agents update their norm perceptions based on observed behavior.
  \item Does the system converge to a cooperative or defection equilibrium?
  \item What intervention could tip the system toward the cooperative equilibrium?
\end{enumerate}
\end{exercise}

\begin{exercise}
Discuss the ethics of an experiment that exposes children to competing anthropological narratives (selfish vs.\ cooperative) and measures long-term behavioral effects.
\begin{enumerate}[nosep]
  \item Under what conditions (if any) would this be justifiable?
  \item What would a thorough debriefing look like for children?
  \item How does this experiment compare ethically to the actual practice of teaching economics (which already exposes students to the rational self-interest narrative without IRB approval)?
\end{enumerate}
\end{exercise}

\begin{exercise}
Design a source credibility experiment (Model~\ref{mod:source-attribution}) with four source conditions.
\begin{enumerate}[nosep]
  \item Choose a specific policy claim as the content.
  \item Specify the four source attributions and justify their selection.
  \item Predict the ordering of treatment effects based on the epistemic capital model.
  \item What would it mean for the epistemic capital model if the ordering came out differently?
\end{enumerate}
\end{exercise}

\begin{exercise}
Design a dose-response experiment (Model~\ref{mod:dose-response}) to estimate the cumulative effect of narrative exposure.
\begin{enumerate}[nosep]
  \item Specify the dose levels and the delivery schedule (how are multiple exposures spaced?).
  \item Discuss the tradeoff between dose intensity and ecological validity.
  \item How would you test for a threshold effect vs.\ a gradual dose-response?
  \item What sample size would you need to estimate the dose-response curve with adequate precision at each level?
\end{enumerate}
\end{exercise}

\begin{exercise}[Computational]
Simulate the norm misperception cascade from Proposition~\ref{prop:norm-misperception}:
\begin{enumerate}[nosep]
  \item Create 100 agents with initial cooperation probability drawn from $\text{Beta}(3, 1.5)$ (most agents tend toward cooperation).
  \item Each agent observes a biased signal of group behavior: $\hat{\mu}_j = (1-\lambda) \cdot \mu_{\text{actual}} + \lambda \cdot \mu_{\text{media}}$, where $\mu_{\text{media}} = 0.7$ (media suggests 70\% are selfish) and $\lambda = 0.3$ (30\% weight on media signal).
  \item Agents update cooperation based on $c_j^{(t+1)} = c_j^{(0)} \cdot (1 - \hat{\mu}_j^{(t)})$.
  \item Iterate for 50 rounds. Plot the average cooperation rate over time.
  \item At what value of $\lambda$ does the system tip from the cooperative to the defection equilibrium?
\end{enumerate}
\end{exercise}
