\chapter{Meritocracy}
\label{ch:meritocracy}

\epigraph{The more a society proclaims its meritocracy, the more its winners can justify their position and the more its losers blame themselves.}{Michael Sandel, \textit{The Tyranny of Merit}}

\noindent
Meritocracy is the belief that outcomes reflect effort and talent.
It is also one of the most heavily subsidized narratives in modern capitalism, functioning as the primary legitimation mechanism for inequality.
Where the selfishness regime (Chapter~\ref{ch:human-nature}) shapes beliefs about what humans \emph{are}, the meritocratic regime shapes beliefs about what humans \emph{deserve}: those who succeed deserve their success; those who fail deserve their failure; and the distribution of rewards, however unequal, is fundamentally just.

This chapter applies the full doxonomic toolkit---genealogy, institutional analysis, formal modeling, cross-cultural comparison---to the meritocratic regime.
The analysis reveals a belief system that is empirically questionable, institutionally constructed, and functionally essential to the political stability of societies with extreme inequality.

\section{The Meritocratic Belief System}
\label{sec:ch22-system}

\subsection{Structure of the regime}

\begin{definition}[Meritocratic Regime]
\label{def:meritocratic-regime}
\index{meritocracy}
The \emph{meritocratic regime} is an ideology $\mathcal{P}_M$ containing four mutually reinforcing propositions:
\begin{enumerate}[nosep]
  \item $p_1$: Success reflects ability and effort (the \emph{desert} claim).
  \item $p_2$: Opportunity is broadly equal (the \emph{access} claim).
  \item $p_3$: Failure reflects personal deficiency (the \emph{blame} claim).
  \item $p_4$: Inequality is therefore justified (the \emph{legitimation} claim).
\end{enumerate}
These propositions form a coherent bundle: high credence in $p_1$ raises credence in $p_3$ and $p_4$.
The regime functions as a \emph{just-world ideology}: it provides a moral framework in which the existing distribution of outcomes is deserved.
\end{definition}

The logical structure is worth noting: $p_4$ follows from $p_1$--$p_3$ only if the background conditions hold.
If opportunity is \emph{not} equal ($p_2$ is false), then success does not reflect desert alone, failure does not reflect deficiency alone, and the legitimation claim collapses.
The empirical question is therefore whether $p_2$ holds---whether opportunity is in fact broadly equal.

\subsection{The empirical record}

The evidence overwhelmingly contradicts $p_2$.
Intergenerational income elasticity in the United States is approximately 0.47: roughly half of a parent's income advantage is transmitted to their children \autocite{chetty2014}.
A child born in the top income quintile has a 40\% chance of remaining there as an adult; a child born in the bottom quintile has a 7.5\% chance of reaching the top.
The ``American Dream'' of unlimited upward mobility is a statistical rarity, not the norm.

\begin{center}
\begin{tabular}{lcc}
\toprule
Country & Intergenerational elasticity & Rank \\
\midrule
Denmark & 0.15 & High mobility \\
Canada & 0.19 & High mobility \\
Germany & 0.32 & Moderate \\
United States & 0.47 & Low mobility \\
United Kingdom & 0.50 & Low mobility \\
\bottomrule
\end{tabular}
\end{center}

The paradox is striking: the United States, the country most committed to meritocratic ideology, has among the lowest actual mobility in the developed world.
The doxonomic hypothesis: meritocratic belief is strongest not where mobility is highest, but where \emph{narrative investment in meritocracy} is highest.
High inequality requires more legitimation effort, and the US has the institutional infrastructure to provide it.

\subsection{The psychological function}

Meritocratic belief serves a psychological function beyond political legitimation.
Just-world belief \autocite{lerner1980} reduces the anxiety of living in an uncertain and unequal world: if outcomes are deserved, then I am safe as long as I am virtuous and hardworking.
This psychological appeal makes the meritocratic narrative ``sticky''---people are motivated to believe it, not just persuaded by evidence for it.

The doxonomic implication: meritocratic belief has both a ``supply side'' (institutional production) and a ``demand side'' (psychological motivation).
The narrative succeeds because institutions produce it \emph{and} because audiences want to believe it.
This dual anchoring makes it especially resistant to counter-evidence.

\section{Institutional Repetition}
\label{sec:ch22-repetition}

The meritocratic regime is reinforced through a network of institutions that treat meritocratic assumptions as operational truth.

\subsection{Education}

Grading, ranking, credentialing, standardized testing, and academic competition constitute an immersive meritocratic environment for children and young adults.
The educational system performs two doxonomic functions simultaneously:
\begin{enumerate}[nosep]
  \item \textbf{Sorting function}: it actually does reward some abilities and efforts (reading comprehension, test-taking skills, persistence), creating a genuine correlation between performance and outcomes.
  \item \textbf{Laundering function}: it transforms inherited advantages (parental education, income, social networks, test preparation resources) into credentials that \emph{appear} to reflect pure merit.
\end{enumerate}

The meritocratic narrative emphasizes the sorting function and denies the laundering function.
But the laundering function is massive: SAT scores correlate with family income at $r \approx 0.4$, and the correlation between parental education and child's educational attainment is even higher.

\subsection{Corporate culture}

\begin{model}[Corporate Meritocratic Reproduction]
\label{mod:corporate-meritocracy}
\index{corporate meritocracy}
Corporate ``talent management'' systems reproduce meritocratic belief through:
\begin{enumerate}[nosep]
  \item \emph{Hiring}: elite credentials (from elite universities, whose admissions correlate with family wealth) as merit proxies.
  \item \emph{Performance review}: individual evaluation systems that attribute team outcomes to individual effort, ignoring structural advantages.
  \item \emph{Promotion}: ``up or out'' systems that frame career outcomes as reflecting individual talent.
  \item \emph{Compensation}: extreme pay ratios (CEO-to-worker ratio $\sim$350:1 in the US) justified as reflecting proportional contribution.
\end{enumerate}
Each practice makes meritocratic assumptions operational, generating behavioral data that appears to confirm the assumptions.
\end{model}

\subsection{Media}

The media meritocratic infrastructure includes:

\paragraph{Rags-to-riches narratives.}
Disproportionate coverage of exceptional upward mobility (Oprah, Steve Jobs, Jay-Z) creates availability bias about its frequency.
The narrative structure---individual overcomes adversity through talent and perseverance---is compelling precisely because it is rare; but the sheer volume of such narratives makes it seem typical.

\paragraph{Billionaire hagiography.}
Biographies, profiles, and documentaries about the ultra-wealthy overwhelmingly attribute their success to vision, hard work, and risk-taking rather than to inherited wealth, network advantages, regulatory capture, or luck.
The genre conventions of business journalism enforce this framing: the ``founder story'' is always a story of individual genius.

\paragraph{Failure as spectacle.}
Conversely, poverty and failure are framed as individual pathology: addiction, laziness, bad decisions.
The structural conditions that produce poverty (deindustrialization, housing markets, healthcare costs, racial discrimination) are backgrounded in favor of individual narratives.

\subsection{Policy}

Means-testing, workfare, conditional welfare, drug testing for benefit recipients, and ``personal responsibility'' rhetoric all operationalize the meritocratic assumption that poverty reflects deficient character.
These policy designs simultaneously:
\begin{itemize}[nosep]
  \item \emph{signal} that the state views the poor as undeserving (reinforcing the narrative),
  \item \emph{impose costs} that make poverty harder to escape (undermining actual mobility),
  \item \emph{generate data} (low take-up rates, compliance failures) that appear to confirm the deficiency narrative.
\end{itemize}

\subsection{Self-help industry}

The self-help and motivational industry---annual revenues exceeding \$13 billion in the US---is perhaps the purest expression of meritocratic doxonomics.
Its core message is that success is a matter of mindset: ``think and grow rich,'' ``the secret,'' ``grit,'' ``growth mindset.''
Structural barriers are reframed as mental barriers; political problems become personal development opportunities \autocite{ehrenreich2009}.

\section{Credential Systems as Epistemic Capital Laundering}
\label{sec:ch22-credentials}

\subsection{The laundering mechanism}

\begin{proposition}[Credential Laundering]
\label{prop:credential}
\index{credential laundering}
Credential systems (degrees, certifications, rankings) convert inherited advantage into apparent merit.
If access to credentials correlates with parental wealth ($r > 0$), then the credential $c_j$ for agent $j$ satisfies:
\[
  c_j = f(\text{ability}_j, \text{effort}_j) + g(\text{wealth}_j) + \varepsilon_j
\]
where $f$ is the genuine merit component and $g$ is the laundering function.
The meritocratic narrative claims $g \equiv 0$.
The doxonomic analysis estimates $g$ and reveals the laundering magnitude.
\end{proposition}

\subsection{Decomposing credentials}

\begin{example}
\label{ex:credential-decomposition}
Consider admission to a selective university.
The credential (admission) is a function of:
\begin{itemize}[nosep]
  \item SAT score (correlates with family income at $r \approx 0.4$),
  \item GPA (correlates with school quality, which correlates with neighborhood wealth),
  \item extracurricular activities (correlates with parental time and money---fencing, oboe, summer research internships),
  \item legacy status (direct inheritance of institutional advantage),
  \item recommendation letters (correlate with access to well-connected adults).
\end{itemize}

A rough variance decomposition: ability and effort explain $\sim$30--40\% of the variance in selective college admission; family background explains $\sim$30--40\%; institutional factors (legacy, geography, recruited athlete status) explain $\sim$20--30\%.
The meritocratic narrative attributes 100\% to the first component.
The laundering function $g$ accounts for the remaining 60--70\%.
\end{example}

\subsection{The credibility premium of credentials}

Credentials not only sort individuals; they confer \emph{epistemic capital} (Chapter~\ref{ch:epistemic-capital}).
A policy recommendation from a Harvard economist carries more weight than the same recommendation from a community college instructor---not because the content differs, but because the credential signals membership in a high-credibility institution.
This credibility premium amplifies the meritocratic regime: those who succeed in the credentialing system gain disproportionate influence over the narratives that define what counts as merit.

\section{A Formal Model of Meritocratic Belief}
\label{sec:ch22-model}

\subsection{Belief as a function of investment and visibility}

\begin{model}[Meritocratic Belief Production]
\label{mod:meritocracy}
\index{meritocratic belief!model}
The strength of meritocratic belief $\belief{M}$ in a population is a function of narrative investment $\phi_M$ and inequality visibility $V$:
\[
  \belief{M} = \frac{\phi_M}{\phi_M + \kappa \cdot V}
\]
where $\kappa > 0$ measures the sensitivity of belief to visible inequality.
Properties:
\begin{itemize}[nosep]
  \item $\belief{M} \to 1$ as $\phi_M \to \infty$: with sufficient investment, any level of inequality can be legitimated.
  \item $\belief{M} \to 0$ as $V \to \infty$: if inequality becomes infinitely visible, no narrative investment suffices.
  \item The elasticity of belief with respect to investment is $\varepsilon_{\phi} = \kappa V / (\phi_M + \kappa V)$: investment is most effective when inequality is moderate.
\end{itemize}
\end{model}

\subsection{The legitimation cost of inequality}

\begin{proposition}[Legitimation Cost]
\label{prop:legitimation-cost}
\index{legitimation cost}
To maintain meritocratic belief at level $\belief{M}^*$ as inequality grows, the required investment is:
\[
  \phi_M^* = \frac{\belief{M}^*}{1 - \belief{M}^*} \cdot \kappa V.
\]
This is linear in $V$: a doubling of visible inequality requires a doubling of narrative investment to maintain the same level of meritocratic belief.
\end{proposition}

\begin{example}
\label{ex:legitimation-cost}
If $\kappa = 1$, $\belief{M}^* = 0.7$, and inequality visibility is $V = 5$ (on an arbitrary scale):
\[
  \phi_M^* = \frac{0.7}{0.3} \cdot 1 \cdot 5 = 11.67 \text{ (investment units)}.
\]
If inequality doubles ($V = 10$):
\[
  \phi_M^* = \frac{0.7}{0.3} \cdot 10 = 23.33.
\]
The required investment doubles.
This implies that societies with rapidly rising inequality (the US since the 1980s) must \emph{increase} their meritocratic narrative investment to maintain the same belief level---and indeed, the growth of the self-help industry, business media, and meritocratic policy rhetoric tracks the growth of inequality quite closely.
\end{example}

\subsection{Dynamic model with inequality feedback}

The static model becomes dynamic when we add a feedback loop: meritocratic belief affects redistributive policy, which affects inequality:

\begin{model}[Meritocracy-Inequality Feedback]
\label{mod:merit-inequality-feedback}
\index{meritocracy!feedback model}
\begin{align}
  \frac{d\belief{M}}{dt} &= \alpha \phi_M (1 - \belief{M}) - \beta V(t) \belief{M} \label{eq:merit-dynamics} \\
  \frac{dV}{dt} &= \gamma (1 - \belief{M}(t))^{-1} - \delta \label{eq:inequality-dynamics}
\end{align}
where Equation~\eqref{eq:merit-dynamics} says belief grows with investment and decays with visible inequality, and Equation~\eqref{eq:inequality-dynamics} says inequality grows faster when meritocratic belief blocks redistribution (high $\belief{M}$ reduces political support for taxation) and decays naturally at rate $\delta$ (through growth, technological diffusion, etc.).

This system can produce a \emph{vicious spiral}: high $\belief{M}$ blocks redistribution $\to$ inequality grows $\to$ more investment needed to maintain $\belief{M}$ $\to$ higher $\phi_M$ $\to$ maintained $\belief{M}$ $\to$ continued inequality growth.
The spiral continues until either narrative investment hits a budget constraint or inequality becomes so extreme that lived experience overwhelms narrative.
\end{model}

\section{The Doxonomic Genealogy of Meritocracy}
\label{sec:ch22-genealogy}

\subsection{Origins of the concept}

The term ``meritocracy'' was coined satirically by Michael Young in \emph{The Rise of the Meritocracy} (1958) as a \emph{dystopian warning}: a society that sorts people by ``merit'' will create a new aristocracy that is even more self-righteous than the old one, because it believes it earned its position.
The transformation of ``meritocracy'' from a satirical coinage into a term of praise is itself a doxonomic achievement of the first order.

\subsection{Institutional genealogy}

\paragraph{Standardized testing (1900s--).}
IQ tests (Binet, 1905), the SAT (1926), and their successors created the apparatus of ``objective'' merit measurement.
The tests correlated heavily with family background but were presented as measuring pure cognitive ability.
The testing industry (College Board, ETS) became a self-sustaining institutional pillar of the meritocratic regime, with annual revenues exceeding \$1 billion.

\paragraph{Human capital theory (1960s--).}
\textcite{becker1964} formalized education as investment in ``human capital''---skills that increase productivity and thereby justify wage differentials.
This framework provides the economic rationale for meritocracy: wage inequality reflects differential investment in skills, which reflects differential ability and effort.
The theory's assumptions (competitive labor markets, wages equal to marginal product) are empirically questionable, but its institutional influence is enormous---it is the foundation of education policy worldwide.

\paragraph{The ``talent'' discourse (1990s--).}
McKinsey's ``War for Talent'' (1997) and the broader talent management movement repackaged meritocratic ideology for corporate consumption.
The key move: redefining organizational success as a function of recruiting and retaining ``A players''---individuals with exceptional talent---rather than building effective systems.
This justified extreme executive compensation and created a self-reinforcing cycle: companies that pay more claim they are attracting better talent, which justifies paying even more.

\paragraph{Silicon Valley meritocracy (2000s--).}
Tech industry culture elevated meritocratic ideology to a founding myth: the garage startup, the dropout billionaire, the disruption narrative.
``Move fast and break things'' implies that outcomes reflect innovation and speed, not capital access, network effects, or regulatory arbitrage.
The tech industry's meritocratic self-image persists despite overwhelming evidence of racial and gender homogeneity in its leadership.

\section{Negative Cases: When Meritocratic Belief Fails}
\label{sec:ch22-negative}

Doxonomic analysis requires attention to failure as well as success.
Not all investments in meritocratic narrative succeed, and the conditions under which they fail are theoretically informative.

\begin{enumerate}
  \item \textbf{Economic crises.}
  In the aftermath of the 2008 financial crisis, meritocratic belief temporarily declined as structural causes became salient.
  Bankers who received million-dollar bonuses while being bailed out with public money provided a vivid counter-example to the desert narrative.
  But the decline was temporary: within five years, meritocratic belief had substantially recovered, suggesting that the institutional infrastructure is resilient to even severe shocks.

  \item \textbf{Strong labor movements.}
  In societies with strong unions and social-democratic traditions (Scandinavia, post-war UK), meritocratic belief is weaker despite similar media exposure.
  The counter-institutional infrastructure (unions, labor parties, cooperative traditions) provides an alternative framing: outcomes reflect bargaining power and institutional design, not individual merit.

  \item \textbf{Extreme inequality.}
  When inequality becomes so extreme that the gap between narrative and lived experience is too large, the legitimation function breaks down.
  The model predicts this: $\belief{M} \to 0$ as $V \to \infty$.
  Historical examples include the Gilded Age populist revolt and the post-2008 Occupy movement.

  \item \textbf{Racial counter-evidence.}
  Among Black Americans, meritocratic belief is consistently lower than among white Americans---the people for whom structural barriers are most visible are least persuaded by the meritocratic narrative \autocite{hochschild2016}.
  This racial gap in meritocratic belief is itself a doxonomic variable: it reflects different lived experiences with the meritocratic system.
\end{enumerate}

These negative cases constrain the model: doxonomic investment is necessary but not sufficient for common-sense capture.
Lived experience, counter-institutions, and competing narratives limit the reach of any regime.

\section{Racial and Gender Dimensions}
\label{sec:ch22-racial}

Meritocratic belief is not distributed uniformly.
It intersects with racial and gender regimes (Chapter~\ref{ch:race-coloniality-gender}) in specific and measurable ways.

\subsection{Racial meritocracy gap}

White respondents in the US consistently score higher on meritocratic belief scales than Black respondents.
The gap is not simply about optimism or pessimism; it reflects structurally different positions within the meritocratic system.
For a white professional whose parents were also professionals, the system does appear meritocratic: their advantages are invisible to them.
For a Black professional whose parents faced redlining, school segregation, and employment discrimination, the same system appears obviously rigged.

\subsection{The ``model minority'' function}

The ``model minority'' narrative applies meritocratic logic selectively to Asian Americans, using their achievements to delegitimize structural claims by other racialized groups.
The doxonomic function: ``If Asian Americans can succeed despite discrimination, the failure of other minorities must reflect their own deficiencies.''
This narrative disaggregates the experiences of very different Asian American communities (Japanese Americans with multi-generational US presence vs.\ recent Hmong refugees), selectively cites educational statistics, and ignores the bamboo ceiling in corporate leadership.

\subsection{Gender and the ``choice'' narrative}

The meritocratic regime reframes gender inequality as reflecting women's ``choices''---to prioritize family, to enter lower-paying fields, to negotiate less aggressively.
This narrative converts structural constraints (lack of affordable childcare, workplace discrimination, gendered socialization) into individual preferences.
The doxonomic analysis reveals that the ``choice'' narrative is itself produced and circulated by institutions with identifiable interests in maintaining the gender pay gap.

\section{Comparative Doxonomics of Meritocracy}
\label{sec:ch22-comparative}

\begin{example}
\label{ex:us-nordic-comparison}
A comparative analysis of meritocratic belief and its institutional infrastructure:

\begin{center}
\begin{tabular}{p{3.5cm}p{4.5cm}p{4.5cm}}
\toprule
& United States & Nordic countries \\
\midrule
Meritocratic belief (survey) & 70--80\% agree ``hard work leads to success'' & 40--50\% agree \\[4pt]
Actual intergenerational mobility & Low ($\beta \approx 0.47$) & High ($\beta \approx 0.15$--$0.20$) \\[4pt]
Key institutions & Competitive education, testing industry, business media, self-help & Universal education, strong unions, comprehensive welfare \\[4pt]
Counter-institutions & Weak unions, limited social safety net & Strong unions, labor parties, cooperative traditions \\[4pt]
Narrative investment (est.) & Very high (\$10B+) & Lower (state-funded, less commercial) \\
\bottomrule
\end{tabular}
\end{center}

The paradox: Americans believe in meritocracy more but experience it less.
The doxonomic explanation: the US has a more developed meritocratic narrative infrastructure and weaker counter-institutional infrastructure, allowing the narrative to persist despite contradictory evidence.
Nordic countries have lower meritocratic belief but higher actual mobility---and the institutional infrastructure to produce an alternative narrative (``success reflects social investment, not just individual effort'').
\end{example}

\section{Notes and References}
\label{sec:ch22-notes}

On meritocracy as ideology, see \textcite{sandel2012} and Young (1958).
Inequality and its legitimation are analyzed in \textcite{piketty2014} and \textcite{stiglitz2012}.
Intergenerational mobility data draws on \textcite{chetty2014} and Corak (2013).
Just-world belief is theorized in \textcite{lerner1980}.
Human capital theory originates with \textcite{becker1964}.

The role of education in reproducing class structure is central to \textcite{bourdieu1984} and \textcite{bowles1976}.
On credentialism, see Collins (1979).
On the racial dimensions of meritocratic belief, see \textcite{hochschild2016}.
The self-help industry as a doxonomic channel connects to \textcite{ehrenreich2009}.

On Silicon Valley's meritocratic ideology, see Marwick (2013).
The ``War for Talent'' is analyzed in Michaels et al.\ (2001).
Standardized testing and its social functions are critiqued in Lemann (1999).
For the Nordic comparison, see Esping-Andersen (1990) and Wilkinson and Pickett (2009).

\section{Exercises}

\begin{exercise}
Using Model~\ref{mod:meritocracy}, compute the narrative investment $\phi_M$ required to maintain $\belief{M} = 0.7$ when inequality visibility doubles (from $V = 5$ to $V = 10$, with $\kappa = 1$).
Then compute the investment for $\belief{M} = 0.9$.
What does the comparison tell you about the escalating cost of legitimation?
\end{exercise}

\begin{exercise}
Design a doxometric survey to measure meritocratic belief across socioeconomic groups.
\begin{enumerate}[nosep]
  \item Write five survey items that capture the four propositions ($p_1$--$p_4$) of Definition~\ref{def:meritocratic-regime}.
  \item Predict which socioeconomic groups will score highest on meritocratic belief and explain the prediction doxonomically.
  \item Describe how you would distinguish between genuine belief and social desirability bias.
  \item How would you test whether meritocratic belief predicts policy preferences (e.g., support for redistribution)?
\end{enumerate}
\end{exercise}

\begin{exercise}
Analyze the genre of ``billionaire biography'' as a doxonomic product.
\begin{enumerate}[nosep]
  \item Select two biographies of ultra-wealthy individuals.
  \item Apply framing analysis (Chapter~\ref{ch:text-media}) to identify how each biography frames the subject's success.
  \item How much of the narrative is attributable to individual talent/effort vs.\ structural advantage (inheritance, networks, timing, luck)?
  \item What doxonomic function does the genre serve?
\end{enumerate}
\end{exercise}

\begin{exercise}
Apply the credential laundering framework (Proposition~\ref{prop:credential}) to a specific credential of your choice (e.g., MBA from a top program, medical board certification, professional engineering license).
\begin{enumerate}[nosep]
  \item Identify the ``merit'' component $f$ and the ``wealth'' component $g$.
  \item Estimate the relative contribution of each to credential attainment.
  \item How does the credential's epistemic capital (credibility, authority) amplify the laundering effect?
\end{enumerate}
\end{exercise}

\begin{exercise}
Using the dynamic model (Model~\ref{mod:merit-inequality-feedback}), analyze the vicious spiral:
\begin{enumerate}[nosep]
  \item For what parameter values does the system reach a stable equilibrium vs.\ diverge?
  \item What real-world mechanism corresponds to the ``budget constraint'' that eventually limits narrative investment?
  \item What policy intervention could break the spiral? Represent it as a modification to the model equations.
\end{enumerate}
\end{exercise}

\begin{exercise}
Construct the US-Nordic comparison (Example~\ref{ex:us-nordic-comparison}) using actual data.
\begin{enumerate}[nosep]
  \item Find survey data on meritocratic belief in both contexts (e.g., World Values Survey, ISSP).
  \item Find intergenerational mobility estimates for both (e.g., Chetty et al., Corak).
  \item Map the institutional infrastructure in both (think tanks, educational systems, media).
  \item Does the doxonomic model (higher narrative investment, lower counter-institutions $\to$ higher meritocratic belief despite lower mobility) fit the data?
\end{enumerate}
\end{exercise}

\begin{exercise}[Computational]
Simulate the meritocracy-inequality feedback model (Model~\ref{mod:merit-inequality-feedback}):
\begin{enumerate}[nosep]
  \item Set initial conditions $\belief{M}(0) = 0.6$, $V(0) = 3$, with $\alpha = 0.1$, $\phi_M = 5$, $\beta = 0.05$, $\gamma = 0.5$, $\delta = 0.3$.
  \item Simulate 100 time steps using Euler's method.
  \item Plot $\belief{M}(t)$ and $V(t)$ on the same graph.
  \item Does the system reach equilibrium or spiral? At what point (if any) does the narrative investment become insufficient to maintain belief?
  \item Experiment with doubling $\phi_M$. How much does this delay the breakdown?
\end{enumerate}
\end{exercise}
