\chapter{Consumerism and Manufactured Desire}
\label{ch:consumerism}

\epigraph{People recognize themselves in their commodities; they find their soul in their automobile, hi-fi set, split-level home, kitchen equipment.}{Herbert Marcuse, \textit{One-Dimensional Man}}

\noindent
Advertising is the largest doxonomic industry.
Global advertising expenditure surpassed one trillion dollars in 2024, making it the single greatest investment in belief production in human history---dwarfing the budgets of all think tanks, universities, and government propaganda agencies combined.
This chapter analyzes consumerism not merely as an economic pattern (people buy things) but as a doxonomic system: a coordinated, massively funded apparatus for producing and maintaining the belief that identity, happiness, and social worth are constituted through consumption.

The individual advertisement promotes a product.
The advertising \emph{system} promotes an anthropological regime: you are what you buy, happiness is purchasable, identity is a consumption choice, and the good life consists in acquisition.
This regime is not self-evident---most human societies throughout history organized self-worth around kinship, craft, faith, civic virtue, or communal belonging rather than purchasing power.
That consumption became the primary idiom of identity in the 20th century is a doxonomic achievement, and this chapter traces its mechanisms.

\section{Advertising as Doxonomic Investment}
\label{sec:ch23-advertising}

\subsection{Scale and structure}

\begin{definition}[Advertising as Doxonomic Production]
\label{def:advertising-doxonomic}
\index{advertising}
Advertising is a doxonomic investment whose primary output is not product awareness but \emph{desire infrastructure}: the narrative environment in which consumption appears as the natural expression of identity, freedom, and self-worth.
\end{definition}

The scale is staggering:

\begin{center}
\begin{tabular}{lr}
\toprule
Category & Annual global expenditure (2023 est.) \\
\midrule
Digital advertising & \$600 billion \\
Television advertising & \$160 billion \\
Print, radio, outdoor & \$120 billion \\
Sponsorships and branded content & \$80 billion \\
\midrule
\textbf{Total advertising} & \textbf{\$960 billion} \\
\midrule
For comparison: & \\
\quad Global military spending & \$2.2 trillion \\
\quad Global public education spending & \$5.0 trillion \\
\quad All US think tanks combined & \$0.01 trillion \\
\bottomrule
\end{tabular}
\end{center}

The advertising industry is roughly 100 times larger than the entire think-tank sector.
In doxonomic terms, the production of consumer desire absorbs far more resources than the production of political ideology---though the two are intertwined, as Section~\ref{sec:ch23-political} explores.

\subsection{The dual product}

The advertising industry produces two outputs simultaneously:
\begin{enumerate}
  \item \textbf{Product-specific persuasion}: buy \emph{this} car, \emph{this} phone, \emph{this} beverage.
  This is the output that advertisers pay for and measure.

  \item \textbf{System-level belief production}: consumption is identity, material goods produce happiness, choice among brands is freedom, dissatisfaction is the natural human condition (solvable only by purchase).
  This is the output that no individual advertiser intends but that the system produces as a collective externality.
\end{enumerate}

The second output is the doxonomically important one.
No single Coca-Cola advertisement makes anyone believe that happiness is purchasable.
But 10,000 advertisements per day---the estimated exposure of the average American---collectively produce an environment in which that belief is the unspoken background assumption of daily life \autocite{adorno1944, debord1967}.

\section{The Desire Production Function}
\label{sec:ch23-desire}

\subsection{Basic model}

\begin{model}[Manufactured Desire]
\label{mod:desire}
\index{manufactured desire}
Let $D_j(t)$ be the desire intensity of agent $j$ for goods beyond subsistence.
Desire is produced by:
\[
  D_j(t) = D_0 + \alpha \sum_k \funding{k}^{\text{ad}} \cdot e_{jk}(t) - \beta \cdot S_j(t) + \gamma \cdot \bar{D}_{-j}(t)
\]
where:
\begin{itemize}[nosep]
  \item $D_0$ is baseline desire (the ``natural'' level absent commercial influence),
  \item $\funding{k}^{\text{ad}}$ is advertising investment through channel $k$, with $e_{jk}$ the exposure of $j$ to channel $k$,
  \item $-\beta \cdot S_j(t)$ captures \emph{induced dissatisfaction}: advertising reduces satisfaction with current possessions ($\beta > 0$),
  \item $\gamma \cdot \bar{D}_{-j}(t)$ captures social comparison: $j$'s desire increases with the average desire in $j$'s reference group ($\gamma > 0$).
\end{itemize}
\end{model}

The three mechanisms---direct stimulation ($\alpha$ term), dissatisfaction induction ($\beta$ term), and social comparison ($\gamma$ term)---operate simultaneously and reinforce each other.

\subsection{The hedonic treadmill}

\begin{proposition}[Advertising Treadmill]
\label{prop:treadmill}
\index{hedonic treadmill}
If satisfaction declines with the gap between desire and consumption, $S_j(t) = C_j(t) - D_j(t)$, and consumption rises to close the gap, $C_j(t+1) = C_j(t) + \eta(D_j(t) - C_j(t))$, then the system produces a treadmill:
\begin{enumerate}[nosep]
  \item Advertising increases desire: $D_j \uparrow$.
  \item The desire-consumption gap reduces satisfaction: $S_j \downarrow$.
  \item Reduced satisfaction further increases desire (via $-\beta S_j$): $D_j \uparrow\uparrow$.
  \item Consumption rises to partially close the gap: $C_j \uparrow$.
  \item But rising consumption raises the reference point for desire (through social comparison): $D_j \uparrow\uparrow\uparrow$.
  \item The cycle continues: consumption rises without lasting satisfaction gains.
\end{enumerate}
At steady state, $D_j^* = (D_0 + \alpha A_j) / (1 - \gamma + \beta)$ where $A_j = \sum_k \funding{k}^{\text{ad}} e_{jk}$.
For $\gamma$ close to 1 (strong social comparison), desire grows without bound unless $\beta$ is very large.
\end{proposition}

This model formalizes the empirical finding that, beyond a threshold of material sufficiency, increases in income and consumption produce diminishing and eventually zero returns to subjective well-being---the Easterlin paradox \autocite{easterlin1974}.
The doxonomic explanation: the advertising system continuously raises the aspiration level, ensuring that satisfaction never catches up to consumption.

\subsection{Worked example: the smartphone cycle}

\begin{example}
\label{ex:smartphone}
The smartphone upgrade cycle illustrates the desire production function:
\begin{enumerate}[nosep]
  \item \emph{Direct stimulation}: Apple spends $\sim$\$1.8 billion annually on advertising, Samsung $\sim$\$2.0 billion.
  \item \emph{Induced dissatisfaction}: each new model is positioned as making the previous model ``obsolete''---slower, lower-resolution, missing features---even when the previous model remains fully functional.
  \item \emph{Social comparison}: visible phone models signal status; using a two-year-old phone signals falling behind.
  \item \emph{Result}: the average smartphone replacement cycle is $\sim$2.7 years, generating $\sim$1.4 billion new phones sold annually, despite functional adequacy of existing devices.
\end{enumerate}
The desire production cost: $\sim$\$4 billion in direct advertising produces $\sim$\$450 billion in smartphone revenue.
The return on doxonomic investment is roughly 100:1---among the highest of any belief-production system.
\end{example}

\section{Consumption as Anthropological Regime}
\label{sec:ch23-moral}

\subsection{The consumerist belief system}

Consumerism is not merely an economic pattern but an anthropological regime: a coherent set of beliefs about what constitutes the good life.

\begin{definition}[Consumerist Regime]
\label{def:consumerist-regime}
\index{consumerism!regime}
The consumerist regime $\alpha_C$ is defined by:
\begin{enumerate}[nosep]
  \item $p_1$: Identity is constituted through consumption choices (``you are what you buy'').
  \item $p_2$: Material acquisition produces happiness.
  \item $p_3$: Consumer choice is the primary form of human agency and freedom.
  \item $p_4$: Dissatisfaction is the natural human condition, solvable by purchase.
  \item $p_5$: Social worth is legible through visible consumption.
\end{enumerate}
\end{definition}

Each proposition has a pre-consumerist alternative that it displaced:
\begin{center}
\begin{tabular}{ll}
\toprule
Consumerist proposition & Displaced alternative \\
\midrule
Identity through consumption & Identity through community, craft, faith \\
Happiness through acquisition & Happiness through relationships, meaning, sufficiency \\
Freedom as consumer choice & Freedom as political participation, self-governance \\
Dissatisfaction as natural & Contentment as virtue, sufficiency as ideal \\
Worth through display & Worth through character, contribution, belonging \\
\bottomrule
\end{tabular}
\end{center}

The displacement was not accomplished through argument but through institutional investment: the advertising system made the consumerist propositions the unspoken background of daily experience, while the alternatives were gradually de-funded and de-institutionalized.

\subsection{The historical construction}

The consumerist regime is historically specific.
Its construction can be periodized:

\paragraph{Early advertising (1880--1920).}
Patent medicines, department stores, and the first national brands created the apparatus of mass persuasion.
The advertising industry professionalized with the founding of agencies (J.\ Walter Thompson, 1864; N.\ W.\ Ayer, 1869) and trade publications.

\paragraph{The Bernays revolution (1920--1950).}
Edward Bernays---Freud's nephew---pioneered ``public relations'' as the application of psychological manipulation to commercial and political persuasion.
His campaigns (getting women to smoke by framing cigarettes as ``torches of freedom''; engineering a banana republic coup for United Fruit) demonstrated that desire could be manufactured at industrial scale \autocite{bernays1928}.

\paragraph{The golden age of mass advertising (1950--1990).}
Television created a delivery mechanism of unprecedented reach and intimacy.
The average American household watched 6+ hours of television daily, absorbing thousands of commercial messages per week.
The ``Mad Men'' era saw advertising expenditure grow from \$5.7 billion (1950) to \$130 billion (1990) in the US alone.

\paragraph{Digital surveillance advertising (2000--present).}
The internet and smartphone transformed advertising from broadcast persuasion to individualized targeting.
Digital platforms (Google, Meta, Amazon) collect granular behavioral data and use it to deliver personalized advertisements at scale.
The advertising model funds the infrastructure of the digital public sphere, making the content people consume financially dependent on the advertising system's need for attention.

\section{The Political Economy of Attention}
\label{sec:ch23-attention}

\subsection{Attention as the scarce resource}

In an information-rich environment, the scarce resource is not information but attention.
The advertising system competes for attention, and the institutional structures it creates---algorithmically curated social media feeds, autoplay video, notification systems, gamified engagement metrics---reshape the cognitive environment in which all beliefs (not just consumer beliefs) are formed.

\begin{model}[Attention Extraction]
\label{mod:attention}
\index{attention economy}
Let $T_j = 16$ hours be agent $j$'s daily waking time.
Attention is allocated across $n$ claims on $j$'s time: work, family, civic participation, leisure, and media consumption.
The advertising system's goal is to maximize $a_j^{\text{ad}}$, the fraction of $T_j$ spent encountering advertisements.
The optimization is:
\[
  \max_{d} \sum_j a_j^{\text{ad}}(d) \quad \text{subject to content design } d \text{ retaining user engagement.}
\]
Content platforms solve this optimization by designing for ``engagement''---content that maximizes time-on-platform---which produces a systematic bias toward emotionally arousing, conflict-laden, and status-signaling content.
\end{model}

The doxonomic consequence: the advertising system's optimization for attention shapes the information environment in which \emph{all} beliefs are formed.
Political beliefs, scientific understanding, and social norms are all filtered through an attention economy designed to sell products.
The consumerist regime is not just about what people buy; it is about the cognitive infrastructure within which all belief formation occurs.

\subsection{The attention tax}

\begin{proposition}[Attention Tax]\footnote{The figures here are order-of-magnitude estimates, not precise measurements. The 5,000 messages/day figure is widely cited in the marketing literature but methodologically contested; the 2-second processing assumption is a stylized average. The proposition illustrates a structural argument, not a calibrated empirical claim.}
\label{prop:attention-tax}
\index{attention tax}
The advertising system imposes an implicit ``attention tax'' on the population.
If the average American is exposed to $\sim$5,000 commercial messages per day and each requires $\sim$2 seconds of cognitive processing, the total daily attention tax is:
\[
  \tau_{\text{attn}} = 5{,}000 \times 2\text{s} = 10{,}000\text{s} \approx 2.8 \text{ hours/day.}
\]
This is time not available for civic engagement, reflection, education, or non-commercial social interaction.
The aggregate US attention tax: $330\text{M people} \times 2.8\text{h/day} \times 365 \approx 337$ billion person-hours per year devoted to processing commercial persuasion.
\end{proposition}

The attention tax is regressive: it falls most heavily on those who cannot afford ad-free environments (premium streaming, ad-free news, private education).
Low-income populations are exposed to more advertising per hour of media consumption, compounding the doxonomic effects.

\section{Consumerism and Other Doxonomic Regimes}
\label{sec:ch23-political}

\subsection{Consumerism reinforces the selfishness regime}

The consumerist and selfishness regimes are mutually reinforcing:
\begin{itemize}[nosep]
  \item Consumerism frames human motivation as desire-satisfaction, which maps onto the economic model of utility maximization.
  \item The selfishness regime legitimates consumerism: if humans are naturally acquisitive, then consumption is the natural expression of human nature.
  \item Both regimes individualize social problems: the selfishness regime says cooperation fails because people are selfish; the consumerist regime says happiness fails because you haven't bought the right things.
\end{itemize}

\subsection{Consumerism as political quietism}

\begin{remark}[Depoliticization Through Consumption]
\label{prop:depoliticization}
\index{depoliticization}
The consumerist regime substitutes consumer identity for political identity and consumer choice for political action.
If ``voting with your wallet'' replaces voting at the ballot box, the locus of agency shifts from the public to the private sphere.
This substitution serves the interests of actors who prefer a depoliticized population: corporations that benefit from weak regulation, and political actors who benefit from low civic participation.
\end{remark}

The mechanism is not conspiracy but structural incentive: the advertising system has no interest in civic engagement (which does not generate revenue) and strong interest in consumption (which does).
The resulting information environment systematically privileges consumer narratives over civic narratives, not because anyone designed it that way but because the revenue model rewards it.

\subsection{Green consumerism as doxonomic containment}

The rise of ``ethical'' or ``green'' consumerism illustrates doxonomic containment: absorbing a systemic critique (environmental destruction) into the consumerist framework by reframing it as a consumption choice (buy organic, drive electric, use reusable bags).

\begin{example}
\label{ex:green-consumerism}
The carbon footprint concept was popularized by BP's 2004 advertising campaign, which reframed climate change from a systemic problem (fossil fuel production) to an individual responsibility problem (your personal carbon footprint).
The campaign invested \$200+ million in shifting the locus of responsibility from producer to consumer---a textbook doxonomic operation.
The ``carbon footprint calculator'' became a widely used tool, embedding the individual-responsibility frame in millions of users' understanding of climate change.
\end{example}

\section{Resistance and Counter-Narratives}
\label{sec:ch23-resistance}

\subsection{Anti-consumerist movements}

Counter-narratives to the consumerist regime exist, though with vastly smaller institutional support:
\begin{itemize}[nosep]
  \item \textbf{Voluntary simplicity}: movements that advocate reduced consumption as a path to well-being (Thoreau's tradition, modern minimalism).
  \item \textbf{Adbusters and culture jamming}: subverting advertising images to reveal their doxonomic function.
  \item \textbf{Degrowth economics}: the academic movement arguing that economic growth beyond sufficiency reduces rather than increases well-being.
  \item \textbf{Religious and philosophical traditions}: ascetic traditions in Buddhism, Christianity, Stoicism, and Islam that value non-material goods.
  \item \textbf{Indigenous alternatives}: many indigenous communities maintain cosmologies in which the good life centers on reciprocity, balance, and communal well-being rather than individual accumulation.
\end{itemize}

\subsection{The funding asymmetry}

The counter-consumerist movement faces an overwhelming funding asymmetry.
The global advertising industry (\$960 billion) outspends all anti-consumerist movements combined by a factor of at least 10,000:1.
Even well-funded environmental organizations (\$10--20 billion globally) are outspent by the advertising budgets of the fossil fuel industry alone (\$200+ billion when including brand marketing for petroleum-derived products).

This asymmetry does not mean resistance is futile---the consumerist regime has cracks, and lived experience of its failures (debt, environmental degradation, social isolation despite material abundance) creates openings for counter-narratives.
But it does mean that the regime's persistence reflects funding rather than evidence: the ``good life through consumption'' hypothesis has been substantially falsified by the well-being literature, yet it remains dominant because of institutional investment.

\section{Subjective Well-Being Evidence}
\label{sec:ch23-wellbeing}

\subsection{The Easterlin paradox}

\begin{remark}[Easterlin Paradox]
\label{prop:easterlin}
\index{Easterlin paradox}
Above a threshold of material sufficiency ($\sim$\$75,000/year in 2010 US dollars, adjusted for household size), additional income and consumption produce diminishing and eventually negligible gains in reported subjective well-being \autocite{easterlin1974}.
Cross-nationally, GDP per capita correlates weakly with life satisfaction among wealthy countries.
(Note: the exact threshold and the strength of the paradox remain debated; Killingsworth 2021 finds continued gains at higher incomes, though with diminishing magnitude.)
\end{remark}

The doxonomic implication: the consumerist regime's central promise (happiness through acquisition) is empirically false for the majority of the population in wealthy countries.
Yet the belief persists---and billions of dollars are spent annually to maintain it.
This is a paradigmatic case of doxonomic belief production overriding empirical evidence.

\subsection{What does produce well-being?}

The subjective well-being literature identifies several factors that matter more than consumption above the sufficiency threshold:
\begin{itemize}[nosep]
  \item social relationships and community belonging,
  \item sense of meaning and purpose,
  \item autonomy and self-determination,
  \item physical health and security,
  \item trust in institutions and other people.
\end{itemize}

Several of these factors are \emph{undermined} by the consumerist regime: social atomization (shopping replaces socializing), meaning deficit (consumption as substitute for purpose), erosion of trust (competitive display replaces cooperative engagement).
The consumerist regime thus undermines the very conditions it claims to fulfill---a structural parallel to the self-fulfilling selfishness prophecy of Chapter~\ref{ch:human-nature}.

\section{Notes and References}
\label{sec:ch23-notes}

The critical theory of consumer society draws on \textcite{adorno1944}, \textcite{debord1967}, and \textcite{fromm1976}.
On the history of advertising and public relations, see \textcite{bernays1928} and Ewen (1976).
On advertising and identity, see \textcite{veblen1899} and \textcite{galbraith1958}.
Induced dissatisfaction and the work-spend cycle are analyzed in \textcite{schor1998}.

The Easterlin paradox is in \textcite{easterlin1974}; for updated evidence, see Kahneman and Deaton (2010) and Killingsworth (2021).
For the economics of manipulation through advertising, see \textcite{akerlof2015}.
On the attention economy, see Wu (2016) and Zuboff (2019).
The political economy of digital advertising is analyzed in McChesney (2013).

Green consumerism and the individualization of environmental responsibility are critiqued in Maniates (2001).
On the degrowth alternative, see Hickel (2020).
For indigenous alternatives to consumerism, see Gudynas (2011) on \emph{buen vivir}.

\section{Exercises}

\begin{exercise}
Estimate the advertising expenditure per capita in your country and compare it to per-capita spending on public education.
Interpret the ratio doxonomically: what does it imply about the relative investment in consumer desire vs.\ civic formation?
\end{exercise}

\begin{exercise}
Using the desire production model (Model~\ref{mod:desire}), show formally that if $\beta > 0$ and $\gamma > 0$, advertising simultaneously increases desire \emph{and} decreases satisfaction, creating a treadmill effect.
\begin{enumerate}[nosep]
  \item Compute the steady-state desire $D^*$ as a function of advertising expenditure $A$.
  \item Show that $\partial D^* / \partial A > 0$ (more advertising, more desire).
  \item Compute steady-state satisfaction $S^*$ and show that $\partial S^* / \partial A < 0$ (more advertising, less satisfaction).
  \item At what level of social comparison ($\gamma$) does the treadmill become unstable?
\end{enumerate}
\end{exercise}

\begin{exercise}
Analyze a specific advertising campaign as a belief production chain (Chapter~\ref{ch:belief-production-chain}).
Identify: funder, institution, channel, narrative content, intended belief change, and behavioral consequence.
Which of the five consumerist propositions (Definition~\ref{def:consumerist-regime}) does the campaign reinforce?
\end{exercise}

\begin{exercise}
Compute the attention tax (Proposition~\ref{prop:attention-tax}) for your own daily media diet.
\begin{enumerate}[nosep]
  \item Estimate the number of commercial messages you encounter per day across all channels.
  \item Estimate the total processing time.
  \item What fraction of your waking hours is devoted to processing commercial persuasion?
  \item How does this compare to time spent on civic engagement, education, or community interaction?
\end{enumerate}
\end{exercise}

\begin{exercise}
Apply the ``doxonomic containment'' analysis (Example~\ref{ex:green-consumerism}) to another social movement that has been absorbed into the consumerist framework.
\begin{enumerate}[nosep]
  \item Describe the original systemic critique.
  \item How was it reframed as a consumption choice?
  \item Who funded the reframing, and through what institutional channels?
  \item Does the consumer-friendly version address the original problem, or does it obscure it?
\end{enumerate}
\end{exercise}

\begin{exercise}
Design a comparative study of consumerist belief strength across two countries with different advertising environments (e.g., a high-advertising country like the US vs.\ a country with advertising restrictions like Norway or Cuba).
\begin{enumerate}[nosep]
  \item Specify the survey items that would measure consumerist belief (Definition~\ref{def:consumerist-regime}).
  \item Specify the advertising environment variables you would measure.
  \item What confounders must you control for?
  \item What result would count as evidence for vs.\ against the doxonomic model of consumerism?
\end{enumerate}
\end{exercise}

\begin{exercise}[Computational]
Simulate the hedonic treadmill model (Proposition~\ref{prop:treadmill}) for 100 agents over 200 time periods:
\begin{enumerate}[nosep]
  \item Set $D_0 = 1$, $\alpha = 0.01$, $\beta = 0.3$, $\gamma = 0.4$, $\eta = 0.5$.
  \item Advertising starts at $A = 50$ and doubles at $t = 100$.
  \item Plot average desire $\bar{D}(t)$, average consumption $\bar{C}(t)$, and average satisfaction $\bar{S}(t)$ over time.
  \item Does the doubling of advertising produce a lasting increase in satisfaction?
  \item What happens if you set $\gamma = 0$ (no social comparison)? Does the treadmill still operate?
\end{enumerate}
\end{exercise}
