\chapter{Nation, Civilization, and Collective Myth}
\label{ch:nation-myth}

\epigraph{Nations are not something eternal. They had their beginnings and they will end.}{Ernest Renan, \textit{What Is a Nation?} (1882)}

\noindent
National narratives are among the most expensive doxonomic products in history.
Every modern state invests in the production and maintenance of national identity: through education, monuments, ceremonies, media, and military ritual.
The budgets are enormous---dwarfing anything a private think tank could deploy---because the state itself is the primary funder, and the survival of the state depends on the narrative's success.

This chapter applies the doxonomic framework to nationalism: the production of collective identity, the management of historical memory, the construction of civilizational narratives, and the conditions under which national myths fracture or are replaced.
Unlike the previous application chapters---where the doxonomic agents are primarily private (corporations, foundations, think tanks)---the state is the central doxonomic actor here, with its unmatched capacity to fund narrative infrastructure through taxation.

\section{National Narrative as Doxonomic Product}
\label{sec:ch27-national}

\subsection{The structure of national myth}

\begin{definition}[National Myth]
\label{def:national-myth}
\index{national myth}
A \emph{national myth} is a structured narrative $n_{\text{nat}} \in \narspace$ specifying:
\begin{enumerate}[nosep]
  \item a \emph{founding story}: origin, sacrifice, triumph (the Revolution, the Landing, the Liberation),
  \item a \emph{collective character}: virtues attributed to the nation (American ingenuity, French rationalism, Japanese harmony, British resilience),
  \item an \emph{enemy or other}: defining the boundary of ``us'' by contrast with ``them'' (the empire, the barbarian, the foreign ideology),
  \item a \emph{destiny}: what the nation is called to achieve (manifest destiny, civilizing mission, historical mission of the proletariat),
  \item a \emph{sacred text or canon}: founding documents, literary traditions, historical narratives that anchor the other four elements.
\end{enumerate}
\end{definition}

The national myth is not a single narrative but a \emph{narrative system}---an interconnected set of stories, images, rituals, and institutions that collectively produce the experience of national belonging.
\textcite{anderson2006} famously described the nation as an ``imagined community'': a group too large for face-to-face interaction whose members nevertheless feel a profound sense of shared identity.
The doxonomic question is: what institutional infrastructure produces and sustains this imagining?

\subsection{The myth maintenance model}

\begin{model}[Myth Maintenance]
\label{mod:myth-maintenance}
\index{myth maintenance}
The strength of national identification $\belief{N}(t)$ requires continuous investment:
\[
  \frac{d\belief{N}}{dt} = \alpha \phi_N - \delta \belief{N} + \beta \belief{N}(1 - \belief{N}) - \gamma \cdot D(t)
\]
where:
\begin{itemize}[nosep]
  \item $\phi_N$ is the institutional investment in national narrative (education budgets, state media, military spending on ceremony and recruitment, holiday infrastructure),
  \item $\delta$ is natural decay (generational replacement, exposure to alternative identities, immigration),
  \item $\beta$ captures social reinforcement (national identity is stronger when peers share it),
  \item $D(t)$ measures dissonance from events contradicting the myth (military defeats, economic crises, revelations of historical atrocities, visible hypocrisy between national values and state actions).
\end{itemize}
\end{model}

The logistic term $\beta \belief{N}(1 - \belief{N})$ is crucial: national identification has the strongest peer reinforcement at intermediate levels.
A deeply nationalistic society reinforces itself; a deeply cosmopolitan one reinforces its alternative.
The unstable point is in the middle, where small doxonomic investments can tip the population in either direction.

\section{The Institutional Infrastructure of National Identity}
\label{sec:ch27-infrastructure}

\subsection{Education}

The school system is the primary delivery mechanism for national narrative.
History curricula, language instruction, civic education, and even mathematics word problems (``a patriot buys 3 flags\ldots'') are sites of national identity production.

\begin{example}
\label{ex:education-national}
A rough accounting of the US national-narrative education budget:
\begin{itemize}[nosep]
  \item K-12 history and social studies instruction: $\sim$\$15--20 billion/year (estimated from teacher salaries, textbook costs, and curriculum development in the relevant fields).
  \item Daily Pledge of Allegiance: performed in $\sim$100,000 schools, $\sim$180 days/year, for $\sim$50 million students. Total ritual-hours: $\sim$150 million hours/year.
  \item Standardized testing of ``American history'': the AP US History exam alone is taken by $\sim$500,000 students annually, requiring months of curriculum dedicated to a specific narrative of national development.
\end{itemize}
No private doxonomic actor commands anything close to this institutional reach.
\end{example}

\subsection{Military and security institutions}

Military institutions serve a dual doxonomic function: they defend the state and they produce the narrative that justifies the state's claim on citizens' loyalty (and lives).

\begin{itemize}[nosep]
  \item \textbf{Recruitment advertising}: the US military spends $\sim$\$1.5 billion/year on recruitment, much of it narrative-driven (``Army Strong,'' ``Be All You Can Be,'' ``The Few. The Proud.'').
  \item \textbf{Ceremony and ritual}: parades, flyovers at sporting events, flag ceremonies, memorial services---all funded by the state and designed to produce emotional identification with the nation.
  \item \textbf{Entertainment partnerships}: the Pentagon's entertainment liaison office reviews and influences Hollywood scripts in exchange for access to military equipment and locations, shaping the cinematic representation of the nation at war.
  \item \textbf{Veterans' institutions}: the VA, veterans' organizations, and Memorial/Veterans Day observances maintain the narrative of national sacrifice.
\end{itemize}

\subsection{State media and public broadcasting}

Many states operate media organizations whose explicit mandate includes national identity production: the BBC (UK), NHK (Japan), France T\'{e}l\'{e}visions, RT (Russia), CCTV (China).
Even ostensibly independent public broadcasters operate within a national framework---their programming assumes a national audience and reinforces national reference points.

\subsection{Monuments, museums, and public space}

The physical landscape is a doxonomic medium.
National monuments, war memorials, museum exhibits, street names, and public art all encode national narratives in material form.
Unlike textbooks, which can be revised, monuments are physically durable---they persist across generations, anchoring national identity in the built environment.

\begin{proposition}[Material Persistence of National Narrative]
\label{prop:material-persistence}
\index{material persistence}
Physical infrastructure (monuments, place names, architecture) has lower depreciation than textual or digital narratives.
A monument can sustain a national narrative for centuries with minimal ongoing investment, while a media campaign requires continuous funding.
This gives physical infrastructure a distinctive role in the doxonomic economy: it provides a ``base layer'' of national identity that persists even when active narrative production declines.
\end{proposition}

\section{Historical Memory Management}
\label{sec:ch27-memory}

\subsection{Selection and omission}

States actively manage historical memory: what is remembered, how it is framed, and what is omitted.
Curriculum design, museum curation, memorial construction, and holiday designation are all doxonomic operations with measurable budgets \autocite{anderson2006}.

\begin{definition}[Historical Memory Regime]
\label{def:memory-regime}
\index{historical memory}
A \emph{historical memory regime} is the institutionally maintained selection of historical events, figures, and narratives that constitute a nation's ``usable past.''
It includes:
\begin{enumerate}[nosep]
  \item \emph{Inclusion}: events that are actively commemorated and taught (victories, foundings, heroic sacrifices).
  \item \emph{Framing}: how included events are interpreted (the Civil War as ``states' rights'' vs.\ ``slavery''; colonialism as ``civilizing mission'' vs.\ ``exploitation'').
  \item \emph{Omission}: events that are systematically excluded from public memory (massacres, forced relocations, failed policies, uncomfortable alliances).
  \item \emph{Periodization}: which periods are treated as ``golden ages'' and which as ``dark ages,'' shaping the narrative arc of national destiny.
\end{enumerate}
\end{definition}

\subsection{Contested memory}

Historical memory is never uncontested.
Counter-memories---maintained by marginalized communities, academic historians, and dissident institutions---challenge the official narrative.
The doxonomic competition between official and counter-memory is visible in:
\begin{itemize}[nosep]
  \item \textbf{Textbook wars}: battles over how to teach the history of slavery, colonialism, indigenous peoples, and immigration (e.g., the Texas textbook controversies, Japan's treatment of WWII atrocities, Turkey and the Armenian genocide).
  \item \textbf{Monument controversies}: debates over Confederate monuments (US), colonial statues (UK, Belgium), and Soviet-era memorials (Eastern Europe).
  \item \textbf{Truth commissions}: post-conflict societies (South Africa, Argentina, Rwanda) that attempt to reconstruct historical memory as a collective project.
  \item \textbf{Memory laws}: legal prohibitions on certain historical claims (Holocaust denial laws, Russia's prohibition on ``falsifying'' WWII history) that use state power to enforce a specific memory regime.
\end{itemize}

\begin{example}
\label{ex:memory-contest}
The memory of the US Civil War has been actively managed for 160 years:
\begin{itemize}[nosep]
  \item \emph{Lost Cause narrative} (1870s--present): the war was about states' rights, not slavery; Confederate soldiers were honorable; Reconstruction was a disaster. Produced by the United Daughters of the Confederacy, textbook campaigns, and monument construction (most Confederate monuments were erected 1890--1950, during peak Jim Crow, not during the war).
  \item \emph{Abolitionist/emancipation narrative}: the war was fundamentally about slavery; the Union cause was morally just; Reconstruction was an unfinished revolution. Maintained by African American communities, academic historians, and civil rights organizations.
  \item The doxonomic competition between these narratives has concrete consequences: which narrative dominates affects support for racial equality policies, attitudes toward Confederate symbolism, and the content of K-12 education in the American South.
\end{itemize}
\end{example}

\section{Civilizational Narratives}
\label{sec:ch27-civilization}

Beyond national myth, larger-scale civilizational narratives operate: ``the West,'' ``the Islamic world,'' ``Asian values,'' ``the Global South.''
These are doxonomic products at a macro scale, grouping diverse societies into supposedly coherent civilizational units.

\begin{proposition}[Civilizational Narrative as Meta-Regime]
\label{prop:civilizational}
\index{civilizational narrative}
Civilizational narratives function as meta-regimes: they provide the background assumptions within which national narratives operate.
``Western civilization'' is not a neutral description of a geographic-cultural zone but a doxonomic construct that:
\begin{enumerate}[nosep]
  \item selects a specific intellectual lineage (Athens $\to$ Rome $\to$ Renaissance $\to$ Enlightenment $\to$ democracy) while omitting others (Byzantium, Islamic golden age, Chinese technological leadership, African kingdoms),
  \item attributes the selected achievements to an essentialized ``Western'' character rather than to contingent historical circumstances,
  \item serves as the foundation for geopolitical narratives (``defending Western values,'' ``the clash of civilizations'') with material consequences.
\end{enumerate}
\end{proposition}

The ``Western civilization'' narrative is institutionally produced through university curricula (``Western Civ'' courses, which became standard in US universities after WWI as a tool for national identity formation), museum curation (European art at the center, everything else as ``world cultures''), and geopolitical discourse (NATO as defender of ``the free world'').

\section{National Myth and Economic Interest}
\label{sec:ch27-economic}

National narratives interact with economic interests in specific ways:

\begin{enumerate}
  \item \textbf{Military-industrial legitimation.}
  The narrative of national threat justifies military spending, which benefits defense contractors.
  The annual US defense budget ($\sim$\$850 billion) is partly a doxonomic product: it requires a narrative of existential threat to sustain political support.
  Defense industry lobbying ($\sim$\$100 million/year) funds the production of this narrative.

  \item \textbf{Trade nationalism.}
  ``Buy American,'' ``Made in Germany,'' ``Cool Japan'' campaigns fuse national identity with consumption, creating doxonomic synergy between nationalism and consumerism (Chapter~\ref{ch:consumerism}).

  \item \textbf{Immigration narratives.}
  The framing of immigration as threat or opportunity serves different economic interests: employers benefit from immigration (cheaper labor), while native workers may face wage competition.
  National narratives about immigration (``nation of immigrants'' vs.\ ``invasion'') track these interests with remarkable fidelity.

  \item \textbf{Austerity nationalism.}
  The Greek case (Chapter~\ref{ch:austerity}) illustrates how national stereotypes (``lazy Greeks'') serve economic interests (justifying creditor-imposed austerity).
  The narrative of national character becomes a tool of transnational economic discipline.
\end{enumerate}

\section{Myth Fracture and Reconstruction}
\label{sec:ch27-fracture}

\subsection{When myths break}

National myths can fracture when dissonance $D(t)$ exceeds the narrative infrastructure's capacity to absorb it:

\begin{enumerate}[nosep]
  \item \textbf{Military defeat}: especially when the myth includes martial prowess (France 1940, Russia in Afghanistan, the US in Vietnam/Iraq).
  \item \textbf{Regime change}: revolutions, coups, and decolonization require wholesale myth reconstruction (post-Soviet states, post-colonial nations).
  \item \textbf{Historical revelation}: declassified documents, archaeological discoveries, or truth commissions that contradict the official narrative.
  \item \textbf{Demographic transformation}: immigration that changes the population's composition faster than the myth can absorb, creating tension between the narrative ``we'' and the actual population.
\end{enumerate}

\subsection{Myth reconstruction}

After fracture, new myths must be constructed---and the construction process reveals the doxonomic machinery in sharp relief.
Post-Soviet states invested heavily in new national narratives; post-apartheid South Africa constructed the ``Rainbow Nation'' myth; post-WWII Germany invested in the ``Vergangenheitsbew\"{a}ltigung'' (coming to terms with the past) narrative.

\begin{example}
\label{ex:german-memory}
Germany's post-WWII memory management is perhaps the most self-conscious exercise in national myth reconstruction:
\begin{itemize}[nosep]
  \item Mandatory Holocaust education in all school curricula.
  \item The Memorial to the Murdered Jews of Europe (Berlin, 2005) as a permanent acknowledgment of national guilt.
  \item \emph{Vergangenheitsbew\"{a}ltigung} as a cultural norm: the expectation that the nation must continuously confront rather than suppress its history.
  \item This reconstruction required enormous institutional investment and produced a distinctive national identity: ``Germany is the nation that confronts its past''---itself a narrative that serves contemporary legitimation functions.
\end{itemize}
\end{example}

\section{Notes and References}
\label{sec:ch27-notes}

On nations as imagined communities, see \textcite{anderson2006}.
State simplification and legibility are analyzed in \textcite{scott1998}.
On propaganda and national myth, see \textcite{lippmann1922} and \textcite{bernays1928}.
For the political economy of nationalism, see \textcite{acemoglu2012}.

On historical memory and its politics, see Assmann (2011) and Nora (1989).
The ``Lost Cause'' narrative is analyzed in Blight (2001).
On ``Western civilization'' as a doxonomic construct, see Appiah (2018).
Civilizational narratives and geopolitics connect to Huntington (1996) and Said's critique in \textcite{said1978}.

On military-entertainment partnerships, see Suid (2002).
For monument controversies, see Savage (2009).
On textbook politics, see Loewen (1995).
Memory laws and their implications are discussed in Belavusau and Gliszczy\'{n}ska-Grabias (2017).

\section{Exercises}

\begin{exercise}
Estimate the annual doxonomic investment in national identity for a country of your choice.
\begin{enumerate}[nosep]
  \item Include: education (history curricula), military (recruitment, ceremony, entertainment partnerships), state media, monuments and museums, holidays and public ceremonies.
  \item Express the total as a fraction of GDP.
  \item Compare this to the investment in any single private doxonomic campaign (e.g., a think tank's annual budget).
  \item What does the comparison reveal about the relative doxonomic power of states vs.\ private actors?
\end{enumerate}
\end{exercise}

\begin{exercise}
Using the myth maintenance model (Model~\ref{mod:myth-maintenance}), analyze the effect of a major dissonance event on national identification.
\begin{enumerate}[nosep]
  \item Choose a specific event (e.g., a military defeat, a historical revelation, an economic crisis).
  \item Estimate $D(t)$ (the dissonance magnitude) and its duration.
  \item At what $D(t)$ does $\belief{N}$ begin to decline despite continued institutional investment?
  \item What additional investment $\Delta \phi_N$ would be needed to maintain $\belief{N}$ at its pre-crisis level?
\end{enumerate}
\end{exercise}

\begin{exercise}
Compare the national myths of two countries using Definition~\ref{def:national-myth}.
\begin{enumerate}[nosep]
  \item Identify the five components (founding story, collective character, enemy/other, destiny, sacred text) for each.
  \item Map the institutional infrastructure that produces each myth.
  \item Identify the structural parallels in their belief production chains.
  \item Where do the myths contradict historical evidence, and how is the contradiction managed?
\end{enumerate}
\end{exercise}

\begin{exercise}
Analyze a textbook war (e.g., US Civil War historiography, Japan's WWII memory, India's partition narratives) as a doxonomic competition.
\begin{enumerate}[nosep]
  \item Identify the competing narratives and the institutions that produce each.
  \item Trace the funding and institutional authority behind each side.
  \item What has determined which narrative dominates in the school curriculum?
  \item How does the winning narrative affect contemporary political attitudes?
\end{enumerate}
\end{exercise}

\begin{exercise}
Apply the civilizational meta-regime analysis (Proposition~\ref{prop:civilizational}) to the concept of ``Western civilization.''
\begin{enumerate}[nosep]
  \item Trace the institutional history of ``Western Civ'' courses in US universities.
  \item What was included and excluded from the canonical narrative?
  \item How has the narrative changed since the 1990s ``culture wars''?
  \item What geopolitical functions does the civilizational narrative serve?
\end{enumerate}
\end{exercise}

\begin{exercise}
Design a study to measure the doxonomic output of a national monument or memorial.
\begin{enumerate}[nosep]
  \item Choose a specific site.
  \item How would you measure its effect on national identification (surveys of visitors vs.\ non-visitors? before/after construction? geographic proximity as instrument?)?
  \item What beliefs does the monument encode, and do visitors absorb them?
  \item How does the monument's effect compare to other channels of national narrative (education, media)?
\end{enumerate}
\end{exercise}

\begin{exercise}[Computational]
Simulate the myth maintenance model for a nation experiencing a dissonance crisis:
\begin{enumerate}[nosep]
  \item Set $\belief{N}(0) = 0.8$, $\alpha = 0.05$, $\phi_N = 5$, $\delta = 0.03$, $\beta = 0.4$.
  \item Run for 200 periods with $D(t) = 0$ (no dissonance). Verify stability.
  \item At $t = 50$, introduce a crisis: $D(t) = 3$ for $t \in [50, 60]$, then $D(t) = 0$.
  \item Plot $\belief{N}(t)$. Does the system recover to its pre-crisis level?
  \item Now simulate a ``myth reconstruction'' scenario: at $t = 100$, increase $\phi_N$ by 50\% (state increases investment in national narrative). How long does recovery take?
  \item At what crisis magnitude $D$ does the system fail to recover even with increased investment?
\end{enumerate}
\end{exercise}
