\chapter{Suggested Datasets, Archives, and Corpora}
\label{app:datasets}

This appendix lists datasets suitable for doxonomic research.

\begin{longtable}{p{4cm} p{9.5cm}}
\toprule
\textbf{Dataset} & \textbf{Description} \\
\midrule
\endhead

World Values Survey & Cross-national survey of beliefs and values in 100+ countries (1981--present). Core resource for doxometric measurement. \\

General Social Survey (GSS) & U.S.\ attitudes, behaviors, and beliefs since 1972. Annual or biennial. \\

Eurobarometer & EU-wide public opinion surveys on trust, institutions, and social values. \\

OpenSecrets Bulk Data & U.S.\ federal lobbying, PAC contributions, and outside spending. Machine-readable. \\

ProPublica Nonprofit Explorer & Searchable database of IRS 990 filings for U.S.\ nonprofits. \\

GDELT Project & Real-time global media monitoring: events, tone, themes, and locations. \\

Media Cloud & Open-source platform for analyzing media ecosystems and narrative flows. \\

Google Ngrams & Word and phrase frequency in the Google Books corpus (1500--2019). \\

Google Trends & Search interest over time by region. Proxy for public attention. \\

Common Crawl & Petabyte-scale web archive for large-scale text analysis. \\

Facebook Ad Library & Archive of political and social issue advertisements on Meta platforms. \\

OECD Education at a Glance & Cross-national data on education spending, enrollment, and curricula. \\

UNESCO Institute for Statistics & Global data on media, education, and cultural spending. \\

Pew Global Attitudes & Cross-national surveys on democracy, economics, technology, and social values. \\

ISSP & International Social Survey Programme: annual thematic cross-national surveys. \\

ICPSR & Inter-university Consortium for Political and Social Research: massive archive of social science datasets. \\

HathiTrust Digital Library & 17+ million digitized volumes for computational text analysis. \\

Internet Archive & Web and media archives including the Wayback Machine. \\

InfluenceWatch & Profiles and funding data for advocacy organizations, unions, and think tanks. \\

Corporate Europe Observatory & Data on EU corporate lobbying and revolving doors. \\

\bottomrule
\end{longtable}
