\chapter{Recommended Reading}
\label{app:reading}

Organized by topic, with brief annotations.

\section*{Political Economy}

\begin{itemize}[nosep]
  \item \textcite{marx1846} --- The foundational text on ideology as material production.
  \item \textcite{polanyi1944} --- How markets were politically constructed, not naturally emergent.
  \item \textcite{north1990} --- Institutions as the rules of the game shaping economic behavior.
  \item \textcite{acemoglu2012} --- How political institutions determine economic outcomes.
  \item \textcite{piketty2014} --- The dynamics of wealth concentration and its legitimation.
  \item \textcite{stiglitz2012} --- How inequality is maintained through institutional design.
  \item \textcite{streeck2016} --- The contradictions of contemporary capitalism.
  \item \textcite{harvey2005} --- The institutional history of neoliberalism.
\end{itemize}

\section*{Sociology of Knowledge and Symbolic Power}

\begin{itemize}[nosep]
  \item \textcite{bourdieu1984} --- The social production of taste, distinction, and doxa.
  \item \textcite{bourdieu1991} --- Language as an instrument of symbolic power.
  \item \textcite{gramsci1971} --- Hegemony and the production of consent through civil society.
  \item \textcite{foucault1971} --- The rules governing what counts as knowledge.
  \item \textcite{lukes2005} --- Three dimensions of power, including the power to shape desires.
  \item \textcite{scott1998} --- How states impose legibility on complex social realities.
  \item \textcite{eyal2019} --- The sociology of expertise and its social authority.
\end{itemize}

\section*{Propaganda, Media, and Public Opinion}

\begin{itemize}[nosep]
  \item \textcite{lippmann1922} --- The manufacture of public opinion through media.
  \item \textcite{dewey1927} --- The public, communication, and democratic life.
  \item \textcite{bernays1928} --- The engineering of consent as a professional practice.
  \item \textcite{adorno1944} --- The culture industry and mass deception.
  \item \textcite{herman1988} --- The propaganda model of media: five structural filters.
  \item \textcite{mcchesney1999} --- Corporate media and the erosion of democratic communication.
  \item \textcite{bagdikian2004} --- Media ownership concentration and its effects.
  \item \textcite{zuboff2019} --- Surveillance capitalism and the commodification of behavior.
  \item \textcite{sunstein2017} --- Echo chambers and the fragmentation of public discourse.
\end{itemize}

\section*{Behavioral Science and Cooperation}

\begin{itemize}[nosep]
  \item \textcite{kahneman2011} --- Cognitive biases and dual-process theory.
  \item \textcite{fischbacher2001} --- Experimental evidence for conditional cooperation.
  \item \textcite{henrich2001} --- Cross-cultural evidence against homo economicus.
  \item \textcite{bowles2011} --- Cooperation as an evolved human capacity.
  \item \textcite{ostrom1990} --- Self-governance of commons without external coercion.
  \item \textcite{gintis2005} --- Moral sentiments and material interests in economic behavior.
  \item \textcite{nowak2006} --- Evolutionary mechanisms for cooperation.
\end{itemize}

\section*{Formal Methods}

\begin{itemize}[nosep]
  \item \textcite{jackson2008} --- Network theory for social and economic systems.
  \item \textcite{schelling1978} --- Micromotives, macrobehavior, and threshold effects.
  \item \textcite{granovetter1978} --- Threshold models of collective behavior.
  \item \textcite{pearl2009} --- Structural causal models and DAGs.
  \item \textcite{angrist2009} --- Causal inference for empirical research.
  \item \textcite{blei2003} --- Latent Dirichlet Allocation for text analysis.
  \item \textcite{shannon1948} --- Information theory: entropy and communication.
\end{itemize}

\section*{Case Studies and Applied Doxonomics}

\begin{itemize}[nosep]
  \item \textcite{oreskes2010} --- How doubt was manufactured about climate science and tobacco.
  \item \textcite{mayer2016} --- The dark-money network behind American conservatism.
  \item \textcite{mirowski2009} --- The institutional history of neoliberal thought.
  \item \textcite{klein2007} --- Crisis as an opportunity for doxonomic capture.
  \item \textcite{mudge2018} --- How left parties adopted neoliberal frameworks.
  \item \textcite{brown2015} --- Neoliberalism and the undoing of democratic subjectivity.
  \item \textcite{graeber2011} --- The doxonomic history of debt and moral obligation.
\end{itemize}

\section*{Alternatives and Futures}

\begin{itemize}[nosep]
  \item \textcite{wright2010} --- Real utopias: institutional designs for a more just society.
  \item \textcite{benkler2006} --- Networked information economy and peer production.
  \item \textcite{sen1999} --- Development as the expansion of capabilities.
  \item \textcite{freire1968} --- Critical pedagogy and the education of political consciousness.
  \item \textcite{srnicek2015} --- Post-work futures and counter-hegemonic strategy.
  \item \textcite{mason2015} --- Postcapitalism and the information economy.
\end{itemize}
